---
title: "Voronoi To The World"
author: "Michael Betancourt"
date: "November 2025"
date-format: "MMMM YYYY"
toc: true
number-sections: true
highlight: pygments
bibliography: voronoi.bib
crossref:
  lst-title: "Stan Program"
filters:
   - include-code-files
format:
  html:
    html-math-method: katex
    theme:
      - lux
      - custom.scss
    standalone: true
    embed-resources: true
    code-overflow: wrap
    linkcolor: "#B97C7C"
  pdf:
    keep-tex: true
    fig-width: 4
    fig-height: 4
    code-overflow: wrap
    monofontoptions:
      - Scale=0.5
  epub:
    css: custom.scss

knitr:
  opts_chunk:
    comment: ''
  opts_knit:
    global.par: TRUE
format-links: false
---

Voronoi diagrams arise naturally in various spatial analyses.
More importantly for this author, they also tend to be
aesthetically attractive, making them useful for crafting
compelling illustrations.

Efficiently constructing a Voronoi diagram, however, is not
particularly straightforward.  In this note I work through the
geometric structure of Voronoi diagrams and an algorithm that
leverages that structure to construct Voronoi diagrams with the
maximum possible performance.

# Tessellating A Plane { #sec:voronoi-def }

A **Voronoi diagram**, also known as a
**Dirichlet tessellation**, is an organization of a plane into
subsets defined by the nearest proximity to a given collection of
points.

From what I can tell the standard reference for Voronoi diagrams,
and other topics in computational geometry, is @deBergEtAl:2008.
This text is often referred to as "the 4Ms", which I can only
infer refers to the fact that three out of the four authors share
the first name "Mark" with the forth, non-Mark author being an
honorary "M".  I humbly suggest that "Mark, Marky, Mark, and the
Funky Bunch" is more appropriate, but to each their own.

To formally define a Voronoi diagram consider a two-dimensional
Euclidean plane $\mathbb{R}^{2}$ and a collection of **sites**
$$
( \, s_{1} = (x_{1}, y_{1}), \ldots, s_{N} = (x_{N}, y_{N}) \, )
\subset \mathbb{R}^{2}.
$$
Give a particular site $s_{n}$ we can construct a subset of
points that are closer to that site than any other site,
$$
\mathsf{c}_{n} =
\{ p \in \mathbb{R}^{2} \mid
d( p, s_{n} ) < d( p, s_{n'} ) \; \mathrm{for all} \; n' \ne n \},
$$
where $d(p, p')$ is a distance function.  These subsets are also
known as **cells**.  The Voronoi diagram of
$(s_{1}, \ldots, S_{N})$ is the corresponding collection of these
proximity-based subsets (@fig-voronoi).

::: {#fig-voronoi layout="[50, 50]"}
![](figures/voronoi/cells/voronoi){#fig-voronoi-cell}

![](figures/voronoi/boundary/voronoi){#fig-voronoi-boundary}

\(a) A Voronoi diagram defined by a collection of sites
$( s_{1}, \ldots, s_{N} )$, here shown in red, is a collection of
subsets, each containing points that are closer to one site than
the others.  (b) The boundary between these subsets consists of
points that are equally distant to two or more sites at the same
time.
:::

The standard Voronoi diagram considers the Euclidean distance
function,
$$
d( p, p' ) = \sqrt{ (x - x')^{2} + (y - y')^{2} }.
$$
In theory, however, this construction can be generalized to other
distance functions.  At the same time Voronoi diagrams can be
defined on higher-dimensional real spaces and even more general
non-Euclidean ometric spaces.

A Voronoi diagram _almost_ forms partition of a Euclidean plane.
The snag is that points that are equally distant from two or more
sites don't fall into any of the cells, but rather form a
singular boundary between the cells.  The combination of the
Voronoi cells and the boundary between them, however, does form
a proper partition.

Because each cell contains one, and only one, site, we can always
label the cells by the associated site (@fig-cell-label).
Similarly we can label the linear segment that forms the boundary
between two neighboring cells by the sites associated with those
two cells, and the points separating the corners of three or more
neighboring cells by the sites associated with those cells
(@fig-line-label, @fig-line-label).

::: {#fig-voronoi-labels layout="[-0.5, 33, 33, 33, -0.5]"}
![](figures/voronoi/cell_labels/labels){#fig-cell-label}

![](figures/voronoi/boundary_labels/line/labels){#fig-line-label}

![](figures/voronoi/boundary_labels/point/labels){#fig-point-label}

\(a) Each of the cells in a Voronoi diagram can be labeled with
the unique site that it contains.  (b, c) Similarly the
boundaries between neighboring cells can be labeled with the
sites contained by those cells.  For the purposes of labeling the
the order of the sites is arbitrary; here $(3, 2)$ is just as
good as $(2, 3)$ while $(2, 7, 3)$, $(3, 7, 2)$, $(3, 2, 7)$,
$(7, 3, 2)$, and $(7, 2, 3)$ are just as good as $(2, 3, 7)$.
:::

# The Geometry of Voronoi Diagrams

The boundary between the cells of a Voronoi diagram is useful in
its own right.  In particular it defines an undirected graph,
with linear edges separating neighboring pairs of cells and
punctual vertices separating neighboring triplets of cells.

In practice it is often easier to construct this
**Voronoi graph** and then derive the Voronoi cells from it
rather than trying to construct the Voronoi cells directly.
The vertices and edges of the Voronoi graph exhibit rich
geometric structure that provides the foundation for powerful
construction algorithms.

## Voronoi Vertices

The vertices of the Voronoi graph are defined by points that are
the same distance from three neighboring Voronoi cells
(@fig-vertex).  Consequently we can uniquely label each vertex
with a triplet the sites contained by those cells,
$$
(s_{n_{1}}, s_{n_{2}}, s_{n_{3}} ).
$$
Because not every triplet of Voronoi cells are neighbors,
however, not every triplet of sites defines a valid vertex.
Fortunately Euclidean geometry provides tools for distinguishing
the valid vertices.

![Each vertex in a Voronoi graph is given by a point on the
boundary between three neighboring Voronoi cells.  This point is
by definition equally distant to the sites contained by those
cells.](
figures/voronoi/vertex/vertex_distances/vertex){
width=50% #fig-vertex}

Any triplet of distinct points falls onto a unique
**circumcircle** with center
\begin{align*}
x_{c}
&=
\frac{1}{2}
\frac{
  \left( x_{n_{1}}^{2} + y_{n_{1}}^{2} \right) \,
    (y_{n_{2}} - y_{n_{3}})
+ \left( x_{n_{2}}^{2} + y_{n_{2}}^{2} \right) \,
    (y_{n_{3}} - y_{n_{1}})
+ \left( x_{n_{3}}^{2} + y_{n_{3}}^{2} \right) \,
    (y_{n_{1}} - y_{n_{2}})
}{
  x_{n_{1}} \, (y_{n_{2}} - y_{n_{3}} )
+ x_{n_{2}} \, (y_{n_{3}} - y_{n_{1}} )
+ x_{n_{3}} \, (y_{n_{1}} - y_{n_{2}} )
}
\\
y_{c}
&=
\frac{1}{2}
\frac{
  \left( x_{n_{1}}^{2} + y_{n_{1}}^{2} \right) \,
    (x_{n_{3}} - x_{n_{2}})
+ \left( x_{n_{2}}^{2} + y_{n_{2}}^{2} \right) \,
    (x_{n_{1}} - x_{n_{3}})
+ \left( x_{n_{3}}^{2} + y_{n_{3}}^{2} \right) \,
    (x_{n_{2}} - x_{n_{1}})
}{
  x_{n_{1}} \, (y_{n_{2}} - y_{n_{3}} )
+ x_{n_{2}} \, (y_{n_{3}} - y_{n_{1}} )
+ x_{n_{3}} \, (y_{n_{1}} - y_{n_{2}} )
}
\end{align*}
and radius
\begin{align*}
r
&=
\sqrt{ (x_{c} - x_{n_{1}})^{2} + (y_{c} - y_{n_{1}})^{2} }
\\
&=
\sqrt{ (x_{c} - x_{n_{2}})^{2} + (y_{c} - y_{n_{2}})^{2} }
\\
&=
\sqrt{ (x_{c} - x_{n_{3}})^{2} + (y_{c} - y_{n_{3}})^{2} }.
\end{align*}
Consequently the center of the circumcircle defined by three
sites is equidistant to those sites, and hence defines a
potential vertex in the Voronoi graph.

The three sites defining a circumcircle are the three _closest_
sites to this potential vertex if and only if no other sites are
inside of the circumcircle.  In other words Voronoi vertices are
defined by _empty_ circumcircles.

::: {#fig-vertex-circum layout="[-17, 33, 33, -17]"}
![](figures/voronoi/vertex/vertex_good/vertex_good){#fig-vertex-circum-good}

![](figures/voronoi/vertex/vertex_bad/vertex_bad){#fig-vertex-circum-bad}

Not every triplet of sites defines a vertex in a Voronoi graph.
(a) Vertices are defined by triplets of sites that define empty
circumcircles.  (b) Triplets of sites whose circumcircles contain
other sites do not define proper vertices.
:::

## Voronoi Edges

The edges in a Voronoi graph are comprised of points equidistant
to two neighboring cells (@fig-edge), allowing us to uniquely
label each Voronoi edge with the pair of sites contained by those
cells.  Once, again, however, we have to take care because
arbitrary pairs of sites do not always define valid edges.

![Each edge in a Voronoi graph is defined by linear boundary
between two neighboring Voronoi cells.  All of the points on a
Voronoi edge are equally distant to the two sites contained by
those cells.](
figures/voronoi/edge/edge_distances/edge){
width=50% #fig-edge}

Before worrying out isolating proper edges let's first work out
the geometry of potential edges.  Any pair of sites defines a
line that connects them and a mid point on that line that is
equally distant to the sites.  The **perpendicular bisector** of
two sites is the line perpendicular to this connecting line that
intersects with the midpoint (@fig-bisector).  Every point along
the perpendicular bisector between two sites is equally distant
from those two sites and, consequently, could contribute to a
Voronoi edge (@fig-bisector-edge).

::: {#fig-potential-edge layout="[-17, 33, 33, -17]"}
![](figures/voronoi/edge/bisector/bisector){#fig-bisector}

![](figures/voronoi/edge/bisector_edge/edge){#fig-bisector-edge}

\(a) The perpendicular bisector between two points is the line
$\ell_{\perp}$ that is perpendicular to line connecting those
points, $\ell$, and intersects the midpoint between them. (b)
Every Voronoi edge falls along a perpendicular bisector between
two sites, but generally is only a segment of that line.
Moreover not every perpendicular bisector contains a Voronoi
edge.
:::

In order to determine which segment, if any, of a perpendicular
bisector forms a Voronoi edge we come back to circles.  Any point
along a perpendicular bisector is the center of a unique circle
that includes the two defining sites.  If no _other_ sites are on
or inside of this circle then those two sites are the nearest
neighbors to each other and that point is, by definition, part of
a Voronoi edge.

As we move along the perpendicular bisector these circles will
eventually intersect with another site.  When they do the point
along the bisector will be equally distant to three sites -- the
two initial sites and this new site -- but the interior of the
circle will remain empty.  In other words the Voronoi edge
terminates as soon as we hit a Voronoi vertex.

::: {#fig-edge-circles layout="[-0.5, 33, 33, 33, -0.5]"}
![](figures/voronoi/edge/circle1/edge){#fig-edge-circle1}

![](figures/voronoi/edge/circle2/edge){#fig-edge-circle2}

![](figures/voronoi/edge/circle3/edge){#fig-edge-circle3}

Circles centered at points along a perpendicular bisector and
intersecting with a pair of sites are useful for identifying
contributions to Voronoi edges.  (a) If there are no other sites
on or inside of the circle then the center is part of a Voronoi
edge.  (b) If there is one or more sites inside of the circle
then the center does not contribute to a Voronoi graph at all.
(c) If there is another site on the circle then the center
defines a Voronoi vertex.  Each Voronoi edge terminates when
these circles first meet a third site.
:::

## Voronoi Faces

The edges of a Voronoi graph trace around each site, defining
faceted **faces** in the graph that correspond to the Voronoi
cells (@fig-faces).  When working over the entire, unbounded,
Euclidean plane not all of these faces will be closed.  While
some of the sites will be entirely enclosed by a path of edges,
some will be bounded on only one side by a path of edges. I will
refer to these as closed faces and open faces, respectively.

::: {#fig-faces layout="[-17, 33, 33, -17]"}
![](figures/voronoi/faces/closed/voronoi){#fig-closed-face}

![](figures/voronoi/faces/open/voronoi){#fig-open-face}

The edges of a Voronoi graph trace out faces which identify the
Voronoi cells.  (a) I will refer to faces that are completely
enclosed by edges as closed faces and (b) faces that are only
partially enclosed as open faces.
:::

# Fortune's Algorithm

There are here are multiple ways to construct a Voronoi graph,
but most of them suffer from wasteful computation.  For large
enough numbers of sites these approaches quickly become too
expensive to be practical.  Fortunately there is one elegant
approach that, while not at all initially obvious, is able to
construct a Voronoi graph as efficiently as possible.

## Inefficient Approaches

Before diving into an efficient solution let's consider some of
the more straightforward but ultimately inefficient approaches.

One way to identify all of the vertices in a Voronoi graph is to
loop over each triplet of sites and then construct the
corresponding circumcircles.  For each of these circumcircles we
then loop over the remaining sites to check for inclusion, with
the centers of the empty circumcircles defining vertices.

We could then construct candidate Voronoi edges by looping over
each pair of sites and checking to see if they are associated
with any vertices.  For each valid pair we could scan across the
corresponding perpendicular bisector, construct circles, and
check if any other sites appear on or within them.

Because there are
$$
{ N \choose 3 }
=
\frac{ N \, (N - 1) \, (N - 2) }{ 6 }
$$
triplets of sites and
$$
{ N \choose 2 }
=
\frac{ N \, (N - 1) }{ 2 }
$$
pairs of sites to consider, however, the number of operations we
need to implement this brute-force approach scales _cubically_
with the number of sites.  Adding twice as many sites requires
six times more operations, and hence six times the cost.

That said, much of this cost is wasted on the redundant queries
of each site.  We should be able to construct Voronoi graphs more
efficiently if we can query each site fewer times.  One more
efficient approach is to loop over each site, construct the
$(N - 1)$ half-planes bounded by the perpendicular bisector
between the active site and each other site, and then construct
the corresponding Voronoi face from the intersection of those
half-planes.  This cost of this approach scales only
quadratically.

We can, however, do even better.  By systematically scanning
across a plane **Fortune's algorithm**
[@Fortune:1987, @deBergEtAl:2008] is able construct a Voronoi
graph with a number of operations proportional to $N \log N$.
This performance turns out to be theoretically optimal.  The main
downside of Fortune's algorithm is that is requires some care to
understand and then implement in a way that achieves that optimal
scaling.

## The Sweep Line

Fortune's algorithm processes the sites not by working through a
list but rather by geometrically scanning across the ambient
plane with the use of a **sweep line**.  Sites on one side of
the sweep line are **active** and can contribute to the
construction of the Voronoi graph while sites on the other side
are **inactive** and cannot yet contribute.  As the sweep line
progresses more sites are processed and more of the Voronoi graph
is built up.

In theory a sweep line can progress in any direction, but most
references consider vertical lines that scan horizontally from
left to right, horizontal lines that scan vertically from top to
bottom, or horizontal lines that scan vertically from bottom to
top.  These axis-aligned orientations of the sweep line make
geometric calculations substantially more straightforward.

Here I will consider a vertical sweep line that scans
horizontally from left to right and denote the horizontal
position of the sweep line at any time by $S$.

## The Beach Line

A sweep line partitions the ambient plane into two subsets,
one containing all of the active sites and one containing all of
the inactive sites.  This allows us to focus the construction of
the Voronoi graph on just one subset of the entire plane and
ignore the other.  That said, not all of the Voronoi graph to the
left of the sweep line is completely determined by the active
sites; some features of the Voronoi diagram can be influenced by
inactive sites just beyond the sweep line.

Consider, for example, a single active site
$$
s_{n} = (x_{n}, y_{n})
$$
and the sweep line position
$$
S > x_{n}.
$$
If a point $(x, y)$ to the left of the sweep line is closer to
$s_{n}$ than the sweep line,
$$
(x - x_{n})^{2} - (y - y_{n})^{2} < (S - x)^{2},
$$
then we it will be closer to $s_{n}$ than any other site that
might be hiding beyond the sweep line.  On the other hand any
point to the left of the sweep line that is closer to the sweep
line than it is to $s_{n}$,
$$
(x - x_{n})^{2} - (y - y_{n})^{2} > (S - x)^{2},
$$
_could_ be be closer to an inactive site on the other side of the
sweep line that we have not yet taken into account.

Consequently when determining site proximity we cannot analyze
the entire region to the left of the sweep line.  Instead we can
analyze only the subset of points that are closer to $s_{n}$ than
they are to the sweep line.  The boundary of this analyzable
subset is defined by the points equidistant to $s_{n}$ and the
sweep line, which traces out the rotated parabola (@fig-parabola)
$$
x
=
f(y)
=
\frac{1}{2}
\left( S + x_{n} - \frac{( y - y_{n} )^{2}}{S - x_{n}} \right).
$$

![The subset of points that are equally distant from a focus,
$s$, and a vertical line $S$, form a rotated parabola.  In
Fortune's algorithm the points to the left of this parabola are
safe to analyze for their contribution to the Voronoi graph.](
figures/fortune/parabola_distances/parabola_distances){
width=50% #fig-parabola}

When there are multiple active sites then the region that we can
safely analyze is comprised of the points that are closer to
_any_ of the active sites than they are to the sweep line.  The
boundary of this region is defined by the segmented envelope of
the parabolas between each active site and the sweep line
(@fig-beachline).  Given its resemblance to the shape made by
waves washing up on a beach (@fig-actual-beachline) this boundary
is known as the **beach line**.

![The beach line is the boundary separating points that are
closer to any active site, here $s_{n_{1}}$, $s_{n_{2}}$, and
$s_{n_{3}}$, and points that are closer to the the sweep line,
$S$.  This boundary is comprised of a sequence of intersecting
parabolic arcs.  Note that the parabola between $s_{n_{1}}$ and
$S$ contributes multiple arcs to the beach line here.](
figures/fortune/beachline/beachline){
width=50% #fig-beachline}

![The parabolic arcs that make up a beach line are similar to the
shape made by waves washing up onto a beach, hence the name.
Photo :copyright: Michael Betancourt.](
figures/fortune/actual_beachline/actual_beachline){
width=100% #fig-actual-beachline}

Each parabolic segment, or **arc**, in a beach line is associated
with a single site.  The parabola between any given site and the
sweep line, however, can contribute to the beach line multiple
times.  Consequently we cannot use the originating sites to
_uniquely_ label the arcs.

Neighboring arcs on the beach line are generated from neighboring
sites.  The discontinuities along the beach line that separate
each pair of neighboring arcs are known as **breakpoints**
(@fig-breakpoints).  Each breakpoint is uniquely labeled by an
ordered pair of the sites associated with those arcs.  Note that
ordering is important here: $(s_{n_{1}}, s_{n_{2}})$ and
$(s_{n_{2}}, s_{n_{1}})$ correspond to two _different_
breakpoints.

![The breakpoints in a beach line separate the component arcs.
Each breakpoint is located at the intersection between the
parabolas generated by two neighboring sites, equally distant
from the sweep line and those two sites while also being further
from any other sites.  Each ordered pair of neighboring sites
defines a unique breakpoint.  Here, for example, $b_{n_{1}n_{2}}$
and $b_{n_{2}n_{1}}$ are distinct breakpoints.](
figures/fortune/breakpoints/breakpoints){
width=50% #fig-breakpoints}

Given a sweep line position $S$ we can readily calculate the
intersection of each pair of neighboring parabolas, and hence the
position of the corresponding breakpoint.  The vertical position
$y$ where the two parabolas generated by $s_{n_{1}}$ and
$s_{n_{2}}$ intersect is implicitly given by the quadratic
equation
\begin{align*}
0
&=\quad
\Bigg( \underbrace{ x_{n_{2}} - x_{n_{1}} }_{a} \Bigg) \,
y^{2}
\\
&\quad
-2 \,
\Bigg(  \underbrace{    y_{n_{2}} \, (S - x_{n_{1}})
                      - y_{n_{1}} \, (S - x_{n_{2}}) }_{b} \Bigg) \,
y
\\
&\quad
+
\Bigg(  \underbrace{   y_{n_{2}}^{2} \, (S - x_{n_{1}})
                     - y_{n_{1}}^{2} \, (S - x_{n_{2}})
                     - (S - x_{n_{1}}) \,
                       (S - x_{n_{2}}) \,
                       (x_{n_{2}} - x_{n_{1}} ) }_{c}
\Bigg),
\end{align*}
which admits two solutions,
$$
y_{\pm} = b \pm \frac{1}{2 \, a} \sqrt{ b^{2} - a \, c }.
$$

To determine which of these solutions is associated with the
breakpoint $b_{n_{1}n_{2}}$, and hence which is associated with
$b_{n_{2}n_{1}}$, we'll need to consider the relative position of
the sites.

If $s_{n_{1}}$ is ahead of $s_{n_{2}}$,
$$
x_{n_{1}} > x_{n_{2}}
$$
then the breakpoint $b_{n_{1}n_{2}}$ will be _above_ the
breakpoint $b_{n_{2}n_{1}}$ (@fig-bp-sol-case1).  In this case
$$
a = x_{n_{2}} - x_{n_{1}}
$$
is negative so the higher solution is given by
$$
y_{n_{1}n_{2}}
=
y_{-}
=
b - \frac{1}{2 \, a} \sqrt{ b^{2} - a \, c }.
$$

On the other hand if $s_{n_{1}}$ is behind $s_{n_{2}}$,
$$
x_{n_{1}} < x_{n_{2}}
$$
then the breakpoint $b_{n_{1}n_{2}}$ will be _below_ the
breakpoint $b_{n_{2}n_{1}}$ (@fig-bp-sol-case2).  Because the
leading coefficient is now positive,
$$
a = x_{n_{2}} - x_{n_{1}},
$$
the appropriate solution is _also_ given by
$$
y_{n_{1}n_{2}}
=
y_{-}
=
b - \frac{1}{2 \, a} \sqrt{ b^{2} - a \, c }.
$$

In other words the the sign of $a$ conveniently adjusts the
geometry correctly so that we can _always_ take
$$
y_{n_{1}n_{2}}
=
b - \frac{1}{2 \, a} \sqrt{ b^{2} - a \, c }
$$
and
$$
y_{n_{2}n_{1}}
=
b + \frac{1}{2 \, a} \sqrt{ b^{2} - a \, c }.
$$

::: {#fig-bp-sol layout="[50, 50]"}
![](figures/fortune/breakpoint_solution/1/solution){#fig-bp-sol-case1}

![](figures/fortune/breakpoint_solution/2/solution){#fig-bp-sol-case2}

The parabolas defined by two points $s_{n_{1}}$ and $s_{n_{2}}$
intersect at two different points, one of which gives the
breakpoint $b_{n_{1}n_{2}}$ and the other giving
$b_{n_{2}n_{1}}$.  (a) If $x_{n_{1}} > x_{n_{2}}$ then
$b_{n_{1}n_{2}}$ is given by the higher of the two intersections,
but (b) if $x_{n_{1}} < x_{n_{2}}$ then $b_{n_{1}n_{2}}$ is given
by the lower of the two intersections.
:::

When it comes to constructing Voronoi graphs the breakpoints are
the most important part of the beach line.  Because they always
falls onto the parabolas generated by two neighboring sites,
each breakpoint is equally distant from those two sites and the
sweep line.  At the same time each breakpoint is further from all
other active sites as well as any inactive sights hiding beyond
the sweep line.

All of this is to say that, by construction, every breakpoint
is the center of a circle containing the two associated sites but
includes no other sites.  In other words, each breakpoint _always
falls on a Voronoi edge no matter the position of the sweep line_.
Specifically the breakpoints $b_{n_{1}n_{2}}$ and
$b_{n_{2}n_{1}}$ both fall on the Voronoi edge between
$s_{n_{1}}$ and $s_{n_{2}}$.

## Evolving The Beach Line

As we scan the sweep line across a Euclidean plane the distances
between the active sites and the sweep line will change.  This,
in turn, changes the shape of the corresponding parabolas and,
ultimately, the shape of the beach line.  As the beach line
evolves the breakpoints trace out the Voronoi edges,
incrementally generating the entire Voronoi graph!

Every now and then the number of arcs comprising the beach line,
known as the **topology** of the beach line, also changes.  When
the sweep line passes over a new site, for example, a new arc is
added to the beach line.  Similarly arcs can be removed from the
beach line when they are squeezed to a point by their neighboring
arcs.

Conveniently we don't need to _continuously_ scan the sweep line
in order to construct a Voronoi diagram.  The structure of a
Voronoi diagram is completely determined by the behavior of the
beach line at the finite number of sweep line positions where the
topology changes.  If we can efficiently determine these
topology-changing **events** then we can efficiently build up a
Voronoi diagram in a finite number of operations.

### Events Where The Beach Line Expands

Whenever the sweep line crosses a new site, that site becomes
active and contributes a new parabolic arc to the beach line.
Because we know the position of all of the sites we know at
exactly which positions these **site events** will occur.

If the sweep line is positioned on a new site then the parabola
of points equidistant from that site and the sweep line is
singular, consisting of just the new site itself.  This forms a
point discontinuity with the rest of the beach line
(@fig-site-event1a).  Although not _technically_ correct, these
singular arcs are often visualized with a horizontal line
extending from the new site to the beach line
(@fig-site-event1b).

::: {#fig-site-event layout="[-0.5, 33, 33, 33, -0.5]"}
![](figures/fortune/site_event/1a/site_event){#fig-site-event1a}

![](figures/fortune/site_event/1b/site_event){#fig-site-event1b}

![](figures/fortune/site_event/2/site_event){#fig-site-event2}

At a site event the sweep line hits an inactive site which then
becomes active.  (a) Initially the new active site introduces a
new singular arc which splits the beach line.  (b) This singular
arc is often visualized with a continuous horizontal line.  (c)
As the sweep line proceeds past the site event, this singular arc
widens into a well-behaved parabola, separating one of the
beach line arcs into two arcs.
:::

As the sweep line progresses past the new active site, the
singular parabola widens into a more well-behaved parabola
(@fig-site-event2).  This not only restores the continuity of the
beach line but also splits an existing parabolic arc into two new
arcs.  The introduction of a new active site always splits an
existing arc into a triplet of arcs, the outer two of which are
generated by the same site.

For example let's say that the sweep line passes the new site
$s_{n_{2}}$.  If the height of this site $y_{n_{2}}$ intersects
with an arc generated by the site $s_{n_{1}}$ then the site event
will split that arc into a triplet of arcs, the outer two of
which are generated by $s_{n_{1}}$ and the center of which is
generated by $s_{n_{2}}$. These three arcs are then separated by
two new breakpoints $b_{ n_{1} n_{2} }$ and $b_{ n_{2} n_{1} }$.

Both of these new breakpoints fall on a Voronoi edge in between
the sites $s_{n_{1}}$ and $s_{n_{2}}$.  As the sweep line
progresses the middle arc widens and the breakpoints separate,
tracing out more and more of that edge
(@fig-site-event-breakpoints).

![As the sweep line moves past a site event, the two new
breakpoints separate and scan across the Voronoi edge between the
two neighboring sites.](
figures/fortune/site_event/breakpoints/breakpoints){
width=50% #fig-site-event-breakpoints}

### Events Where The Beach Line Contracts { #sec:contract }

The topology of the beach line can also change when one of the
parabolic arcs collapses in between its two neighboring arcs
(@fig-vertex-event-collapse).  In this case the collapsed arc is
removed from the beach line and its two neighboring arcs become
new neighbors.

![As the sweep line progresses, arcs on the beach line can
collapse in between their two neighboring arcs.  When this
happens the neighboring breakpoints converge to a point.  This
convergence point is equally distant from the three active sites
but further from all other sites, defining a vertex in the
Voronoi graph.](
figures/fortune/vertex_event/collapse/vertex_event){
width=50% #fig-vertex-event-collapse}

Consider three neighboring arcs generated from the active sites
$s_{n_{2}}$, $s_{n_{2}}$, and $s_{n_{3}}$.  If any two of these
sites are the same then the central arc will not collapse as the
sweep line progresses so we can safely assume that these sites
are distinct.

When the central arc collapses its two neighboring breakpoints
$b_{ n_{1} n_{2} }$ and $b_{ n_{2} n_{3} }$ will merge into a
new breakpoint $b_{ n_{1} n_{3} }$.  At that moment
$$
b_{ n_{1} n_{2} } = b_{ n_{2} n_{3} } = b_{ n_{1} n_{3} }
$$
are equally distant from $s_{n_{2}}$, $s_{n_{2}}$, and
$s_{n_{3}}$.  Because this intersection is on the beach line it
also cannot be closer to any inactive sites hiding beyond the
sweep line.  In other words a collapsing arc on the beach line
always defines a Voronoi vertex!

This duality between arc collapses and Voronoi vertices is useful
not only for building up the Voronoi graph but also for
predicting when arcs will collapse.  By now it should be no
surprise that this prediction will involve circles.

For any triplet of neighboring arcs that correspond to the
distinct sites $s_{n_{1}}$, $s_{n_{2}}$, and $s_{n_{3}}$ we can
always construct a unique circumcircle with center
$(x_{c}, y_{c})$ and radius $r$.  By construction the center
$(x_{c}, y_{c})$ is equally distant to all three sites, and hence
identifies exactly where the central arc will collapse to a
point.

The center is also equally distant to vertical lines at
$x = x_{c} - r$ and $x = x_{c} + r$, which define the two
possible sweep line configurations where the central arc will
collapse.  When $S = x_{c} - r$, however, the sweep line will not
yet have reached any of the three sites and they will not yet
have contributed any arcs to the beach line.  Consequently a
central arc generated by $s_{n_{2}}$ can collapse only when
$S = x_{c} + r$ (@fig-vertex-event-circumcircle).

![If the central arc in a triplet of neighboring arcs collapses
then it will do so when the sweep line is at the far edge of the
circumcircle spanned by those three sites.  This geometric
extrapolation allows us to schedule future vertex events.](
figures/fortune/vertex_event/circumcircle/vertex_event){
width=50% #fig-vertex-event-circumcircle}

That said, this collapse can contribute to the Voronoi graph only
if it occurs to the right of the beach line, where the structure
of the Voronoi graph has not yet been established.  If $x_{c}$
falls to the left of the current beach line then advancing the
sweep line forward will cause the central arc to not collapse but
rather expand, with the breakpoints _diverging_ away from each
other.

One particularly straightforward way to determine if a central
arc will collapse as the sweep line progresses forward is to
directly check if $y_{ n_{1} n_{2} } = y_{ n_{2} n_{3} }$ at
$S = x_{c} + r$.  While this works fine in theory, it can be
awkward to implement in practice as checking the equality of two
floating point numbers is always tricky.

We can also use the relative positions of $s_{n_{1}}$,
$s_{n_{2}}$, and $s_{n_{3}}$ to determine whether some arcs will
converge or diverge.  For example if site $s_{n_{2}}$ is _behind_
both $s_{n_{1}}$ and $s_{n_{3}}$,
\begin{align*}
x_{n_{2}} &< x_{n_{1}}
\\
x_{n_{2}} &< x_{n_{3}},
\end{align*}
then the Voronoi edges, and the breakpoints that flow along them,
will converge (@fig-vertex-convergence1).  On the other hand if
$s_{n_{2}}$ is _ahead_ of the other two sites,
\begin{align*}
x_{n_{2}} &> x_{n_{1}}
\\
x_{n_{2}} &> x_{n_{3}},
\end{align*}
then the Voronoi edges will diverge, pulling the breakpoints
away from each other (@fig-vertex-convergence2).

The mixed cases where
$$
x_{n_{1}} < x_{n_{2}} < x_{n_{3}}
$$
or
$$
x_{n_{1}} > x_{n_{2}} > x_{n_{3}}
$$
are indeterminate
(@fig-vertex-convergence3, @fig-vertex-convergence4).  Depending
upon the vertical positioning of the sites the central arc can
collapse or expand as the sweep line progresses.  The behavior in
these cases can probably be worked out with some clever geometry,
but I have yet not been able to figure out that geometry myself.
Consequently I will use the more direct approach mentioned above.

::: {#fig-vertex-convergence layout="[ [-17, 33, 33, -17], [-17, 33, 33, -17]] ]"}
![](figures/fortune/vergence/convergence/voronoi){#fig-vertex-convergence1}

![](figures/fortune/vergence/divergence/voronoi){#fig-vertex-convergence2}

![](figures/fortune/vergence/mixed1/voronoi){#fig-vertex-convergence3}

![](figures/fortune/vergence/mixed2/voronoi){#fig-vertex-convergence4}

The central arc of any triplet of neighboring beach line arcs,
here generated by the sites $s_{n_{1}}$, $s_{n_{2}}$, and
$s_{n_{3}}$, will not always collapse as the sweep line
progresses forward.  (a) If $s_{n_{2}}$ is behind the other two
sites then the arc it generates will eventually collapse, but (b)
if $s_{n_{2}}$ is ahead of the other two sites then the arc it
generates will only ever expand.  (c, d) When $s_{n_{2}}$ is in
between the two other sites we need additional information to
determine if the corresponding arc will collapse.
:::

If a central arc will eventually collapse as the sweep line
progresses then that collapse defines **vertex event**, also
known as a **circle event**, at $S = x_{c} + r$.  At this sweep
line position the breakpoints $b_{ n_{1} n_{2} }$ and
$b_{ n_{2} n_{3} }$ will intersect at $(x_{c}, y_{c})$, defining
a new vertex that we can add to the Voronoi graph
(@fig-vertex-event-voronoi).  This vertex then spawns a new
Voronoi edge along which the new merged breakpoint
$b_{ n_{1} n_{3} }$ will flow.

![When a central arc collapses, two Voronoi edges terminate at a
Voronoi vertex and a new Voronoi edge between the sites that
generate the remaining arcs is created.](
figures/fortune/vertex_event/voronoi/voronoi){
width=50% #fig-vertex-event-voronoi}

## The Algorithm

Now that we know how and when the topology of a beach line can
change we can organize those changes into an algorithm that
builds up a Voronoi graph dynamically.

Fortune's algorithm starts with an **event queue** consisting of
events defined by sweep line positions where an arc is added to
or removed from the beach line.  The algorithm then proceeds
iteratively, processing each event in the queue one at a time,
expanding the structure of the Voronoi graph, and adding new
events as necessary.  Initially the event queue is filled with
only site events, one for each site, but vertex events are also
added as the site events are processed.

In this section we'll review the basic steps required to process
both site and vertex events, as well as clean up after the event
queue has been fully processed.  [Section 4](@sec:implementation)
will focus on the implementation of these steps, which becomes a
good bit more complicated if we want to ensure efficient
computation.

### Site Events

Processing a site event requires adding a new arc to the beach
line.  To do this we have to scan across the current beach line
to find the existing arc that meets the height of the new event.
We then split that arc in two, inserting a new arc generated by
the new site into the beach line along with two new, neighboring
breakpoints.

Inserting a new arc into the beach line can also modify the
remaining event queue.

For one the new arc will disrupt any triplets of neighboring
arcs that contain the split arc.  If any of those disrupted arc
triplets correspond to future vertex events then they have to be
removed from the event queue.

At the same time inserting a new arc also creates two triplets of
neighboring arcs, each of which could potentially spawn a vertex
event.  If the central arc of any of these triplets eventually
collapses then we need to add a new vertex event to the event
queue where the central arc will be removed.

To make this all a bit more concrete consider an initial beach
line comprised of three successive arcs, $\alpha_{1}$,
$$
( \alpha_{1}, \alpha_{2}, \alpha_{4}, \alpha_{2}, \alpha_{3}).
$$
When the sweep line then passes a new site it introduces a new
arc $\alpha_{4}$.  Let's say that this arc splits$\alpha_{2}$,
resulting in the new beach line
$$
( \alpha_{1}, \alpha_{2}, \alpha_{4}, \alpha_{2}, \alpha_{3}).
$$

This site event breaks the initial triple
$( \alpha_{1}, \alpha_{2}, \alpha_{3})$ and invalidates any
associated vertex events.  It also, however, introduces two new
triples, $( \alpha_{1}, \alpha_{2}, \alpha_{4})$ and
$( \alpha_{4}, \alpha_{2}, \alpha_{3})$, which have to be
considered for new vertex events.

### Vertex Events

Our first task when we encounter a vertex event in the event
queue is to expand the current Voronoi graph.  This starts
by adding a new Voronoi vertex, connecting it to two of the
previously existing Voronoi edges, and then creating a new
Voronoi edge.  Exactly how we accomplish this expansions depends
on how the graph itself is being implemented.

Once the Voronoi graph has been updated we then remove the
central arc from the beach line and update the breakpoints
appropriately.

Lastly we have to clean up the event queue.  The removal of the
central arc will invalidate any vertex event in the event queue
where the central arc was previously neighbors with the removed
arc.  On the other hand the removal also creates two new triplets
of neighboring arcs along the beach line, each of which could
spawn a new vertex event.  If any of these triplets feature a
collapsing central arc then we have to add a corresponding vertex
event to the event queue.

Again to make this a bit more concrete let's consider an initial
beach line comprised of the five successive arcs,
$$
( \alpha_{1}, \alpha_{2}, \alpha_{3}, \alpha_{4}, \alpha_{5}).
$$
At a vertex event the central arc $\alpha_{3}$ collapses,
resulting in the new beach line
$$
( \alpha_{1}, \alpha_{2}, \alpha_{4}, \alpha_{5}).
$$.

This vertex event breaks two initial triples,
$( \alpha_{1}, \alpha_{2}, \alpha_{3} )$ and
$( \alpha_{3}, \alpha_{4}, \alpha_{5} )$, invalidating any
associated vertex events still in the event queue.  At the same
time the deletion of $\alpha_{3}$ introduces two new triples,
$( \alpha_{1}, \alpha_{2}, \alpha_{4} )$ and
$( \alpha_{2}, \alpha_{4}, \alpha_{5} )$, each of which could
define new vertex events.

### Clean Up

As we process events the queue will eventually empty.  When there
are no more events to process the algorithm will have found all
of the Voronoi vertices and all of the Voronoi edges.  That said
some of those edges will be dangling, anchored to a vertex on
only one side because they stretch out to infinity in the other
direction (@fig-clean-up1).

::: {#fig-clean-up layout="[-0.5, 33, 33, 33, -0.5]"}
![](figures/fortune/cleanup/dangling/voronoi){#fig-clean-up1}

![](figures/fortune/cleanup/bounding_box/voronoi){#fig-clean-up2}

![](figures/fortune/cleanup/pseudo_vertices/voronoi){#fig-clean-up3}

\(a) After running Fortune's algorithm some of the edges in the
Voronoi graph will be connected to vertices on only one side.
The other side extends out towards infinity.  (b) When
visualizing a Voronoi graph in practice we can consider only a
finite range of values.  The best we can do is visualize the
graph within a finite bounding box that contains all of the
Voronoi vertices.  (c) For visualization purposes we can anchor
any dangling edges to points where they meet the bounding box.
:::

We can handle these dangling edges in a few different ways.  For
example from a graph theory perspective we could anchor any
dangling edges to a single "vertex at infinity" that quantifies
the common unboundedness of those edges.  This approach, however,
does not yield particularly compelling visualizations.

For visualization purposes it is more useful to create a
rectangular **bounding box** that contains all of the Voronoi
vertices (@fig-clean-up2), and then anchor the dangling edges to
points along this rectangle (@fig-clean-up3).  Specifically the
new anchor for a dangling edge will be the point where the
perpendicular bisector between the two associated sites
intersects with the bounding box.

# Implementing Fortune's Algorithm { #sec:implementation }

Once we've built up enough familiarity with planar geometry, the
motivation for Fortune's algorithm, particularly the utility of a
beach line and its evolution, becomes less obscure.  That said,
there is still work to do to ensure that we can implement the
algorithm in a way that achieves its maximum possible
performance.

For example when processing a site event we need to be able to
efficiently search through the arcs on the current beach line.
If we naively scan through the _entire_ beach line every time we
process a new site event then the the number of operations needed
to implement Fortune's algorithm with scale with not $N \log N$
but rather $N^{2}$.  To achieve that $N \log N$ scaling we need
to implement the beach line in a way that allows for efficient
searching.

Similarly we need to keep track of the next event while we add
and delete events from the event queue.  We could accomplish this
by sorting the entire event queue after every update, but this
is wasteful and also spoils the optimal $N \log N$ scaling.

Fortunately (pun very much intended) if we leverage some
standard data structures from computer science then we can avoid
these potential slowdowns and implement Fortune's algorithm as
efficiently as possible.

## Implementing Voronoi Graphs

In theory graphs can be implemented with a just list of points
for each vertex and list of pairs of points for each edge.  While
these lists allow us to draw a graph easily enough, they don't
provide enough information to perform basic graph operations,
such as tracing along paths of edges to derive faces.

The **doubly-connected edge list** [@deBergEtAl:2008] is a data
structure that efficiently captures not only the components of a
graph but also the relationships between those components.  These
relationships in turn make it straightforward to efficiently
implement graph algorithms.

A doubly-connected edge list is itself built up from lists of
three primitive data structures: half-edges, vertices, and faces.

### Half-Edges

Doubly-connected edge lists don't actually represent undirected
edges.  Instead the data structure splits each undirected edge
into two directed edges known as **half-edges**
(@fig-dcel-half-edge).

![Doubly-connected edge lists represent undirected edges with a
pair of directed half-edges.  Each half-edge includes references
to other structures in the doubly-connected edge lists, including
the vertex from which the half-edge originates and neighboring
half-edges.](
figures/dcel/half_edge/1/edge){
width=50% #fig-dcel-half-edge}

To situate them within the full graph, each half-edge is endowed
with a collection of references to other objects in the
doubly-connected edge list.  For example each half-edge includes
a reference to the vertex from which the half-edge originates.
Similarly each half-edge includes references to three other
half-edges -- the _previous_ half-edge that points into it, the
_next_ half-edge that it points into, and the _twin_ half-edge
that completes it to form a full undirected edge
(@fig-dcel-half-edge-refs).  Finally each half-edge includes a
reference to the face located to its left.

![Each half-edge in a doubly-connected edge lists includes
reference three other half-edges.  The previous half-edge, here
labeled "he.prv", identifies the half-edge that points into it
while the next half-edge, here labeled "he.nxt",  identifies the
half-edge that it points into.  Finally the twin half-edge, here
labeled "he.twin", identifies the half-edge that completes it to
form a full undirected edge.](
figures/dcel/half_edge/2/edge){
width=50% #fig-dcel-half-edge-refs}

These references make a variety of graph operations
straightforward to implement.  For instance the originating
vertex and twin half-edge references provide enough information
to draw a half-edge as a directed line segment, starting at the
position of the originating vertex and then ending at the
position of the twin's originating vertex.

Similarly we can use the previous and next references to trace
through paths in the graph.  These paths can then be used to
quantify topological information about the graph, such as the
configuration of the faces.

Voronoi graphs generally feature unbounded edges, with only one
side terminating at a vertex and the other shooting off towards
infinity.  In this case some of the half-edges will feature
empty next half-edge references while their twins will feature
empty originating vertex and previous half-edge references.

### Vertices

Each vertex in a doubly-connected edge list is specified with its
spatial position and a reference to any one half-edge that
originates from it (@fig-dcel-vertex).  The relationship of a
vertex to the rest of graph can then be completely reconstructed
from this information.

![Each vertex in a doubly-connected edge list includes its
spatial position, here denoted "v.pos" and a reference to any one
half-edge that originates from it, here denoted "v.he".](
figures/dcel/vertex1/vertex){
width=50% #fig-dcel-vertex}

For example all of the half-edges that originate at a given
vertex can always be accessed by repeatedly following the twin
and next references from the one included half-edge
(@fig-dcel-vertex-algo).  Equivalently we can iteratively follow
the previous and then twin references.  In the same way we can
also access the half-edges that _terminate_ at a given vertex.

::: {#fig-dcel-vertex-algo layout="[ [50, 50], [50, 50] ]"}
![](figures/dcel/vertex2/1/vertex){#fig-dcel-vertex-algo1}

![](figures/dcel/vertex2/2/vertex){#fig-dcel-vertex-algo2}

![](figures/dcel/vertex2/3/vertex){#fig-dcel-vertex-algo3}

![](figures/dcel/vertex2/4/vertex){#fig-dcel-vertex-algo4}

\(a) The one half-edge referenced in a vertex object is enough to
reconstruct the entire structure of the graph around that vertex.
For example if we follow the reference to (b) the previous
half-edge, "v.he.p", (c) and then its twin, "v.he.p.t", we
recover another half-edge that originates from that vertex.
(d) Repeating this operation recovers all of the half-edges that
originate from that vertex.
:::

Once we can access all of the neighboring half-edges we can then
access additional information about the local structure of the
graph.  Neighboring vertices, for instance, are given by the
originating vertex references of the twin of any half-edge that
originates at the initial vertex.

### Faces

Finally each face in the graph is specified by any one half-edge
along its boundary (@fig-dcel-face1).  This provides enough
information to trace out the entire face boundary by following
the next or previous references from that one included half-edge
(@fig-dcel-face2, @fig-dcel-face3).

::: {#fig-dcel-face layout="[-0.5, 33, 33, 33, -0.5]"}
![](figures/dcel/face1/face){#fig-dcel-face1}

![](figures/dcel/face2/face){#fig-dcel-face2}

![](figures/dcel/face3/face){#fig-dcel-face3}

\(a) Each face in a doubly-connected edge list includes a
reference to any one half-edge along its boundary.  By following
the next and previous references from that one half-edge we can
always trace out the entire boundary of the face. (b) If the
face is closed then we can follow _either_ the next or previous
references until we return to the initial half-edge.  (c)  If the
face is open then we have to follow _both_ the next and previous
half-edges until they we encounter an empty reference.
:::

When working with more complex graphs, such as graphs containing
disconnected faces or faces that are completely surrounded by
other faces, we need to include additional half-edge references,
such as one along the exterior boundary and one along each
internal boundary.  Because Voronoi graphs do not exhibit these
topological features, however, we won't need to worry about this
additional information here.

### Satellite Data

Depending upon the application the half-edge, vertex, and face
data structures can also include additional information in the
form of **satellite data**.  For example when using a
doubly-connected edge list to represent a Voronoi graph it will
be helpful to include information about the associated sites,
three for each vertex object and one for each face object.  Note
that we don't need to store the two sites associated with an
edge; this information can be recovered from the sites of the
left faces of the corresponding half-edges.

### Bounding Box and Pseudo-Vertices

The visualization of Voronoi graphs is easier if we add a few
more objects to the doubly-connected edge list data structure.
In particular, a natural addition is the configuration of a
bounding box as well as **pseudo-vertices** along that bounding
box that can be used to terminate any dangling half-edges.  This
ensures, for example, that every half-edge includes a non-empty
originating vertex references which in turns makes visualization
functions more uniform.

## Implementing Beach Lines

A beach line consists of parabolic arcs and the breakpoints that
separate them.  Critically these objects exhibit a consistent
_ordering_.

As we work through Fortune's algorithm we will need to be able to
efficiently search through these ordered objects.  Moreover, we
need to be able to add and remove arcs and breakpoints without
disrupting the ordering.  Conveniently all of these operations
can be implemented with a slightly modified
**binary search tree** [@CormenEtAl:2022] that lacks an explicit
key.

Many references, in particular @deBergEtAl:2008, recommend a
binary search tree with two types of nodes: one to represent arcs
and one to represent the breakpoints.  Unfortunately this
representation introduces some awkward implementation problems
that are poorly documented.  Your humble author has not yet been
able to work these issues out.

In this note I will adopt a different representation inspired by,
although slightly different from, the approach taken in
[pvigier's blog](https://pvigier.github.io/2018/11/18/fortune-algorithm-details.html).  Here beach lines will be
implemented with a binary search tree where every node represents
an arc and its lower breakpoint paired together (@fig-bst).

::: {#fig-bst layout="[ [-25, 50, -25], [100], [100] ]"}
![](figures/bst/beachline/beachline){#fig-bst1}

![](figures/bst/beachline_arcs/arcs){#fig-bst2}

![](figures/bst/beachline_bst/bst){#fig-bst3}

\(a) Every beach line (b) can be decomposed into a collection of
ordered arc and lower breakpoint pairs.  (c) These ordered
arc/breakpoint pairs can then implemented with a key-less
binary search tree data structure.
:::

Beyond a reference to their parent and child nodes in the tree,
each nodes contains satellite data that allows us to reconstruct
the local structure of the beach line for any sweep line
configuration, and hence implement the sequential updates in
Fortune's algorithm.

### Node Data

Each node in a beach line tree includes a reference to the site
that generates the corresponding arc.  Because the parabola
generated by a single site can contribute to the beach line
multiple times, multiple nodes can share the same site reference.

Once we know the generating site we can dynamically compute the
shape of the arc for any position of the sweep line.  If we also
know the position of the neighboring breakpoints we can then
plot the full extent of the arc in the beach line.

To that end the site information is complemented with a reference
to the Voronoi half-edge that originates from the lower
breakpoint.  From that reference we can derive the sites
associated with the Voronoi edge, in particular from the left
face of the half-edge and the left face of its twin.  In turn
those sites allow us to compute the intersection of the
corresponding parabolas, and hence the position of the lower
breakpoint.

Because the first arc in a beach line is always unbounded below,
it does not feature a lower breakpoint.  Consequently the
smallest node in any beach line tree will always feature an
empty half-edge reference.  Every other node, however, will
feature a non-trivial reference.

As the sweep line progresses the half-edges associated with each
breakpoint will not yet have an originating Voronoi vertex.  For
visualization purposes we can always display these half-edges as
if they originated from the current position of the corresponding
breakpoint (@fig-bst-edges).

![Initially the half-edges stored in each node are not anchored
to any Voronoi vertices.  When visualizing the progression of
Fortune's algorithm we can instead anchor each half-edge to the
current position of the corresponding breakpoint.  This ensures
that the half-edges always span at least a segment of full
Voronoi edge, shown here in dark teal.](
figures/bst/bp_edges/edges){
width=50% #fig-bst-edges}

Finally, to avoid repeated searches through the event queue each
node also includes a reference to any vertex events that are
centered on the corresponding arc.  This allows us to, for
example, quickly find events that need to be removed from the
event queue when an arc is removed from the beach line.

### Tree Operations

The key difference (absolutely an intentional pun) between these
beach line trees and generic binary search trees is that the
latter don't feature a traditional key.  Ordering of the nodes in
a beach line tree is determined not by static key values but
rather by the relative vertical position of the arcs and
breakpoints along the corresponding beach line.  As the sweep
line progresses the absolute positions change but the relative
positions, and hence the structure of the binary tree, remains
invariant.

Beach line trees share most of the same operations as standard
binary search trees, such as predecessor, successor, and deletion
operations.  Without static keys, however, beach line trees do
not feature traditional key search or key insertion operations.
In order to implement Fortune's algorithm we need to introduce
slight different versions of these operations.

For example Fortune's algorithm requires an **arc search** where
we find the arc in a beach line that intersects with a particular
vertical position $y$ when the sweep line is at position $S$.  To
find the corresponding arc node in a beach line tree we can
modify the standard key search operation, comparing $y$ to
dynamically-computed breakpoint heights at each step instead of
keys (@fig-bst-tree-search).

::: {#fig-bst-tree-search layout="[ [100] ]"}
![](figures/bst/tree_search/1/bst){#fig-bst-tree-search1}

![](figures/bst/tree_search/2/bst){#fig-bst-tree-search2}

Unlike the nodes in standard binary search trees, the nodes in
beach line trees are not equipped with static keys.  Given a
sweep line position, however, we can search for positional
intersections.  (a) To find the arc that intersects $y = -1.931$
we start at the root node.  We first compute the height of the
lower breakpoint for the current sweep line position, here
$-0.517$, and then move to the left child because the breakpoint
height is above $y$.  If $y$ were above the lower breakpoint
height then we would compute the current height of the upper
breakpoint and move to the right child if it is below $y$.
Otherwise we would terminate the search.  (b, c) We then repeat
this process, progressing down the tree until the search
terminates.  Here we terminate after one iteration.
:::

Once we find an appropriate node we then need to be able to
insert nodes immediately before or after it, expanding the beach
line with a new arc and breakpoint.  Fortunately these
**insert-before** and **insert-after** operations are
straightforward to implement (@fig-bst-insert-above).

For example to insert a new node after an existing node we check
to see if it has a right child.  If it does not then we can
insert the new node as the existing node's right child.
If there is a right child then we find the successor to the
existing node, which will always have no left child otherwise it
would not be a proper successor.  Then we can insert the new node
in that empty left child position.

::: {#fig-bst-insert-above layout="[ 50, 50 ]"}
![](figures/bst/insert_above/1/insert){#fig-bst-insert-above1}

![](figures/bst/insert_above/2/insert){#fig-bst-insert-above2}

Without static keys beach line trees do not support key insertion
operations.  They do, however, allow for insertion of a new node,
here shown in light teal, before or after a given node, here
shown in dark teal.  (a) For example the insert-after operation
begins by checking if the right child of the given node is empty
and, if it is, then inserts the new node as a new right child.
(b) If the right child is occupied then the insert-after
operation moves to the successor node.  Because the left child of
the successor will always be empty if the initial right child is
not, the operation can safely insert the new node there.  The
insert-below operation is symmetric, only moving backwards
instead of forwards.
:::

### Balancing

The number of operations needed to implement all of these beach
line tree operations scales with the height of a particular tree.
Now the height of most trees scales logarithmically with the
number of nodes, with in the context of Fortune's algorithm also
scales logarithmically with the number of sites.  Consequently
for most trees these operations require only a logarithmic number
of operations, and a _typical_ evaluation of Fortune's algorithm
will achieve an overall $N \log N$ cost scaling.

Unfortunately the height of _every_ possible binary tree is not
always logarithmic in the number of nodes.  In the worst case the
height of a tree can be linear in the number of nodes, pushing
up the overall cost scaling of Fortune's algorithm to quadratic.

In order to realize the optimal performance of Fortune's
algorithm we need to be able to avoid excessively tall, or
**unbalanced** trees.  Conveniently this can be accomplished with
the use of **self-balancing binary trees**, in particular
**red-black trees** [@CormenEtAl:2022].  These data structures
modify the standard insertion and deletion operations to ensure
that nodes are well-distributed across the entire tree.

With the use of red-black trees and balanced insertion and
deletion operations we can ensure $N \log N$ cost scaling for
any evaluation of Fortune's algorithm.

## The Event Queue

The event queue in Fortune's algorithm can be efficiently
implemented with a **priority queue** data structure
[@CormenEtAl:2022].  Priority queues are equipped with an
operation for "popping" the next element off of the queue as well
as operations for inserting and deleting elements while
maintaining the overall order of the elements.

To implement Fortune's algorithm we'll need to consider two types
of events.  Vertex events contain the position of the sweep line
where an arc will collapse and a reference to the node in the
beach line tree that represents that arc.  Site events, on the
other hand, contain the position where sweep line will encounter
a new site as well as a reference to that site.

## Event Processing

Fortune's algorithm proceeds by popping the next event off of
the event queue, using that event to advanced the sweep line, and
then updating the Voronoi graph as necessary.  With the
operations of doubly-connected edge lists and binary search
trees, processing each event is relatively straightforward.

### Processing Site Events

Processing a site event begins by using the arc search operation
of the beach line tree to find the arc that intersects with
the height of the new site by searching through the beach line
tree.

Because splitting this arc will invalidate any existing vertex
event centered on it, we have to remove any corresponding
vertex event from the event queue.  This is straightforward given
the vertex event reference held in each node.

As the sweep line passes the new site, the intersecting arc will
split into three arcs, with two new breakpoints between them
(@fig-bst-site-event1, @fig-bst-site-event2).  This expansion is
implemented by creating two new nodes, one associated with the
new site and one with the site generating the existing arc, and
then inserting them into the beach line tree with two
balanced insert-after operations (@fig-bst-site-event3).

::: {#fig-bst-site-event layout="[ [100], [100], [100] ]"}
![](figures/bst/site_event/beachlines/site_event){#fig-bst-site-event1}

![](figures/bst/site_event/arcs/site_event){#fig-bst-site-event2}

![](figures/bst/site_event/bsts/site_event){#fig-bst-site-event3}

\(a) A site event splits one of the arcs in the initial beach
line into three arcs, with two new breakpoints.  (b)  This
corresponds to two new arc-breakpoint pairs being introduced.
(c) The addition of the two new arc-breakpoint pairs is readily
implemented by twice applying the insert-after operation to the
initial beach line tree.
:::

The half-edges of these two new nodes are always twinned to each
other (@fig-site-event-edges).  Together the half-edges define a
new Voronoi edge associated with the old and new sites.

![The introduction of a new site introduces a new Voronoi edge
which is implemented with two new half-edges in the Voronoi graph
that are twinned with each other.  A reference to each of these
half-edges is stored in the the two new nodes added to the
beach line tree.  For example here a reference to the half-edge
`he1` is stored in the node $(b_{n_{6} n_{2}}, s_{n_{2}})$ while
a reference to `he2` is stored in the node
$(b_{n_{2} n_{6}}, s_{n_{6}})$.](
figures/fortune/edges/site/edges){
width=50% #fig-site-event-edges}

Lastly incorporating the new site creates up to two new triplets
of neighboring arcs that might generate new vertex events.
Fortunately the nodes in the beach line tree representing these
new triplets can be readily accessed with the predecessor and
successor beach line operations.

### Processing Vertex Events

The first step when processing a new vertex event is to remove
any other vertex events that will become invalid after we remove
the collapsing arc.  This can be done efficiently by using the
predecessor and successor operations of the beach line tree to
find the neighboring arcs, and then removing any vertex events
referenced in those arcs.

Next we manage the expansion of the Voronoi graph.  We begin by
adding a new vertex to the doubly-connected edge list at the
point where the collapsing arc becomes singular and the
neighboring breakpoints intersect.

Then we have to deal with the Voronoi edges.  We create a new
half-edge originating at the new vertex and its twin that, for
the moment, is left unanchored (@fig-vertex-event-edges).  Lastly
we have to carefully connect the previous and next references
between these two new half-edges, the two half-edges associated
with the neighboring breakpoints, and their twins
(@fig-vertex-event-edge-connections).

::: {#fig-vertex-event-edges layout="[ [50, 50], [-25, 50, -25] ]"}
![](figures/fortune/edges/vertex/1/edges){#fig-vertex-event-edges1}

![](figures/fortune/edges/vertex/2/edges){#fig-vertex-event-edges2}

![](figures/fortune/edges/vertex/3/edges){#fig-vertex-event-edges3}

Vertex events require substantial updates to the half-edges of
the Voronoi graph.  (a) The breakpoints that merge in a vertex
event contain references to two half-edges, here `he1` and `he3`.
(b) When processing a vertex event we delete these references
(c) Then we create two new half-edges, one of which is referenced
by the new vertex, here `he5`, and one of which is referenced by
the new breakpoint, here `he6`.
:::

![The trickiest part of processing a vertex event is making sure
that the four existing half-edges and the two new half-edges all
properly reference each other.  These references ultimately
define the structure of the Voronoi graph.](
figures/fortune/edges/connections/edges){
width=50% #fig-vertex-event-edge-connections}

At this point we're ready to remove the collapsing arc and
replace its two neighboring breakpoints with a new merged
breakpoint (@fig-bst-vertex-event1, @fig-bst-vertex-event2).
This can be done by updating the half-edge reference of the
successor node and then applying a balanced deletion operation
(@fig-bst-vertex-event3).

::: {#fig-bst-vertex-event layout="[ [100], [100], [100] ]"}
![](figures/bst/vertex_event/beachlines/vertex_event){#fig-bst-vertex-event1}

![](figures/bst/vertex_event/arcs/vertex_event){#fig-bst-vertex-event2}

![](figures/bst/vertex_event/bsts/vertex_event){#fig-bst-vertex-event3}

\(a) A vertex event removes one of the arcs in the initial beach
line, replacing the two neighboring breakpoints with a new merged
breakpoint.  (b)  This corresponds to removing one of the initial
arc-breakpoint pairs and then updating the breakpoint information
in another arc-breakpoint pair.  (c) The removal of the
deprecated arc-breakpoint pair is implemented by applying the
deletion operation to the initial beach line tree.
:::

Lastly there are two new triplets of neighboring arcs that we
have to check for possible new vertex events.  Once again we can
efficiently access these arcs by repeatedly applying the
predecessor and successor beach line tree operations.

## Clean Up

Once the event queue has been exhausted we are left with a
doubly-connected edge list that contains all of the information
about the Voronoi graph, and hence the Voronoi diagram.

At this point the beach line tree will _not_ be empty.  In fact
the remaining breakpoint nodes reference all of the half-edges
with unbounded origins.  If we want to visualize the Voronoi
graph neatly then we want to anchor these dangling edges by
creating a bounding box and then adding a pseudo-vector to the
doubly-connected edge list for the intersection of each dangling
half-edge with that boundary.

# Demonstration

To put all of this discussion into practice let's work through a
full implementation of Fortune's algorithm in `R`.  Although
doubly-connected edge lists, binary search trees, and event
queues can all be implemented with functional programming I
strongly prefer a more object-oriented approach.  To achieve
object-oriented programming in `R` I will use `R6` classes
[@Chang:2025], at the cost of some clumsy interactions with base
`R` graphics.  At least using `R6` classes gets us
pass-by-reference semantics for class instances instead of `R`'s
usual call-by-need/lazy evaluation semantics!

```{r}
library(R6)
```

Speaking of graphics, let's tweak the local graphics
configuration.

```{r}
par(family="serif", las=1, bty="l",
    cex.axis=1, cex.lab=1, cex.main=1,
    xaxs="i", yaxs="i", mar = c(5, 5, 3, 1))

c_light <- c("#DCBCBC")
c_light_highlight <- c("#C79999")
c_mid <- c("#B97C7C")
c_mid_highlight <- c("#A25050")
c_dark <- c("#8F2727")
c_dark_highlight <- c("#7C0000")

c_light_teal <- c("#6B8E8E")
c_mid_teal <- c("#487575")
c_dark_teal <- c("#1D4F4F")
```

## Beach Line

To define a beach line we'll first need nodes.

This class implements arc-breakpoint nodes with satellite data
for the associated site, half-edge, and vertex event.  Note that
the site variables are integer indices which will be used to
access site information from global arrays.

The variable `red` denotes the coloring of the node which is used
for balancing binary trees.

```{r}
node <- R6Class("node",
  public = list(
    site = NULL,
    parent = NULL,
    left = NULL,
    right = NULL,
    red = FALSE,
    id = NULL,
    he = NULL,  # Half-edge originating at (predecessor$site, site)
    scheduled_vertex = NULL,
    initialize = function(beachline) {
      if (is.integer(beachline)) {
        self$id <- beachline
      } else {
        self$parent <- beachline$nil
        self$left <- beachline$nil
        self$right <- beachline$nil

        beachline$last_node_id <- beachline$last_node_id + 1
        self$id <- beachline$last_node_id
      }
    }
  )
)
```

The `id` variable allows us to determine if any two `node`
instances refer to the same object.  In many programming
languages this is done by directly comparing memory addresses,
but because `R6` classes are built up from `R` environments this
is not possible here.

```{r}
`==.node` = function(x, y) { x$id == y$id }
`!=.node` = function(x, y) { x$id != y$id }
```

A beach line tree itself is defined by its root node.  Here we'll
be using a self-balancing red-black binary tree which uses the
coloring of the nodes to limit the height of any given tree.
Red-black binary trees requires a dedicated null, or nil, node.

```{r}
beachline <- R6Class("beachline",
  public = list(
   nil = NULL,
   root = NULL,
   last_node_id = NULL,
   initialize = function() {
     self$nil <- node$new(as.integer(0))
     self$root <- self$nil
     self$last_node_id <- as.integer(1)
   }
  ))
```

The `beachline` class also gives us a place to keep track of the
node indices.

### Black-Red Binary Tree Operations

The `insert_fixup` function balances a binary tree after we
add the node `z`.

```{r}
beachline$set("private", "left_rotate",
function(x) {
  if (x$right == self$nil)
    return()

  y <- x$right
  x$right <- y$left

  if (y$left != self$nil)
    y$left$parent <- x

  y$parent <- x$parent

  if (x$parent == self$nil) {
    self$root <- y
  } else {
    if (x$parent$left == x)
      x$parent$left <- y
    else
      x$parent$right <- y
  }

  y$left <- x
  x$parent <- y
})
```

```{r}
beachline$set("private", "right_rotate",
function(x) {
  if (x$left == self$nil)
    return()

  y <- x$left
  x$left <- y$right

  if (y$right != self$nil)
    y$right$parent <- x

  y$parent <- x$parent

  if (x$parent == self$nil) {
    self$root <- y
  } else {
    if (x$parent$right == x)
      x$parent$right <- y
    else
      x$parent$left <- y
  }

  y$right <- x
  x$parent <- y
})
```

```{r}
beachline$set("private", "insert_fixup",
function(z) {
  while (z$parent$red) {
    if (z$parent == z$parent$parent$left) {
      y <- z$parent$parent$right
      if (y$red) {
        z$parent$red <- FALSE
        y$red <- FALSE
        z$parent$parent$red <- TRUE
        z <- z$parent$parent
      } else {
        if (z == z$parent$right) {
          z <- z$parent
          private$left_rotate(z)
        }
        z$parent$red <- FALSE
        z$parent$parent$red <- TRUE
        private$right_rotate(z$parent$parent)
      }
    } else {
      y <- z$parent$parent$left

      if (y$red) {
        z$parent$red <- FALSE
        y$red <- FALSE
        z$parent$parent$red <- TRUE
        z <- z$parent$parent
      } else {
        if (z == z$parent$left) {
          z <- z$parent
          private$right_rotate(z)
        }
        z$parent$red <- FALSE
        z$parent$parent$red <- TRUE
        private$left_rotate(z$parent$parent)
      }
    }
  }
  self$root$red <- FALSE
})
```

The `insert_fixup` function then allows us to implement
balanced `insert-before` and `insert-after` operations.

```{r}
beachline$set("public", "insert_before",
function(x, y) {
  if (x$left == self$nil) {
    x$left <- y
    y$parent <- x
  } else {
    z <- self$predecessor(x)
    z$right <- y
    y$parent <- z
  }

  y$left <- self$nil
  y$right <- self$nil
  y$red <- TRUE

  private$insert_fixup(y)
})
```

```{r}
beachline$set("public", "insert_after",
function(x, y) {
  if (x$right == self$nil) {
    x$right <- y
    y$parent <- x
  } else {
    z <- self$successor(x)
    z$left <- y
    y$parent <- z
  }

  y$left <- self$nil
  y$right <- self$nil
  y$red <- TRUE

  private$insert_fixup(y)
})
```

Deleting a node from a red-black tree takes some care.  To
properly remove a node we'll need to be able to calculate the
boundaries of sub-trees.

```{r}
beachline$set("public", "min",
function(x) {
  if (x == self$nil)
    return(self$nil)
  while (x$left != self$nil)
    x <- x$left
  x
})
```

```{r}
beachline$set("public", "max",
function(x) {
  if (x == self$nil)
    return(self$nil)
  while (x$right != self$nil)
    x <- x$right
  x
})
```

Then we'll need to carefully balance the tree.

```{r}
beachline$set("private", "transplant",
function(u, v) {
  if (u$parent == self$nil) {
    self$root <- v
  } else {
    if (u == u$parent$left) {
      u$parent$left <- v
    } else {
      u$parent$right <- v
    }
  }
  v$parent <- u$parent
})
```

```{r}
beachline$set("private", "delete_fixup",
function(x) {
  while (x != self$root && !x$red) {
    if (x == x$parent$left) {
      w <- x$parent$right

      if (w$red) {
        w$red <- FALSE
        x$parent$red <- TRUE
        private$left_rotate(x$parent)
        w <- x$parent$right
      }

      if (!w$left$red && !w$right$red) {
        w$red <- TRUE
        x <- x$parent
      } else {
        if (!w$right$red) {
          w$left$red <- FALSE
          w$red <- TRUE
          private$right_rotate(w)
          w <- x$parent$right
        }

        w$red <- x$parent$red
        x$parent$red <- FALSE
        w$right$red <- FALSE
        private$left_rotate(x$parent)
        x <- self$root
      }
    } else {
      w <- x$parent$left

      if (w$red) {
        w$red <- FALSE
        x$parent$red <- TRUE
        private$right_rotate(x$parent)
        w <- x$parent$left
      }

      if (!w$right$red && !w$left$red) {
        w$red <- TRUE
        x <- x$parent
      } else {
        if (!w$left$red) {
          w$right$red <- FALSE
          w$red <- TRUE
          private$left_rotate(w)
          w <- x$parent$left
        }

        w$red <- x$parent$red
        x$parent$red <- FALSE
        w$left$red <- FALSE
        private$right_rotate(x$parent)
        x <- self$root
      }
    }
  }
  x$red <- FALSE
})
```

```{r}
beachline$set("public", "delete",
function(z) {
  y <- z
  y_old_red <- y$red

  if (z$left == self$nil) {
    x <- z$right
    private$transplant(z, z$right)
  } else if (z$right == self$nil) {
    x <- z$left
    private$transplant(z, z$left)
  } else {
    y <- self$min(z$right)
    y_old_red <- y$red
    x <- y$right

    if (y != z$right) {
      private$transplant(y, y$right)
      y$right <- z$right
      y$right$parent <- y
    } else {
      x$parent <- y
    }

    private$transplant(z, y)
    y$left <- z$left
    y$left$parent <- y
    y$red <- z$red
  }

  if (!y_old_red)
    private$delete_fixup(x)
})
```

Neighboring nodes can be accessed with `precursor` and
`successor` operations.

```{r}
beachline$set("public", "predecessor",
function(x) {
  if (x$left != self$nil) {
    return(self$max(x$left))
  } else {
    y <- x$parent
    while (   y != self$nil
              && x == y$left) {
      x <- y
      y <- y$parent
    }
    return(y)
  }
})
```

```{r}
beachline$set("public", "successor",
function(x) {
  if (x$right != self$nil) {
    return(self$min(x$right))
  } else {
    y <- x$parent
    while (   y != self$nil
           && x == y$right) {
      x <- y
      y <- y $parent
    }
    return(y)
  }
})
```

### Visualization Functions

Visualizing the nodes in a binary search tree is particularly
elegant with recursive functions.  That said any function that
interacts with base `R` graphics objects has to be external to
the `R6` class.

```{r}
beachline$set("public", "height",
function(node=NULL) {
  if (is.null(node))
    node <- self$root

  if (node == self$nil)
    return(0)

  left_height <- self$height(node$left)
  right_height <- self$height(node$right)

  1 + max(left_height, right_height)
})
```

Just to minimize the text, we'll display only the site that
generates the arc associated with each node and not any
information about the lower breakpoint.

```{r}
plot_node <- function(nil, node,
                      x, y,
                      delta_x, delta_y,
                      text_cex) {

  if (node$left != nil) {
    lines(c(x, x - delta_x), c(y, y - delta_y),
          col=c_mid_teal, lwd=2)
    plot_node(nil, node$left,
              x - delta_x, y - delta_y,
              delta_x / 2, delta_y, text_cex)
  }

  if (node$right != nil) {
    lines(c(x, x + delta_x), c(y, y - delta_y),
          col=c_mid_teal, lwd=2)
    plot_node(nil, node$right,
              x + delta_x, y - delta_y,
              delta_x / 2, delta_y, text_cex)
  }

  if (node$red)
    points(x, y, pch=16, col=c_dark, cex=3)
  else
    points(x, y, pch=16, col="black", cex=3)

  text(x, y, col="white", cex=text_cex,
       labels=node$site)
}
```

```{r}
plot_tree <- function(tree, delta=0.5, text_cex=0.75) {
  H <- tree$height()

  par(mar=c(0, 0, 0, 0))

  plot(NULL,
       xlab='', ylab='', xaxt="n", yaxt="n", frame.plot=F,
       xlim=c(-delta * (2**(H - 1) + 1), delta * (2**(H - 1) + 1)),
       ylim=c(-(H - 0.5) * delta, 0.5 * delta))

  plot_node(tree$nil, tree$root, 0, 0,
            delta * 2**(H - 2), delta, text_cex)
}
```

### Fortune's Algorithm Geometry Calculations

As we've seen, implementing Fortune's algorithm requires a lot of
geometric calculations.

For example we'll need to be able to compute the vertical
position of the breakpoint separating two neighboring arcs on a
beach line.  Site positions are accessed by indexing the global
arrays `xs` and `ys`.

```{r}
beachline$set("public", "pair_intersection",
function(n1, n2, S) {
  if (xs[n1] == S)
    return(ys[n1])
  if (xs[n2] == S)
    return(ys[n2])

  a <- xs[n2] - xs[n1]
  b <- ys[n2] * (S - xs[n1]) - ys[n1] * (S - xs[n2])
  c <- (   ys[n2]**2 * (S - xs[n1])
         - ys[n1]**2 * (S - xs[n2])
         - (S - xs[n1]) * (S - xs[n2]) * (xs[n2] - xs[n1]) )

  e <- b / a
  d <- sqrt( e**2 - c / a )
  e - sign(a) * d
})
```

The horizontal position of a breakpoint can be recovered from the
parabola defining each arc.  Here `(fx, fy)` defines the focal
point of the parabola and `S` the vertical directrix.

```{r}
beachline$set("public", "parabola_xs",
function(parabola_ys, fx, fy, S) {
  0.5 * ( (fx + S) - (parabola_ys - fy)**2 / (S - fx) )
})
```

We can't forget the circumcircles.

```{r}
beachline$set("public", "circumcircle",
function(n1, n2, n3) {
  D <- 2 * (     xs[n1] * (ys[n2] - ys[n3])
               + xs[n2] * (ys[n3] - ys[n1])
               + xs[n3] * (ys[n1] - ys[n2]) )
  xc <- (    (xs[n1]**2 + ys[n1]**2) * (ys[n2] - ys[n3])
           + (xs[n2]**2 + ys[n2]**2) * (ys[n3] - ys[n1])
           + (xs[n3]**2 + ys[n3]**2) * (ys[n1] - ys[n2]) ) / D

  yc <- (    (xs[n1]**2 + ys[n1]**2) * (xs[n3] - xs[n2])
           + (xs[n2]**2 + ys[n2]**2) * (xs[n1] - xs[n3])
           + (xs[n3]**2 + ys[n3]**2) * (xs[n2] - xs[n1]) ) / D

  dx <- xc - xs[n1]
  dy <- yc - ys[n1]

  if (dx > dy) {
    r <- abs(dx) * sqrt(1 + (dy / dx)**2)
  } else {
    r <- abs(dy) * sqrt(1 + (dx / dy)**2)
  }

  c(xc, yc, r)
})
```

We can use these geometric functions to, for example, find the
node representing the arc that intersects with a given vertical
position.

```{r}
beachline$set("private", "arc_search",
function(z, y, S) {
  while (1) {
    z_left <- self$predecessor(z)
    if (z_left != self$nil) {
      by <- self$pair_intersection(z_left$site, z$site, S)
      if (y <= by) {
        z <- z$left
        next
      }
    }

    z_right <- self$successor(z)
    if (z_right != self$nil) {
      by <- self$pair_intersection(z$site, z_right$site, S)
      if (y > by) {
        z <- z$right
        next
      }
    }

    break
  }

  return(z)
})
```

### Fortune's Algorithm Event Functions

With all of these auxiliary functions we are finally ready to
implement event processing functions.

One operating that appears over and over again when processing
events is checking triplets of arcs to see if they define a valid
vertex event and, if so, adding that event to the event queue
`queue`.

As we discussed in [Section 3.4.2](@sec:contract) there is
undoubtably a more elegant way to verify that `alpha` will
collapse as the sweep line progresses than comparing `upper_int`
to `lower_int`, but it will do for now.

```{r}
beachline$set("private", "schedule_new_vertex_event",
function(queue, alpha_l, alpha, alpha_r, S) {
  # Skip vertex if any of the neighboring arcs are nil
  if (alpha_l == self$nil || alpha_r == self$nil) {
    return()
  }

  # Skip vertex if any of the sites are the same
  if (alpha_l$site == alpha_r$site) {
    return()
  }

  # Construct circumcircle
  int <- self$circumcircle(alpha_l$site,
                           alpha$site,
                           alpha_r$site)
  S_new <- int[1] + int[3]


  lower_int <- self$pair_intersection(alpha_l$site,
                                      alpha$site, S_new)
  upper_int <- self$pair_intersection(alpha$site,
                                      alpha_r$site, S_new)

  if (   S_new > S
      && (abs(upper_int - lower_int) < (1e6 * .Machine$double.eps)) ) {
    vertex_event <- event$new(NULL, S_new, NULL, alpha)
    alpha$scheduled_vertex <- vertex_event
    queue$insert_event(vertex_event)
  }
})
```

Now we can process site events.

```{r}
beachline$set("private", "process_site_event",
function(queue, graph, new_event) {
  new_site <- new_event$site

  # Initialize beachline if empty
  if (self$root == self$nil) {
    self$root <- node$new(self)
    self$root$site <- new_site
    return()
  }

  # Find beachline arc that intersects
  # vertical position of new site
  alpha <- private$arc_search(self$root,
                              ys[new_site],
                              new_event$x)

  # Delete any scheduled vertex
  # events at intersecting arc
  if (!is.null(alpha$scheduled_vertex)) {
    queue$delete_event(alpha$scheduled_vertex)
    alpha$scheduled_vertex <- NULL
  }

  # Split intersecting arc
  alpha_r <- node$new(self)
  alpha_r$site <- new_site

  self$insert_after(alpha, alpha_r)

  alpha_rr <- node$new(self)
  alpha_rr$site <- alpha$site

  self$insert_after(alpha_r, alpha_rr)

  # Instantiate and connect new half-edges
  new_he <- graph$new_half_edge(left_face=alpha$site)     # (alpha, alpha_r)
  new_he$twin <- graph$new_half_edge(left_face=new_site,  # (alpha_r, alpha_rr)
                                     twin=new_he)

  alpha_r$he <- new_he
  alpha_rr$he <- new_he$twin

  # Attempt to schedule vertex events
  # for the two new arc triplets
  alpha_l <- self$predecessor(alpha)
  private$schedule_new_vertex_event(queue,
                                    alpha_l, alpha, alpha_r,
                                    new_event$x)

  alpha_rrr <- self$successor(alpha_rr)
  private$schedule_new_vertex_event(queue,
                                    alpha_r, alpha_rr, alpha_rrr,
                                    new_event$x)
})
```

Vertex event processing follows.

```{r}
beachline$set("private", "process_vertex_event",
function(queue, graph, new_event) {
  alpha <- new_event$arc

  alpha_r <- self$successor(alpha)
  alpha_rr <- self$successor(alpha_r)

  alpha_l <- self$predecessor(alpha)
  alpha_ll <- self$predecessor(alpha_l)

  # Clean up deprecated vertex events in the queue
  if (!is.null(alpha_r$scheduled_vertex)) {
    queue$delete_event(alpha_r$scheduled_vertex)
    alpha_r$scheduled_vertex <- NULL
  }

  if (!is.null(alpha_l$scheduled_vertex)) {
    queue$delete_event(alpha_l$scheduled_vertex)
    alpha_l$scheduled_vertex <- NULL
  }

  # Create vertex
  int <- self$circumcircle(alpha_l$site, alpha$site, alpha_r$site)
  v <- vertex$new(int[1:2], alpha_l$site, alpha$site, alpha_r$site)
  graph$add_vertex(v)

  # Connect vertex
  alpha_r$he$origin <- v
  alpha$he$origin <- v

  new_he <- graph$new_half_edge(origin=v,
                                left_face=alpha_r$site)
  new_he$twin <- graph$new_half_edge(left_face=alpha_l$site,
                                     twin=new_he)

  v$he <- new_he

  # Connect half-edges
  alpha$he$twin$nxt <- alpha_r$he
  alpha_r$he$prv <- alpha$he$twin

  alpha_r$he$twin$nxt <- new_he
  new_he$prv <- alpha_r$he$twin

  new_he$twin$nxt <- alpha$he
  alpha$he$prv <- new_he$twin

  # Update active half-edges
  alpha_r$he <- new_he$twin

  # Delete deprecated arcs
  self$delete(alpha)

  # Check for new vertex events
  private$schedule_new_vertex_event(queue,
                                    alpha_ll, alpha_l, alpha_r,
                                    new_event$x)

  private$schedule_new_vertex_event(queue,
                                    alpha_l, alpha_r, alpha_rr,
                                    new_event$x)
})
```

Lastly we'll wrap both of these event processing functions
together.

```{r}
beachline$set("public", "process_next_event",
function(queue, graph) {
  next_event <- queue$pop_next_element()

  if (!is.null(next_event$site)) {
    private$process_site_event(queue,
                               graph,
                               next_event)
  } else {
    private$process_vertex_event(queue,
                                 graph,
                                 next_event)
  }
})
```

### Fortune's Algorithm Visualization Functions

To monitor the progress of Fortune's algorithm we'll define some
functions that allow us to plot the shape of the beach line for
any sweep line position, as well as the behavior of any
preliminary Voronoi edges.

```{r}
plot_arcs <- function(bl, S, x=NULL, y_min, y_max) {
  if (is.null(x))
    x <- bl$root

  if (x == bl$nil)
    return(y_max)

  # Recurse right
  y_max <- plot_arcs(bl, S, x$right, y_min, y_max)

  # Plot arc
  if (is.null(x$he)) {
    # Plot parabola
    if (y_min + 1e-3 < y_max) {
      parabola_ys <- seq(y_min, y_max, 1e-3)
      lines(bl$parabola_xs(parabola_ys, xs[x$site], ys[x$site], S),
            parabola_ys, lty=2, lwd=2, col=c_mid)
    }
  } else {
    # Compute left breakpoint
    by <- bl$pair_intersection(x$he$left_face, x$site, S)
    bx <- bl$parabola_xs(by, xs[x$site], ys[x$site], S)

    # Plot parabola
    if (by + 1e-3 < y_max) {
      parabola_ys <- seq(by, y_max, 1e-3)
      lines(bl$parabola_xs(parabola_ys, xs[x$site], ys[x$site], S),
            parabola_ys, lty=2, lwd=2, col=c_mid)
    }

    # Plot left breakpoint
    points(bx, by, col=c_light, pch=16)
    text(bx + 0.1, by,
        labels=paste0(x$he$left_face, ",", x$site))

    # Save breakpoint position for plotting half-edges
    x$he$pseudo_origin <- c(bx, by)

    # Update upper bound
    y_max <- by
  }

  # Recurse left
  y_max <- plot_arcs(bl, S, x$left, y_min, y_max)

  return(y_max)
}
```

```{r}
plot_beachline <- function(bl, S,
                           xlim=c(-4, 4),
                           ylim=c(-4, 4)) {

  plot(xs, ys, col=c_dark, pch=16,
       xlim=xlim, ylim=ylim,
       main="")

  # Plot sweep line
  abline(v=S, col=c_mid_teal, lty=2)

  # Plot site parabolas given the sweep line
  for (i in which(xs < S)) {
    text(xs[i] + 0.1, ys[i], i)
    parabola_ys <- seq(ylim[1], ylim[2], 0.01)
    lines(bl$parabola_xs(parabola_ys, xs[i], ys[i], S),
          parabola_ys, lwd=2, col="#DDDDDD")
  }

  # Plot beach line
  if (bl$root != bl$nil)
    r <- plot_arcs(bl, S, y_min=ylim[1], y_max=ylim[2])
}
```

```{r}
plot_beachline_half_edges <- function(bl, S, x=NULL,
                                      lwd=2, col=c_mid_teal,
                                      delta=0.08, eps=0.04,
                                      theta=60, gamma=0.0075) {
  if (is.null(x))
    x <- bl$root

  if (x != bl$nil) {
    # Recurse right
    plot_beachline_half_edges(bl, S, x$right,
                              lwd, col, delta,
                              eps, theta, gamma)

    # Plot arc
    if (!is.null(x$he)) {
      if (is.null(x$he$origin)) p1 <- x$he$pseudo_origin
      else                      p1 <- x$he$origin$pos

      if (is.null(x$he$twin$origin)) p2 <- x$he$twin$pseudo_origin
      else                           p2 <- x$he$twin$origin$pos

      plot_half_edge(p1, p2, lwd, col, delta, eps, theta, gamma)
    }

    # Recurse left
    plot_beachline_half_edges(bl, S, x$left,
                              lwd, col, delta,
                              eps, theta, gamma)
  }
}
```

## Event Queue

We'll overload the `event` class to handle both site and vertex
events at the same time.  The variable `site` is used for only
site events while the variable `arc` is used for only vertex
events.

```{r}
event <- R6Class("event",
  public=list(
  x=NULL,
  site=NULL,
  arc=NULL,
  queue_idx=NULL,
  initialize = function(i, x, site, arc) {
    self$queue_idx <- i
    self$x <- x
    self$site <- site
    self$arc <- arc
  }))
```

The event queue itself is a pretty straightforward implementation
of a priority queue data structure.  When creating a new event
queue instance we pass in a vector of indices for the available
sites which are then used to create the site events that
initially populate the queue.

```{r}
event_queue <- R6Class("event_queue",
  public=list(
   idxs=NULL,
   events=NULL,
   length=0,
   initialize = function(sites, xs) {
     self$length <- length(sites)
     self$idxs <- 1:self$length
     for (i in seq_along(sites)) {
       new_event <- event$new(i, xs[i], i, NULL)
       self$events <- append(self$events,
                             new_event)
     }
     for (i in (self$length %/% 2):1) {
       private$min_heapify(i)
     }
   },
   is_empty = function() {
     self$length == 0
   },
   is_not_empty = function() {
     self$length > 0
   }))
```

```{r}
event_queue$set("private", "min_heapify",
function(i) {
  l <- 2 * i
  r <- l + 1

  if (   l <= self$length
      &&   self$events[[self$idxs[l]]]$x
         < self$events[[self$idxs[i]]]$x) {
    smallest <- l
  } else {
    smallest <- i
  }

  if (   r <= self$length
         &&   self$events[[self$idxs[r]]]$x
         < self$events[[self$idxs[smallest]]]$x) {
    smallest <- r
  }

  if (smallest != i) {
    n <- self$events[[self$idxs[i]]]$queue_idx
    self$events[[self$idxs[i]]]$queue_idx <-
      self$events[[self$idxs[smallest]]]$queue_idx
    self$events[[self$idxs[smallest]]]$queue_idx <- n

    n <- self$idxs[i]
    self$idxs[i] <- self$idxs[smallest]
    self$idxs[smallest] <- n

    private$min_heapify(smallest)
  }
})
```

```{r}
event_queue$set("public", "insert_event",
function(new_event) {
  self$length <- self$length + 1
  self$idxs <- c(self$idxs, length(self$events) + 1)

  new_event$queue_idx <- self$length
  self$events <- append(self$events, new_event)

  i <- self$length
  while (   i > 1
            &&   self$events[[self$idxs[i %/% 2]]]$x
            > self$events[[self$idxs[i]]]$x) {
    n <- self$events[[self$idxs[i]]]$queue_idx
    self$events[[self$idxs[i]]]$queue_idx <-
      self$events[[self$idxs[i %/% 2]]]$queue_idx
    self$events[[self$idxs[i %/% 2]]]$queue_idx <- n

    n <- self$idxs[i]
    self$idxs[i] <- self$idxs[i %/% 2]
    self$idxs[i %/% 2] <- n

    i <- i %/% 2
  }
})
```

```{r}
event_queue$set("public", "insert_site_event",
function(site, xs) {
  self$insert_event(event$new(NULL, xs[site], site, NULL))
})
```

```{r}
event_queue$set("public", "insert_vertex_event",
function(S, arc) {
  self$insert_event(event$new(NULL, S, NULL, arc))
})
```

```{r}
event_queue$set("public", "pop_next_element",
function() {
  if (self$length == 0) {
    print("The queue is empty!")
    return(NULL)
  }

  next_element <- self$events[[self$idxs[1]]]

  self$idxs[1] <- self$idxs[self$length]
  self$events[[self$idxs[1]]]$queue_idx <- 1

  self$idxs <- self$idxs[-self$length]
  self$length <- self$length - 1

  private$min_heapify(1)

  next_element
})
```

```{r}
event_queue$set("public", "delete_event",
function(event) {
  event$x <- self$events[[self$idxs[1]]]$x - 1

  i <- event$queue_idx
  while (   i > 1
         &&   self$events[[self$idxs[i %/% 2]]]$x
            > self$events[[self$idxs[i]]]$x) {
    n <- self$events[[self$idxs[i]]]$queue_idx
    self$events[[self$idxs[i]]]$queue_idx <-
      self$events[[self$idxs[i %/% 2]]]$queue_idx
    self$events[[self$idxs[i %/% 2]]]$queue_idx <- n

    n <- self$idxs[i]
    self$idxs[i] <- self$idxs[i %/% 2]
    self$idxs[i %/% 2] <- n

    i <- i %/% 2
  }
  e <- self$pop_next_element()
})
```

```{r}
event_queue$set("public", "check_min_heapify",
function() {
  if (self$length > 0) {
    for (i in 1:self$length)  {
      if (i %/% 2 == 0) next
      print(   self$events[[self$idxs[i %/% 2]]]$x
            <= self$events[[self$idxs[i]]]$x)
    }
  }
})
```

```{r}
event_queue$set("public", "next_event",
function() {
  if (self$length > 0)
    self$events[[self$idxs[1]]]
})
```

## Doubly-Connected Edge List

In order to represent a Voronoi graph we'll need classes for
half_edges, vertices, and faces and then the doubly-connected
edge list that encapsulates them.

Like the beach line nodes the half-edges uses a global indexing
to allow for equality comparisons.  The `pseudo_origin` variable
is used for plotting dangling half-edges that have not yet been
anchored to vertices.

```{r}
half_edge <- R6Class("half_edge",
  public=list(
   pseudo_origin=NULL,
   origin=NULL,
   left_face=NULL,
   twin=NULL,
   prv=NULL,
   nxt=NULL,
   visited=FALSE,
   id=NULL,
   initialize = function(graph, o, lf, t, p, n) {
     self$origin <- o
     self$left_face <- lf
     self$twin   <- t
     self$prv    <- p
     self$nxt    <- n

     graph$last_half_edge_id <- graph$last_half_edge_id + 1
     self$id <- graph$last_half_edge_id
   }))
```

```{r}
`==.half_edge` = function(x, y) { x$id == y$id }
`!=.half_edge` = function(x, y) { x$id != y$id }
```

```{r}
vertex <- R6Class("vertex",
  public=list(
    pos=NULL,
    n1=NULL,
    n2=NULL,
    n3=NULL,
    he=NULL,
    initialize = function(p, n1, n2, n3) {
      self$pos <- p
      self$n1 <- n1
      self$n2 <- n2
      self$n3 <- n3
    }))
```

```{r}
face <- R6Class("face",
  public=list(
    site=NULL,
    he=NULL,
    initialize = function(site, he) {
      self$site <- site
      self$he <- he
    }))
```

```{r}
dcel <- R6Class("dcel",
  public=list(
    vertices=NULL,
    pseudo_vertices=NULL,
    faces=NULL,
    xlim=NULL,
    ylim=NULL,
    last_half_edge_id = NULL,
    initialize = function() {
      self$vertices <- list()
      self$pseudo_vertices <- list()
      self$faces <- list()
      self$last_half_edge_id <- as.integer(1)
    },
    add_vertex = function(v) {
      self$vertices <-
        append(self$vertices, v)
    },
    N_vertices = function() {
      length(self$vertices)
    },
    add_pseudo_vertex = function(pv) {
      self$pseudo_vertices <-
        append(self$pseudo_vertices, pv)
    },
    N_pseudo_vertices = function() {
      length(self$pseudo_vertices)
    },
    add_face = function(f) {
      self$faces <-
        append(self$faces, f)
    },
    N_faces = function() {
      length(self$faces)
    }))
```

```{r}
dcel$set("public", "new_half_edge",
function(origin=NULL, left_face=NULL,
         twin=NULL, prv=NULL, nxt=NULL) {
  half_edge$new(self, origin, left_face, twin, prv, nxt)
})
```

There are all kinds of operations that we can implement from the
structure encoded in a doubly-connected edge list.  For example
we can automatically compute a bounding box that contains all of
the vertices.  The parameter `q` specifies how much
multiplicative margin we add around the smallest bounding box.

```{r}
dcel$set("public", "compute_bounding_box",
function(q=0.05) {
  # Set bounding box edges to
  # extreme vertex positions
  self$xlim <- c(self$vertices[[1]]$pos[1],
                 self$vertices[[1]]$pos[1])
  self$ylim <- c(self$vertices[[1]]$pos[2],
                 self$vertices[[1]]$pos[2])

  for (v in self$vertices[-1]) {
    if (v$pos[1] > self$xlim[2]) {
      self$xlim[2] <- v$pos[1]
    } else if (v$pos[1] < self$xlim[1]) {
      self$xlim[1] <- v$pos[1]
    }

    if (v$pos[2] > self$ylim[2]) {
      self$ylim[2] <- v$pos[2]
    } else if (v$pos[2] < self$ylim[1]) {
      self$ylim[1] <- v$pos[2]
    }
  }

  # Add proportional margins to bounding box
  dx <- self$xlim[2] - self$xlim[1]
  self$xlim[1] <- self$xlim[1] - q * dx
  self$xlim[2] <- self$xlim[2] + q * dx

  dy <- self$ylim[2] - self$ylim[1]
  self$ylim[1] <- self$ylim[1] - q * dy
  self$ylim[2] <- self$ylim[2] + q * dy
})
```

Given a bounding box we can then introduce pseudo-vertices to
anchor any dangling half-edges.

```{r}
dcel$set("public", "anchor_dangling_edges",
function(z, xs, ys) {
  # Dangling edge
  he <- z$he

  if (!is.null(he)) {
    nu <- he$twin$origin$pos
    zeta <- -nu

    i_left <- he$left_face
    i_right <- he$twin$left_face
    dangling_he <- he

    # Anchored edges
    he <- he$nxt
    zeta <- zeta + he$twin$origin$pos

    he <- he$twin$nxt
    zeta <- zeta + he$twin$origin$pos

    # Compute bisecting vector in the opposite
    # direction of the anchored edges
    delta <- c(ys[i_right] - ys[i_left],
               xs[i_left]  - xs[i_right])

    # Compute closest bounding box intersection
    bh <- ifelse(delta[1] > 0, self$xlim[2], self$xlim[1])
    bv <- ifelse(delta[2] > 0, self$ylim[2], self$ylim[1])

    pv_x <- bh
    pv_y <- nu[2] + delta[2] * (pv_x - nu[1]) / delta[1]
    if (pv_y < self$ylim[1] || pv_y > self$ylim[2]) {
      pv_y <- bv
      pv_x <- nu[1] + delta[1] * (pv_y - nu[2]) / delta[2]
    }

    # Anchor dangling edge to a new pseudo-vertex
    # positioned at the bounding box intersection
    pv <- vertex$new(c(pv_x, pv_y), NULL, NULL, NULL)
    dangling_he$origin <- pv
    pv$he <- dangling_he
    self$add_pseudo_vertex(pv)

    # Recurse
    self$anchor_dangling_edges(z$left, xs, ys)
    self$anchor_dangling_edges(z$right, xs, ys)
  }
})
```

More relevant to the construction of a Voronoi diagrams, we also
can follow the half-edges to automatically identify all of the
faces in the graph, and hence cells in the diagram.

```{r}
dcel$set("private", "reset_visitation",
function() {
  for (v in self$vertices) {
    he <- v$he
    he$visited <- FALSE
    he$twin$visited <- FALSE
    he <- he$prv
    he$visited <- FALSE
    he$twin$visited <- FALSE
    he <- he$twin$prv
    he$visited <- FALSE
    he$twin$visited <- FALSE
  }
})
```

```{r}
dcel$set("public", "find_faces",
function(he=NULL) {

  if (is.null(he)) {
    private$reset_visitation()
    he <- self$vertices[[1]]$he
  }

  init_he <- he
  open_face <- FALSE
  closed_face <- FALSE

  while (!he$visited) {
    he$visited <- TRUE
    he <- he$nxt
    if (is.null(he)) {
      open_face <- TRUE
      break
    }
    if (he == init_he) {
      closed_face <- TRUE
      break
    }
  }

  he <- init_he
  if (open_face) {
    # Reverse to beginning of open boundary
    while(!is.null(he$prv)) {
      he <- he$prv
    }

    self$add_face(face$new(he$left_face, he))

    # Shoot along open boundary
    while(!is.null(he$nxt)) {
      if(!he$twin$visited) self$find_faces(he$twin)
      he <- he$nxt
    }
  }
  if (closed_face) {
    self$add_face(face$new(he$left_face, he))

    # Shoot along closed boundary
    while(1) {
      if(!he$twin$visited) self$find_faces(he$twin)
      he <- he$nxt
      if (he == init_he) break
    }
  }
})
```

Why spend all of this time creating graphs if we can't visualize
them with aesthetic flair?

```{r}
plot_half_edge <- function(pi, pf, lwd, col=c_mid_teal,
                           delta, eps, theta, gamma) {
  r <- pf - pi
  v <- r / sqrt(sum(r**2))

  r_perp <- c(r[2], -r[1])
  w <- r_perp / sqrt(sum(r_perp**2))

  # Perpendicular offset
  gammax <- -gamma * w[1]
  gammay <- -gamma * w[2]

  # Arrow body
  lines(c(pi[1], pf[1]) + gammax,
        c(pi[2], pf[2]) + gammay,
        lwd=lwd, col=col)

  # Arrow head
  phi_rad <- (90 - theta) * (3.14159265 / 180)
  p3 <- pf - (delta + eps * tan(phi_rad)) * v - eps * w
  p4 <- pf - delta * v

  polygon(c(pf[1], p3[1], p4[1], pf[1]) + gammax,
          c(pf[2], p3[2], p4[2], pf[2]) + gammay,
          col=col, border=NA)
}
```

```{r}
plot_vertex_edges <- function(v, lwd=2, ucol=c_mid_teal,
                              delta=0.1, eps=0.05,
                              theta=60, gamma=0.025) {
  points(v$pos[1], v$pos[2],
         pch=16, col=c_light_teal)

  he <- v$he

  while(1) {
    if (is.null(he$origin)) p1 <- he$pseudo_origin
    else                    p1 <- he$origin$pos

    if (is.null(he$twin$origin)) p2 <- he$twin$pseudo_origin
    else                         p2 <- he$twin$origin$pos

    plot_half_edge(p1, p2,
                   lwd, ifelse(he$visited, c_mid, ucol),
                   delta, eps, theta, gamma)

    he <- he$prv$twin
    if (is.null(he) || he == v$he) break
  }
}
```

```{r}
plot_face_boundary <- function(f, lwd=2, col=c_mid_teal,
                               delta=0.1, eps=0.05,
                               theta=60, gamma=0.025) {
  he <- f$he

  while (1) {
    points(he$origin$pos[1], he$origin$pos[2],
           pch=16, col=col)

    plot_half_edge(he$origin$pos, he$twin$origin$pos,
                   lwd, col,
                   delta, eps, theta, gamma)

    he <- he$nxt
    if (is.null(he) || he == f$he) break
  }
}
```

```{r}
plot_face <- function(f, col=c_mid_teal) {
  # Trace boundary
  he <- f$he

  perimeter_xs <- c()
  perimeter_ys <- c()

  while (1) {
    perimeter_xs <- c(perimeter_xs, he$origin$pos[1])
    perimeter_ys <- c(perimeter_ys, he$origin$pos[2])

    if (is.null(he$nxt)) {
      perimeter_xs <- c(perimeter_xs, he$twin$origin$pos[1])
      perimeter_ys <- c(perimeter_ys, he$twin$origin$pos[2])
      break
    }

    he <- he$nxt

    if (he == f$he) {
      # Close the perimeter
      perimeter_xs <- c(perimeter_xs, he$origin$pos[1])
      perimeter_ys <- c(perimeter_ys, he$origin$pos[2])
      break
    }
  }

  polygon(perimeter_xs, perimeter_ys, col=col, border=NULL)
}
```

## Running Fortune's Algorithm

Alright, let's cut the pedagogical tension and finally build a
Voronoi diagram.

We start by defining a collection of sites.

```{r}
N <- 10

set.seed(54482248)
xs <- runif(N, -1, 1)
ys <- runif(N, -1, 1)


par(mfrow=c(1, 1), mar=c(5, 5, 4, 1))
plot(xs, ys, col=c_dark, pch=16,
     xlim=c(-1.1, 1.1), ylim=c(-1.1, 1.1))
```

Next we initialize the event queue and beach line tree.

```{r}
queue <- event_queue$new(1:N, xs)

bl <- beachline$new()
voronoi <- dcel$new()

cat('Initial queue:\n')
for (i in queue$idxs) {
  arc <- queue$events[[i]]$arc
  if (is.null(queue$events[[i]]$arc)) {
    cat(paste('  S =', sprintf('%.3f', queue$events[[i]]$x), ':',
              'Site Event at', queue$events[[i]]$site), '\n')
  } else {
    cat(paste('  S =', sprintf('%.3f', queue$events[[i]]$x), ':',
              'Vertex Event at', queue$events[[i]]$arc$site), '\n')
  }
}
```

We can completely exhaust the event queue by running.

```{r}
#| eval: false
while (queue$is_not_empty()) {
  bl$process_next_event(queue, voronoi)
}
```

Here, however, we'll proceed a little bit more deliberately.
This function processes the next event with extreme verbosity,
communicating the type of the event, visualizing both the current
beach line tree and the current beach line, and then displaying
the status of the event queue.

```{r}
process_next_event <- function() {
  if (is.null(queue$next_event()$arc)) {
    cat(paste0('Processing Site Event (',
               queue$next_event()$site, ')\n'))
  } else {
    cat(paste('Processing Vertex Event (',
              queue$next_event()$arc$site, ')\n'))
  }

  bl$process_next_event(queue, voronoi)

  par(mfrow=c(2, 1))
  plot_tree(bl, text_cex=0.6)
  if (is.null(queue$next_event()$x)) {
    plot_beachline(bl, 2,
                   c(-4, 2), c(-2, 2))
    plot_beachline_half_edges(bl, S, col=c_dark_teal,
                              delta=0.08, eps=0.04, gamma=0.02)
    for (v in voronoi$vertices)
      plot_vertex_edges(v, ucol=c_light_teal,
                        delta=0.08, eps=0.04, gamma=0.02)
  } else {
    plot_beachline(bl, queue$next_event()$x,
                   c(-4, 2), c(-2, 2))
    plot_beachline_half_edges(bl, S, col=c_dark_teal,
                              delta=0.08, eps=0.04, gamma=0.02)
    for (v in voronoi$vertices)
      plot_vertex_edges(v, ucol=c_light_teal,
                        delta=0.08, eps=0.04, gamma=0.02)
  }

  cat('\n')
  cat('Remaining queue:\n')
  for (i in queue$idxs) {
    arc <- queue$events[[i]]$arc
    if (is.null(queue$events[[i]]$arc)) {
      cat(paste('  S =', sprintf('%.3f', queue$events[[i]]$x),
                ':',
                'Site Event at', queue$events[[i]]$site),
                '\n')
    } else {
      cat(paste('  S =', sprintf('%.3f', queue$events[[i]]$x),
                ':',
                'Vertex Event at', queue$events[[i]]$arc$site),
                '\n')
    }
  }
}
```

Initially the beach line is empty.  The first site event
populates the beach line with a single parabola.

```{r}
process_next_event()
```

The next site event splits this monolithic parabola up into three
arcs and introduces the first Voronoi half-edges.  Note that the
lower arc is too small to be seen in this plot.

```{r}
process_next_event()
```

At this point we'll jump through the next few events until we're
about to encounter our first vertex event.

```{r}
for (i in 1:3) {
  bl$process_next_event(queue, voronoi)
}

par(mfrow=c(2, 1))

plot_tree(bl, text_cex=0.6)

plot_beachline(bl, queue$next_event()$x,
               c(-4, 2), c(-2, 2))
plot_beachline_half_edges(bl, S, col=c_dark_teal,
                          delta=0.08, eps=0.04, gamma=0.02)
for (v in voronoi$vertices)
  plot_vertex_edges(v, ucol=c_light_teal,
                    delta=0.08, eps=0.04, gamma=0.02)

cat('')
cat('Remaining queue:')
for (i in queue$idxs) {
  arc <- queue$events[[i]]$arc
  if (is.null(queue$events[[i]]$arc)) {
    cat(paste('  S =', sprintf('%.3f', queue$events[[i]]$x),
              ':',
              'Site Event at', queue$events[[i]]$site),
              '\n')
  } else {
    cat(paste('  S =', sprintf('%.3f', queue$events[[i]]$x),
              ':',
              'Vertex Event at', queue$events[[i]]$arc$site),
              '\n')
  }
}
```

In the first vertex event we start paring the beach line down.
Here the arc node generated by site $2$ that sits between the
breakpoints $b_{32}$ and $b_{27}$ is removed while our first
Voronoi vertex is created.

```{r}
process_next_event()
```

To not draw this process out _too_ long let's run Fortune's
algorithm to completion.

```{r}
while (queue$is_not_empty()) {
  bl$process_next_event(queue, voronoi)
}
```

The remaining breakpoint nodes each contain a dangling Voronoi
half-edge.

```{r}
par(mfrow=c(2, 1))
plot_tree(bl, text_cex=0.6)

plot_beachline(bl, 2,
               c(-4, 2), c(-2, 2))
plot_beachline_half_edges(bl, S, col=c_dark_teal,
                          delta=0.08, eps=0.04, gamma=0.02)
for (v in voronoi$vertices)
  plot_vertex_edges(v, ucol=c_light_teal,
                    delta=0.08, eps=0.04, gamma=0.02)
```

To visualize the entire Voronoi graph within the confines of a
finite plot we'll need a bounding box.

We can always specify a bounding box of our own.  This can be
useful, for example, if we want particular large or symmetric
bounding boxes for visualization purposes.

```{r}
#| eval: false
voronoi$xlim <- c(-3, 3)
voronoi$ylim <- c(-3, 3)
```

In most cases, however, it's most convenient to dynamically
adapt a bounding box to the Voronoi vertices.

```{r}
voronoi$compute_bounding_box(0.25)
```

Given a bounding box geometry we can anchor dangling half-edges
to pseudo-vertices on the bounding box.

```{r}
voronoi$anchor_dangling_edges(bl$root, xs, ys)
```

Perhaps the simplest way to plot the Voronoi graph is to plot
each Voronoi vertex along with the half-edges originating from
and terminating at those vertices.

```{r}
par(mfrow=c(1, 1), mar=c(0, 0, 0, 0))
plot(xs, ys, axes=FALSE, ann=FALSE, col=c_dark, pch=16,
     xlim=voronoi$xlim, ylim=voronoi$ylim,
     main="")

for (n in 1:N)
  text(xs[n], ys[n] - 0.1, n)

for (v in voronoi$vertices)
  plot_vertex_edges(v, delta=0.08, eps=0.04, gamma=0.0075)

for (pv in voronoi$pseudo_vertices)
  plot_vertex_edges(pv, delta=0.08, eps=0.04, gamma=0.0075)
```

The real content of the Voronoi diagram, however, is in the
faces.

```{r}
voronoi$find_faces()
```

```{r}
par(mfrow=c(1, 1), mar=c(0, 0, 0, 0))
plot(xs, ys, axes=FALSE, ann=FALSE, col=c_light, pch=16,
     xlim=voronoi$xlim, ylim=voronoi$ylim,
     main="")

for (n in 1:N)
  text(xs[n], ys[n] - 0.1, n)

for (f in voronoi$faces) {
  plot_face_boundary(f, col=c_light_teal,
                     delta=0.08, eps=0.04, gamma=0.01)
}

f <- voronoi$faces[[1]]
plot_face(f, col=c_light_teal)
plot_face_boundary(f, col=c_dark_teal,
                   delta=0.08, eps=0.04, gamma=0.01)
points(xs[f$site], ys[f$site], col=c_dark, pch=16)
text(xs[f$site], ys[f$site] - 0.1, f$site)
```

The beautiful thing about Fortune's algorithm is that scaling up
to more sights is not particularly expensive.

```{r}
N <- 1000

xs <- runif(N, -1, 1)
ys <- runif(N, -1, 1)


queue <- event_queue$new(1:N, xs)

bl <- beachline$new()
voronoi <- dcel$new()


while (queue$is_not_empty()) {
  bl$process_next_event(queue, voronoi)
}
```

```{r}
voronoi$compute_bounding_box(0.25)

voronoi$anchor_dangling_edges(bl$root, xs, ys)
```

```{r}
voronoi$find_faces()
```

```{r}
par(mfrow=c(1, 1), mar=c(0, 0, 0, 0))

plot(xs, ys, col=c_light, pch=16,
     xlim=c(-1, 1), ylim=c(-1, 1),
     main="")

for (f in voronoi$faces) {
  plot_face(f, col=c_light_teal)
}
```

# Conclusion

Voronoi diagrams are straightforward to define but subtle to
construct in practice, especially if we want to maximize
computational efficiency.  Fortune's algorithm systematizes the
construction of Voronoi diagrams, but understanding how the
algorithm actually works requires immersing ourselves in planar
geometry.  Moreover, effectively implementing Fortune's algorithm
in practice requires familiarity with a variety of nontrivial
data structures.

From a more optimistic perspective, Fortune's algorithm provides
a wealth of learning opportunities...

# Acknowledgements {-}

A very special thanks to everyone supporting me on Patreon:
Adam Fleischhacker, Alejandro Morales, Alessandro Varacca,
Alex D, Alexander Noll, Amit, Andrea Serafino, Andrew Mascioli,
Andrew Rouillard, Ara Winter, Ari Holtzman, Austin Rochford,
Aviv Keshet, Avraham Adler, Ben Matthews, Ben Swallow,
Benoit Essiambre, boot, Brendan Galdo, Bryan Chang,
Brynjolfur Gauti Jónsson, Cameron Smith, Canaan Breiss,
Cat Shark, Cathy Oliveri, Charles Naylor, Chase Dwelle,
Chris Jones, Christina Van Heer, Christopher Mehrvarzi,
Colin Carroll, Colin McAuliffe, Damien Mannion, dan mackinlay,
Dan W Joyce, Dan Waxman, Dan Weitzenfeld, Daniel Edward Marthaler,
Danny Van Nest, David Burdelski, Doug Rivers, Dr. Jobo,
Dr. Omri Har Shemesh, Dylan Maher, Dylan Spielman, Ed Cashin,
Edgar Merkle, Edoardo Marcora, Eric LaMotte, Erik Banek,
Eugene O'Friel, Felipe González, Fergus Chadwick, Finn Lindgren,
Francesco Corona, Geoff Rollins, Granville Matheson,
Gregor Gorjanc, Guilherme Marthe, Håkan Johansson,
Hamed Bastan-Hagh, haubur, Hector Munoz, Henri Wallen, hs,
Hugo Botha, Ian Costley, idontgetoutmuch, Ignacio Vera,
Ilaria Prosdocimi, Isaac Vock, Isidor Belic, jacob pine,
Jair Andrade, James C, James Hodgson, James Wade, Janek Berger,
Jarrett Byrnes, Jason Martin, Jason Pekos, Jason Wong, jd, Jed ,
Jeff Burnett, Jeff Dotson, Jeff Helzner, Jeffrey Erlich,
Jerry Lin , Jessica Graves, Joe Sloan, John Flournoy,
Jonathan H. Morgan, Josh Knecht, JU, Julian Lee, June,
Justin Bois, Kádár András, Karim Naguib, Karim Osman,
Konstantin Shakhbazov, Kristian Gårdhus Wichmann, Lars Barquist,
lizzie , LOU ODETTE, Luís F, Mads Christian Hansen,
Marek Kwiatkowski, Mariana Carmona, Mark Donoghoe, Markus P.,
Márton Vaitkus, Matěj, Matthew, Matthew Kay, Matthieu LEROY,
Mattia Arsendi, Maurits van der Meer, Max, Michael Colaresi,
Michael DeWitt, Michael Dillon, Michael Lerner, Mick Cooney,
MisterMentat , N Sanders, N.S. , Name, Nathaniel Burbank,
Nicholas Cowie, Nick S, Nikita Karetnikov, Octavio Medina,
Ole Rogeberg, Oliver Crook, Olivier Ma, Patrick  Kelley,
Patrick Boehnke, Pau Pereira Batlle, Peter Johnson,
Pieter van den Berg , ptr, quasar, Ramiro Barrantes Reynolds,
Raúl Peralta Lozada, Ravin Kumar, Rex, Riccardo Fusaroli,
Richard Nerland, Robert Frost, Robert Goldman, Robert kohn,
Robin Taylor, Ryan Gan, Ryan Grossman, S Hong, Sean Wilson,
Sergiy Protsiv, Seth Axen, shira, Simon Duane, Simon Lilburn,
Simon Steiger, Simone, Spencer Boucher, sssz, Stefan Lorenz,
Stephen Lienhard, Steve Harris, Steven Forrest, Stew Watts,
Stone Chen, Susan Holmes, Svilup, Tate Tunstall, Tatsuo Okubo,
Teresa Ortiz, Theodore Dasher, Thomas Siegert, Thomas Vladeck,
Tobychev , Tomas Capretto, Tony Wuersch, Virgile Andreani,
Virginia Fisher, Vitalie Spinu, Vladimir Markov,
VO2 Maximus Decius, Will Farr, Will Lowe, Will Wen, woejozney,
yolhaj , yureq , Zach A, and Zhengchen Cai.

# References {-}

<div id="refs"></div>

# License {-}

A repository containing all of the files used to generate this chapter is
available on [GitHub](
https://github.com/betanalpha/quarto_misc/tree/main/voronoi).

The code in this case study is copyrighted by Michael Betancourt and licensed
under the new BSD (3-clause) license:

[https://opensource.org/licenses/BSD-3-Clause](https://opensource.org/licenses/BSD-3-Clause)

The text and figures in this chapter are copyrighted by Michael Betancourt
and licensed under the CC BY-NC 4.0 license:

[https://creativecommons.org/licenses/by-nc/4.0/](https://creativecommons.org/licenses/by-nc/4.0/)

# Original Computing Environment {-}

```{r, comment=NA}
sessionInfo()
```
