---
title: "Voronoi To The World"
author: "Michael Betancourt"
date: "November 2025"
date-format: "MMMM YYYY"
toc: true
number-sections: true
highlight: pygments
bibliography: voronoi.bib
crossref:
  lst-title: "Stan Program"
filters:
   - include-code-files
format:
  html:
    html-math-method: katex
    theme:
      - lux
      - custom.scss
    standalone: true
    embed-resources: true
    code-overflow: wrap
    linkcolor: "#B97C7C"
  pdf:
    keep-tex: true
    fig-width: 4
    fig-height: 4
    code-overflow: wrap
    monofontoptions:
      - Scale=0.5
  epub:
    css: custom.scss

knitr:
  opts_chunk:
    comment: ''
  opts_knit:
    global.par: TRUE
format-links: false
---

Voronoi diagrams arise naturally in many spatial analyses.  More
importantly for this author they also tend to be aesthetically
attractive, making them useful for crafting
compelling illustrations.

Efficiently constructing a Voronoi diagram, however, is not
particularly straightforward.  In this note we'll work through
the geometric structure of Voronoi diagrams and an algorithm that
leverages that structure to implement Voronoi diagrams with the
maximum possible performance.

# Tessellating A Plane { #sec:voronoi-def }

A **Voronoi diagram**, also known as a
**Dirichlet tessellation**, is an organization of a plane into
subsets defined by the nearest proximity to given collection of
points.

From what I can tell the standard reference for Voronoi diagrams,
and other topics in computational geometry, is @deBergEtAl:2008.
This text is sometimes referred to as "the 4Ms" which I can only
infer refers to three out of the four authors having the first
name "Mark" with the forth, non-Mark author being an honorary
"M".  I humbly suggest that "Mark, Marky, Mark, and the Funky
Bunch" is more appropriate, but to each their own.

To define a Voronoi diagram formally consider a two-dimensional
Euclidean plane $\mathbb{R}^{2}$ and a collection of **sites**
$$
( \, s_{1} = (x_{1}, y_{1}), \ldots, s_{N} = (x_{N}, y_{N}) \, )
\subset \mathbb{R}^{2}.
$$
Give a particular site $s_{n}$ we can construct a subset of
points, also known as a **cell**, that are closer to that site
than any other site,
$$
\mathsf{c}_{n} =
\{ p \in \mathbb{R}^{2} \mid
d( p, s_{n} ) < d( p, s_{n'} ) \; \mathrm{for all} \; n' \ne n \},
$$
where $d(p, p')$ is a distance function.  The Voronoi diagram of
a collections of sites is the corresponding collection of
nearest-proximity subsets (@fig-voronoi).

::: {#fig-voronoi layout="[50, 50]"}
![](figures/voronoi/cells/voronoi){#fig-voronoi-cell}

![](figures/voronoi/boundary/voronoi){#fig-voronoi-boundary}

\(a) A Voronoi diagram defined by a collection of sites
$( s_{1}, \ldots, s_{N} )$, here shown in red, is a collection of
subsets, each containing points that are closer to one site than
the others.  (b) The boundary between these subsets consists of
points that are equally distant to two or more sites at the same
time.
:::

The standard Voronoi diagram considers the Euclidean distance
function,
$$
d( p, p' ) = \sqrt{ (x - x')^{2} + (y - y')^{2} }.
$$
This construction, however, can be generalized to other distance
functions.  At the same time Voronoi diagrams can be defined on
higher-dimensional real spaces and even more general metric
spaces.

A Voronoi diagram _almost_ forms partition of a Euclidean plane.
The obstruction is that points that are equally distant from two
or more sites don't fall into any of the cells, but rather form a
singular boundary between the cells.  The combination of the
Voronoi cells and the boundary between them, however, does form
a proper partition.

Because each cell contains one, and only one, site, we can always
label the cells by the associated site (@fig-cell-label).
Similarly we can label the linear segment that forms the boundary
between two neighboring cells by the sites associated with those
two cells, and the punctual boundaries separating the corners of
three neighboring cells by the sites associated with those cells
(@fig-line-label, @fig-line-label).

::: {#fig-voronoi-labels layout="[-0.5, 33, 33, 33, -0.5]"}
![](figures/voronoi/cell_labels/labels){#fig-cell-label}

![](figures/voronoi/boundary_labels/line/labels){#fig-line-label}

![](figures/voronoi/boundary_labels/point/labels){#fig-point-label}

\(a) Each cell comprising a Voronoi diagram can be labeled with
the unique site that it contains.  (b, c) Similarly the
boundaries between neighboring cells can be labeled with the
sites contained by those cells.  For the purposes of labeling the
the order of the sites is arbitrary, so $(3, 2)$ is just as good
as $(2, 3)$ while $(2, 7, 3)$, $(3, 7, 2)$, $(3, 2, 7)$,
$(7, 3, 2)$, and $(7, 2, 3)$ are just as good as $(2, 3, 7)$.
:::

# The Geometry of Voronoi Diagrams

The boundary between the cells of a Voronoi diagram is useful in
its own right.  In particular it defines an undirected graph,
with linear edges separating neighboring pairs of cells and
punctual vertices separating neighboring triplets of cells.

Often it is easier in practice to construct this
**Voronoi graph** and then derive the Voronoi cells from it
rather than trying to construct the Voronoi cells directly.
The vertices and edges of the Voronoi graph exhibit rich
geometric structure that provides the foundation for powerful
algorithms.

## Voronoi Vertices

The vertices of the Voronoi graph are defined by points that are
the same distance from three neighboring Voronoi cells
(@fig-vertex).  Consequently we can uniquely label each vertex
with a triplet the sites contained by those cells,
$$
(s_{n_{1}}, s_{n_{2}}, w_{n_{3}} ).
$$
That said, not every triplet of Voronoi cells are neighbors and
not every triplet of cells defines a valid vertex.  Fortunately
Euclidean geometry provides tools for identifying all of the
valid vertices.

![Each vertex in a Voronoi graph is given by a point on the
boundary between three neighboring Voronoi cells.  This point is
by definition equally distant to the three sites contained by
those cells.](
figures/voronoi/vertex/vertex_distances/vertex){
width=50% #fig-vertex}

Each triplet of sites falls onto a unique **circumcircle** with
center
\begin{align*}
x_{c}
&=
\frac{1}{2}
\frac{
  \left( x_{n_{1}}^{2} + y_{n_{1}}^{2} \right) \,
    (y_{n_{2}} - y_{n_{3}})
+ \left( x_{n_{2}}^{2} + y_{n_{2}}^{2} \right) \,
    (y_{n_{3}} - y_{n_{1}})
+ \left( x_{n_{3}}^{2} + y_{n_{3}}^{2} \right) \,
    (y_{n_{1}} - y_{n_{2}})
}{
  x_{n_{1}} \, (y_{n_{2}} - y_{n_{3}} )
+ x_{n_{2}} \, (y_{n_{3}} - y_{n_{1}} )
+ x_{n_{3}} \, (y_{n_{1}} - y_{n_{2}} )
}
\\
y_{c}
&=
\frac{1}{2}
\frac{
  \left( x_{n_{1}}^{2} + y_{n_{1}}^{2} \right) \,
    (x_{n_{3}} - x_{n_{2}})
+ \left( x_{n_{2}}^{2} + y_{n_{2}}^{2} \right) \,
    (x_{n_{1}} - x_{n_{3}})
+ \left( x_{n_{3}}^{2} + y_{n_{3}}^{2} \right) \,
    (x_{n_{2}} - x_{n_{1}})
}{
  x_{n_{1}} \, (y_{n_{2}} - y_{n_{3}} )
+ x_{n_{2}} \, (y_{n_{3}} - y_{n_{1}} )
+ x_{n_{3}} \, (y_{n_{1}} - y_{n_{2}} )
}
\end{align*}
and radius
\begin{align*}
r
&=
\sqrt{ (x_{c} - x_{n_{1}})^{2} + (y_{c} - y_{n_{1}})^{2} }
\\
&=
\sqrt{ (x_{c} - x_{n_{2}})^{2} + (y_{c} - y_{n_{2}})^{2} }
\\
&=
\sqrt{ (x_{c} - x_{n_{3}})^{2} + (y_{c} - y_{n_{3}})^{2} }.
\end{align*}
By construction the center of the circumcircle is equidistant
from all three sites, and hence defines a potential vertex in the
Voronoi graph.

The three sites defining the circumcircle are the three
_closest_ sites to this potential vertex if and only if no other
sites are inside of the circumcircle.  In other words Voronoi
vertices are defined by _empty_ circumcircles.

::: {#fig-vertex-circum layout="[-17, 33, 33, -17]"}
![](figures/voronoi/vertex/vertex_good/vertex_good){#fig-vertex-circum-good}

![](figures/voronoi/vertex/vertex_bad/vertex_bad){#fig-vertex-circum-bad}

Not every triplet of sites defines a vertex in the Voronoi graph.
(a) Vertices are defined by triplets of sites that define empty
circumcircless.  (b) Triplets of sites whose circumcircles
contain other sites do not define proper vertices.
:::

## Voronoi Edges

The edges in a Voronoi graph are comprised of points equidistant
to two neighboring cells (@fig-edge), allowing us to uniquely
label each Voronoi edge with the pair of sites contained by those
cells.  Once, again, however, we have to take care because
arbitrary pairs of sites do not always define valid edges.

![Each edge in a Voronoi graph is defined by linear boundary
between two neighboring Voronoi cells.  All of the points on a
Voronoi edge is equally distant to the two sites contained by
those cells.](
figures/voronoi/edge/edge_distances/edge){
width=50% #fig-edge}

Before worrying out isolating proper edges let's first work out
how to identify the points that define potential edges.  Any pair
of sites defines a line that connects them, and a mid point on
that line that is equally distant to the sites.  The
**perpendicular bisector** of two sites is the line perpendicular
to this connecting line that intersects with the midpoint
(@fig-bisector).  Every point along the perpendicular bisector
between two sites is equally distant from those two sites and,
consequently, could contribute to a Voronoi edge
(@fig-bisector-edge).

::: {#fig-potential-edge layout="[-17, 33, 33, -17]"}
![](figures/voronoi/edge/bisector/bisector){#fig-bisector}

![](figures/voronoi/edge/bisector_edge/edge){#fig-bisector-edge}

\(a) The perpendicular bisector between two points is the line
$\ell_{\perp}$ that is perpendicular to line connecting those
points $\ell$ and intersects the midpoint between them. (b) Every
Voronoi edge falls along a perpendicular bisector between two
sites, but not every perpendicular bisector contains a Voronoi
edge and not every point of a perpendicular bisector that does
contain an edge is part of that edge.
:::

In order to determine which segment, if any, of a perpendicular
bisector forms a Voronoi edge we come back to circles.  Any point
along a perpendicular bisector is the center of a unique circle
that includes the two defining sites.  If no _other_ sites are on
or inside of this circle then those two sites are the nearest
neighbors to each other and that point is part of a Voronoi edge.

As we move along the perpendicular bisector these circles will
eventually intersect with another site.  When they do the point
along the bisector is equally distant to three sites -- the two
initial sites and this new site -- but the interior of the circle
will remain empty.  In other words the Voronoi edge terminates
as soon as we hit a Voronoi vertex.

::: {#fig-edge-circles layout="[-0.5, 33, 33, 33, -0.5]"}
![](figures/voronoi/edge/circle1/edge){#fig-edge-circle1}

![](figures/voronoi/edge/circle2/edge){#fig-edge-circle2}

![](figures/voronoi/edge/circle3/edge){#fig-edge-circle3}

Circles centered at points along a perpendicular bisector and
intersecting with a pair of sites are useful for identifying
contributions to Voronoi edges.  (a) If there are no other sites
on or inside of the circle then the center contributes to a
Voronoi edge.  (b) If there is one or more sites inside of the
circle then the center does not contribute to a Voronoi graph at
all.  (c) If there is another site on the circle then the center
defines a Voronoi vertex.  Each Voronoi edge terminates when
these circles first meet a third site.
:::

## Voronoi Faces

The edges of a Voronoi graph trace around each site, defining
faceted **faces** in the graph that correspond to the Voronoi
cells (@fig-faces).  When working over the entire, unbounded,
Euclidean plane not all of these faces will be closed.  While
some of the sites will be entirely fully enclosed by a path of
edges, some will be bounded on only one side by a path of edges.
I will refer to these as closed faces and open faces,
respectively.

::: {#fig-faces layout="[-17, 33, 33, -17]"}
![](figures/voronoi/faces/closed/voronoi){#fig-closed-face}

![](figures/voronoi/faces/open/voronoi){#fig-open-face}

The edges of a Voronoi graph trace out faces which identify the
Voronoi cells.  (a) I will refer to faces that are completely
enclosed by edges as closed faces and (b) faces that are only
partially enclosed as open faces.
:::

# Fortune's Algorithm

There are here are multiple ways to construct a Voronoi graph,
most of them suffer from wasteful computation.  For large enough
numbers of sites these approaches quickly become too expensive to
be practical.  Fortunately there is one elegant approach that,
while not at all initially obvious, is able to construct a
Voronoi graph as efficiently as possible.

## Inefficient Approaches

Before diving into efficient solutions let's consider some of the
more straightforward but ultimately inefficient approaches.

One way to identify all of the vertices in a Voronoi graph is to
loop over each triplet of sites and then construct the
corresponding circumcircles.  For each of these circumcircles we
then loop over the remaining sites to check for inclusion, with
the centers of the empty circumcircles defining vertices.

We could then construct candidate Voronoi edges by looping over
each pair of sites and scanning across the points between the two
Voronoi vertices associated with those cites.  Next we could take
any point along each candidate edge, construct a circle centered
on that point and intersecting with the two associated sites,
and check if any other sites are inside.  The candidates with
empty circles define Voronoi edges.

Because there are
$$
{ N \choose 3 }
=
\frac{ N \, (N - 1) \, (N - 2) }{ 6 }
$$
triplets of sites and
$$
{ N \choose 2 }
=
\frac{ N \, (N - 1) }{ 2 }
$$
pairs of sites to consider, however, the number of operations we
need to implement this brute-force approach scales _cubically_
with the number of sites.

That said, much of this cost is wasted on the repeated querying
of each site; we should be able to construct Voronoi graphs more
efficiently if we can query each site fewer times.  One more
efficient approach is to loop over each site, construct the
half-planes defined by the perpendicular bisector between the
active site and each other site, and then construct the
corresponding Voronoi face from the intersection of those
half-planes.  This cost of this approach requires only quadratic
number of operations.

We can, however, do even better.  By systematically scanning
across a plane **Fortune's algorithm**
[@Fortune:1987, @deBergEtAl:2008] is able construct Voronoi graph
with a number of operations proportional to $N \log N$.  This
performance turns out to be theoretically optimal.  The main
downside of Fortune's algorithm is that is requires some care to
understand and then implement in a way that achieves that optimal
scaling.

## The Sweep Line

Fortune's algorithm processes the sites not by working through a
list but rather by systematically scanning across the ambient
plane with the use of a **sweep line**.  Sites on one side of
the sweep line are active and can contribute to the construction
of the Voronoi graph while sites on the other side are inactive
and cannot yet contribute.  As the sweep line progresses more
sites are processed and more of the Voronoi graph is built up.

In theory a sweep line can progress in any direction, but most
references consider vertical lines that scan horizontally from
left to right, horizontal lines that scan vertically from top to
bottom, or horizontal lines that scan vertically from bottom to
top.  These axis-aligned orientations of the sweep line make
geometric calculations substantially more straightforward.

Here I will consider a vertical sweep line that scans
horizontally from left to right.  I will denote the horizontal
position of the scan at any time by $S$.  I will also refer to
any site to the left of the sweep line, $x_{n} < S$ as active.

## The Beach Line

A sweep line partitions the ambient plane into two pieces,
allowing the construction of the Voronoi graph to focus on just
a subset of the entire plane.  That said, not all of the Voronoi
graph to the left of the sweep line is completely determined by
the active sites; some features of the Voronoi diagram can be
influenced by sites just beyond the sweep line.

Consider, for example, a single active site
$$
s_{n} = (x_{n}, y_{n})
$$
and the sweep line position
$$
S > x_{n}.
$$
If a point $(x, y)$ on the left of the sweep line is closer to
$s_{n}$ than the sweep line,
$$
(x - x_{n})^{2} - (y - y_{n})^{2} < (S - x)^{2},
$$
then we it will be closer to $s_{n}$ than any other site that
might be hiding beyond the sweep line.  On the other any point to
the left of the sweep line that is closer to the sweep line than
it is to $s_{n}$,
$$
(x - x_{n})^{2} - (y - y_{n})^{2} > (S - x)^{2},
$$
could be be closer to an inactive site on the other side of the
sweep line.

When determining geometric features such as the nearest sites we
cannot analyze the entire region to the left of the sweep line.
Instead we can analyze only the subset of points that are closer
to $s_{n}$ than they are to the sweep line.  The boundary of this
analyzable subset is defined by the points equidistant to $s_{n}$
and the sweep line, which traces out a rotated
parabola (@fig-parabola)
$$
x
=
f(y)
=
\frac{1}{2}
\left( S + x_{n} - \frac{( y - y_{n} )^{2}}{S - x_{n}} \right).
$$

![The subset of points that are equally distant from a focus,
$s$, and a vertical line $S$, form a rotated parabola.  In
Fortune's algorithm the points interior to this parabola are safe
to analyze for their contribution to the Voronoi graph.](
figures/fortune/parabola_distances/parabola_distances){
width=50% #fig-parabola}

When there are multiple active sites then the region that we can
safely analyze the points that are closer to _any_ of the active
sites than the sweep line.  The boundary of this region is
defined by the segmented envelope of the parabolas between each
active site and the sweep line (@fig-beachline).  Given its
resemblance to the shape made by waves washing up on a beach
(@fig-actual-beachline) this boundary is known as the
**beach line**.

![The beach line is the boundary separating points that are
closer to any active site, here $s_{n_{1}}$, $s_{n_{2}}$, and
$s_{n_{3}}$, and points that are closer to the the sweep line,
$S$, is comprised of multiple parabolic arcs.  Note that the
parabola between $s_{n_{1}}$ and $S$ contributes multiple arcs to
the beach line here.](
figures/fortune/beachline/beachline){
width=50% #fig-beachline}

![The parabolic arcs that make up a beach line are similar to the
shape made by waves washing up onto a beach, hence the name.](
figures/fortune/actual_beachline/actual_beachline){
width=100% #fig-actual-beachline}

Each parabolic segment, or **arc**, in a beach line is associated
with a single site.  The parabola between each site and the sweep
line, however, can contribute to the beach line multiple times.
Consequently we cannot use the originating sites to _uniquely_
label the arcs.

By construction neighboring arcs on the beach line are derived
from neighboring sites.  The discontinuities along the beach line
that separate each pair of neighboring arcs are known as
**breakpoints** (@fig-breakpoints).  Each breakpoint is uniquely
labeled by an ordered pair of the sites associated with those
arcs.  Ordering is important here: $(s_{n_{1}}, s_{n_{2}})$ and
$(s_{n_{2}}, s_{n_{1}})$ correspond to two _different_
breakpoints.

![The breakpoints in a beach line separate the component arcs.
Each breakpoint is located at the intersection between the
parabolas defined by two neighboring sites, equally distant from
those two sites and further from any other sites.  We can use an
ordered pair of these sites to uniquely label each breakpoint.](
figures/fortune/breakpoints/breakpoints){
width=50% #fig-breakpoints}

Given a sweep line configuration we can readily calculate the
intersection of each pair of neighboring parabolas and hence the
position of any given breakpoint.  Formally if the lower arc
is a segment of the parabola generated by $s_{n_{1}}$ and the
higher arc is a segment of the parabola generated by $s_{n_{2}}$
then the vertical position of the breakpoint between those arcs
is given by one of the two solutions to the quadratic equation
\begin{align*}
0
&=\quad
\Bigg( x_{n_{2}} - x_{n_{1}} \Bigg) \, y_{n_{1}n_{2}}^{2}
\\
&\quad
+
\Bigg(  y_{n_{2}} \, (S - x_{n_{1}})
      - y_{n_{1}} \, (S - x_{n_{2}}) \Bigg) \, y_{n_{1}n_{2}}
\\
&\quad
+
\Bigg(  y_{n_{2}}^{2} \, (S - x_{n_{1}})
      - y_{n_{1}}^{2} \, (S - x_{n_{2}})
      - (S - x_{n_{1}}) \,
        (S - x_{n_{2}}) \,
        (x_{n_{2}} - x_{n_{1}} ).
\Bigg)
\end{align*}
The correct solution is the one that falls in between the height
of the two neighboring sites,
$$
y_{n_{1}} < y_{n_{1}n_{2}} < y_{n_{2}}.
$$

When it comes to constructing Voronoi graphs the breakpoints are
the most important part of the beach line.  Because each
breakpoint falls onto the parabolas of two neighboring sites, it
is equally distant from those two sites and the sweep line.  At
the same time no other active site is closer than those two
sites, otherwise the parabolas generated by them would not define
neighboring arcs on the beach line.  In addition every inactive
site is also further because they're all beyond the sweep line.

All of this is to say that, by construction, each breakpoint
is the center of a circle that intersects with the two associated
sites but includes no other sites.  Consequently each breakpoint
_is always a part of a Voronoi edge_.

## Evolving The Beach Line

As we scan the sweep line across a Euclidean plane the distances
between the active sites and the sweep line change.  This in
turn changes the shape of the corresponding parabolas, and hence
the shape of the beach line.  As the beach line evolves in this
way the breakpoints trace out the edges of the Voronoi diagram.

Every now and then the number of arcs comprising the beach line,
referred to as the **topology** of the beach line, also changes.
When the sweep line passes over a new site, for example, a new
arc is added to the beach line.  Similarly arcs can be removed
from the beach line when they are squeezed to a point by their
neighboring arcs.

Conveniently we don't need to continuously scan the sweep line in
order to construct a Voronoi diagram.  All of the information we
need is completely determined by the behavior of the beach line
at the finite number of sweep line positions where the topology
changes.  If we can efficiently determine these topology-changing
**events** then we can efficiently build up a Voronoi diagram in
a finite number of operations.

### Events Where The Beach Line Expands

Whenever the sweep line crosses a new site that site becomes
active and contributes a new parabolic arc to the beach line.
Because we know the position of all of the sites in advance we
at exactly which positions these **site events** will occur.

If the sweep line falls exactly on a new site then the parabola
of points equidistant from the site and the sweep line is
singular, consisting of just the new site itself.  This forms a
point discontinuity with the rest of the beach line
(@fig-site-event1a).  Although not _technically_ correct, these
singular arcs are often visualized with a horizontal line
extending from the new site to the beach line
(@fig-site-event1b).

::: {#fig-site-event layout="[-0.5, 33, 33, 33, -0.5]"}
![](figures/fortune/site_event/1a/site_event){#fig-site-event1a}

![](figures/fortune/site_event/1b/site_event){#fig-site-event1b}

![](figures/fortune/site_event/2/site_event){#fig-site-event2}

At a site event the sweep line passes an inactive site which then
becomes active.  (a) Initially the new active site introduces a
new singular arc which forces a discontinuity in the beach line.
(b) This singular arc is often visualized with a straight line.
(c) As the sweep line proceeds past the site event the singular
arc widens into a well-behaved parabola, splitting one of the
initial beach line arcs into a triplet of arcs.
:::

As the sweep line progresses past the new active site the
singular parabola widens into a more well-behaved parabola
(@fig-site-event2).  This not only restores the continuity of the
beach line but also splits an existing parabolic arc into two new
arcs.  The introduction of a new active site always splits an
existing arc into a triplet of arcs.

For example let's say that the sweep line passes the new site
$s_{n_{2}}$.  If the height of this site $y_{n_{2}}$ intersects
with an arc generated by the site $s_{n_{1}}$ then the site event
will split that arc into a triplet of arcs, the outer two of
which are generated by $s_{n_{1}}$ and the center of which is
generated by $s_{n_{2}}$. These three arcs are separated by two
new breakpoints $b_{ n_{1} n_{2} }$ and $b_{ n_{2} n_{1} }$.

Both of the breakpoints fall onto a Voronoi edge in between the
sites $s_{n_{1}}$ and $s_{n_{2}}$.  As the sweep line progresses
the middle arc widens and the breakpoints $b_{ n_{1} n_{2} }$ and
$b_{ n_{2} n_{1} }$ separate, tracing out more and more of the
edge (@fig-site-event-breakpoints)

![As the sweep line moves past a site event the two new
breakpoints separate and scan across the Voronoi edge between the
two neighboring sites.](
figures/fortune/site_event/breakpoints/breakpoints){
width=50% #fig-site-event-breakpoints}

### Events Where The Beach Line Contracts

The topology of the beach line can also change when one of the
parabolic arcs collapses in between its two neighboring arcs
(@fig-vertex-event-collapse).  In this case the collapsed arc is
removed from the beach line and its two neighbors become
neighbors themselves.

![As the sweep line progresses arcs on the beach line can
collapse in between their two neighboring arcs, with the
neighboring breakpoints converging to a point.  This point is
equally distant from the three active sites and further from all
other sites, defining a vertex in the Voronoi graph.](
figures/fortune/vertex_event/collapse/vertex_event){
width=50% #fig-vertex-event-collapse}

Now an arc in between two arcs that are generated from the same
active site, such as the central arc introduced when the sweep
line crosses a site event, always widens faster than its
neighboring arcs as the sweep line progresses.  Consequently a
triplet of arcs generated by only two sites,
$(s_{n_{1}}, s_{n_{2}}, s_{n_{1}})$, will never collapse.  Rather
collapsing arcs are _always_ defined by triplets of arcs
generated by three distinct sites,
$(s_{n_{1}}, s_{n_{2}}, s_{n_{3}})$.

When the central arc collapses its two neighboring breakpoints
$b_{ n_{1} n_{2} }$ and $b_{ n_{2} n_{3} }$ will intersect.  At
that moment the breakpoints are equally distant from the sweep
line and all tree of the associated sites, $s_{n_{1}}$,
$s_{n_{2}}$, and $s_{n_{3}}$ .  Moreover, because this
intersection is on the beach line it cannot be closer to any
in active sites hiding beyond the sweep line.  In other words a
collapsing arc on the beach line always defines a Voronoi vertex.

This duality between arc collapses and Voronoi vertices is useful
not only for building up the Voronoi graph but also for
predicting when arcs will collapse.  At this point it should not
be a surprise that this prediction will involve circles.

For any triplet of neighboring arcs on a beach line that
correspond to the _distinct_ sites
$(s_{n_{1}}, s_{n_{2}}, s_{n_{3}})$ we can always construct a
unique circumcircle with center $(x_{c}, y_{c})$ and radius $r$.
By construction the center $(x_{c}, y_{c})$ is equally distant to
all three sites.  The center is also equally distant to vertical
lines at $x = x_{c} - r$ and $x = x_{c} + r$.  Consequently if
the central arc collapses then it will do so when the sweep
line is at either $S = x_{c} - r$ or $S = x_{c} + r$
(@fig-vertex-event-circumcircle).

![If the central arc in a triplet of neighboring arcs generated
by distinct sites will collapse then it will do so when the
sweep line is at the far edge of the circumcircle spanned by
those tree sites.  This geometric extrapolation allows us to
schedule future vertex events.](
figures/fortune/vertex_event/circumcircle/vertex_event){
width=50% #fig-vertex-event-circumcircle}

A collapse at $S = x_{c} - r$ can't contribute information about
the Voronoi graph because it is ahead of the current beach line
where the structure of the Voronoi graph has already been
established.  If the central arc collapses at this point then it
will _expand_ as the sweep line progresses further.

Instead we can focus on a new **vertex event**, also known as a
**circle event**, at $S = x_{c} + r$.  If the the central arc
generated by $s_{n_{2}}$ collapses when the sweep line reaches
this position then $b_{ n_{1} n_{2} }$ and $b_{ n_{2} n_{3} }$
will intersect at $(x_{c}, y_{c})$.  In turn this intersection
defines a new vertex that we can add to the Voronoi graph
(@fig-vertex-event-voronoi).

![When a central arc collapses two Voronoi edges terminate at a
Voronoi edge and a new Voronoi edge between the sites that
generate the remaining arcs is created.](
figures/fortune/vertex_event/voronoi/voronoi){
width=50% #fig-vertex-event-voronoi}

## The Algorithm

Now that we know how and when the topology of a beach line can
change we can organize those changes into a dynamic algorithm
that builds up a Voronoi graph dynamically.

Fortune's algorithm starts with an **event queue** consisting of
events defined by sweep line positions where an arc is added to
or removed from the beach line.  The algorithm then proceeds
iteratively, processing the next event in the queue to expand the
structure of the Voronoi graph and add new events as necessary.
Initially the event queue is filled with only site events, one
for each site, and vertex events are added as the site events are
processed.

In this section we'll review the basic steps required to process
both site and vertex events, as well as clean up after the event
queue has been fully processed.  The details depend on how we
implement Fortune's algorithm, which will be the focus of the
next section.

### Site Events

Processing a site event requires adding a new arc to the beach
line.  To do this we have to scan across the current beach line
to find the existing arc at the same height as the new event.  We
then split that arc in two, inserting a new arc generated by the
new site into the beach line along with two new breakpoints.

Inserting a new arc into the beach line can also modify the
remaining event queue.  For example inserting a new arc will
disrupt some of the triplets of neighboring arcs generated by
distinct sites in the previous beach line.  If any of those
arc triplets correspond to a future vertex events in the event
queue then they have have to be removed.

At the same time inserting a new arc also creates new triplets of
neighboring arcs, each of which could potentially spawn a vertex
event.  After checking that the individual arcs are generated by
distinct sites we have to check whether or not the central arc
actually collapses beyond the current sweep line.  If all of
these conditions are met then we add a new vertex event to the
event queue where the central arc will be removed.

### Vertex Events

When we encounter a vertex event we first have to remove the
central arc from the beach line, along with one of its
neighboring breakpoints.

At this point we expand the current Voronoi graph.  This starts
by adding a new Voronoi vertex, connecting it to two of the
previously dandling Voronoi edges, and then creating a new
Voronoi edge.  Exactly how we accomplish this depends on how the
graph is implemented.

Lastly we have to clean up the event queue.  The removal of the
central arc, for example, will invalidate any vertex event in the
remaining event queue where the central arc was previously
neighbors with the removed arc.  On the other hand the removal
also creates two new triplets of neighboring arcs along the beach
line, each of which could spawn a new vertex event.  If any of
these triplets consist of arcs generated by distinct sites with
the central arc collapses beyond the current sweep line then we
add a corresponding vertex event to the event queue.

### Clean Up

As we process the events the event queue will eventually empty.
When there are no more events to process the algorithm will have
found all of the Voronoi vertices and all of the Voronoi edges.
That said some of those edges will be dangling, anchored to a
vertex on only one side because they stretch out to infinity
(@fig-clean-up1).

::: {#fig-clean-up layout="[-0.5, 33, 33, 33, -0.5]"}
![](figures/fortune/cleanup/dangling/voronoi){#fig-clean-up1}

![](figures/fortune/cleanup/bounding_box/voronoi){#fig-clean-up2}

![](figures/fortune/cleanup/pseudo_vertices/voronoi){#fig-clean-up3}

\(a) After running Fortune's algorithm some of the edges in the
Voronoi graph will be connected to vertices on only one side.
The other side dangles out towards infinity.  (b) When
visualizing a Voronoi graph we can consider only a finite range
of values.  This requires limiting the visualization to a
bounding box that contains all of the Voronoi vertices.  (c)
For visualization purposes we can anchor the dangling edges to
points where they meet the bounding box.
:::

We can handle these dangling edges in a few different ways.  For
example from a graph theory perspective we can anchor any
dangling edges to a single "vertex at infinity" that quantifies
the common unboundedness of those edges.  This approach, however,
does not yield particularly compelling visualizations.

For visualization purposes it is more useful to create a
**bounding box** that contains all of the Voronoi vertices, and
ideally all of the sites, (@fig-clean-up2) and then anchor the
dandling edges to points along this rectangle (@fig-clean-up3).
Specifically the anchor for a dangling edge will be at the point
where the perpendicular bisector between the two sites associated
with the edge intersect with the bounding box.

# Implementing Fortune's Algorithm

Once we've built up enough familiarity with planar geometry, the
motivation for Fortune's algorithm, particularly the utility of a
beach line and its evolution, becomes less obscure.  That said,
there is still work to do to ensure that we can implement the
algorithm in a way that achieves the maximum possible
performance.

For example when processing a site event we need to be able to
efficiently search through the arcs on the current beach line.
If we naively scan through the _entire_ beach line every time we
process a new site event then the the number of operations needed
to implement Fortune's algorithm with scale with not $N \log N$
but rather $N^{2}$.  To achieve that $N \log N$ scaling we need
to implement the beach line in a way that allows for efficient
searching.

Similarly we need to keep track of the next event while we add
and delete events from the event queue.  We could accomplish this
by sorting the entire event queue after every update, but this
is wasteful and spoils the optimal $N \log N$ scaling.

Fortunately (pun very much intended) if we leverage some
standard data structures from computer science then we can avoid
these potential slowdowns and implement Fortune's algorithm as
efficiently as possible.

## Implementing Voronoi Graphs

In theory graphs are straightforward to specify with a just list
of points for each vertex and list of pairs of points for each
edge.  While these lists allow us to draw a graph easily enough,
they don't provide enough information to perform basic graph
operations, such as tracing along paths of edges to derive faces.

The **doubly-connected edge list** [@deBergEtAl:2008] is a data
structure that efficiently captures not only the components of a
graph but also the relationships between those components.  These
relationships in turn make it straightforward to efficiently
implement basic graph algorithms.

A doubly-connected edge list is itself built up from lists of
three primitive data structures: half-edges, vertices, and faces.

### Half-Edges

Doubly-connected edge lists don't represent undirected edges
directly.  Instead the data structure splits undirected edges
into two directed edges known as **half-edges**
(@fig-dcel-half-edge).

![Doubly-connected edge lists represent undirected edges with a
pair of directed half-edges.  Each half-edge includes references
to other structures in the doubly-connected edge lists, including
the vertex from which the half-edge originates and neighboring
half-edges.](
figures/dcel/half_edge/1/edge){
width=50% #fig-dcel-half-edge}

To situate them within the full graph, each half-edge is endowed
with a collection of references to other objects in the
doubly-connected edge list.  For example each half-edge includes
a reference to the vertex from which the half-edge originates.
Similarly each half-edge includes references to three other
half-edges -- the _previous_ half-edge that points into it, the
_next_ half-edge that it points into, and the _twin_ half-edge
that completes it to form a full undirected edge
(@fig-dcel-half-edge-refs).  Finally each half-edge includes a
reference to the face located to the left of the half-edge.

![Each half-edge in a doubly-connected edge lists includes
reference three other half-edges.  The previous half-edge, here
labeled "he.prv", identifies the half-edge that points into it
while the next half-edge, here labeled "he.nxt",  identifies the
half-edge that it points into.  Finally the twin half-edge, here
labeled "he.twin", identifies the half-edge that completes it to
form a full undirected edge.](
figures/dcel/half_edge/2/edge){
width=50% #fig-dcel-half-edge-refs}

These references make a variety of graph operations
straightforward to implement.  For instance the originating
vertex and twin half-edge references provide enough information
to draw a half-edge as a directed line segment, starting at the
position of the originating vertex and then ending at the
position of the twin's originating vertex.

Similarly we can use the previous and next references to trace
through paths in the graph.  These paths can then be used to
quantify topological information about the graph, such as the
configuration of the faces.

Voronoi graphs generally feature unbounded edges, with only one
side terminating at a vertex and the other shooting off towards
infinity.  In this case some of the half-edges will feature
empty next half-edge references and their twins will features
empty originating vertices and previous half-edge references.

### Vertices

Each vertex in a doubly-connected edge list is specified with its
spatial position and a reference to any one half-edge that
originates from it (@fig-dcel-vertex).  The relationship of a
vertex to the rest of graph can be completely reconstructed from
this information alone.

![Each vertx in a doubly-connected edge list includes its
spatial position, here denoted "v.pos" and a reference to any one
half-edge that originates from it, here denoted "v.he".](
figures/dcel/vertex1/vertex){
width=50% #fig-dcel-vertex}

For example all of the half-edges that originate at a given
vertex can always be accessed by repeatedly following the twin
and next references from the included half-edge
(@fig-dcel-vertex-algo).  Equivalently we can iteratively follow
the previous and then twin references.  In the same way we can
access the half-edges that _terminate_ at a given vertex.

::: {#fig-dcel-vertex-algo layout="[ [50, 50], [50, 50] ]"}
![](figures/dcel/vertex2/1/vertex){#fig-dcel-vertex-algo1}

![](figures/dcel/vertex2/2/vertex){#fig-dcel-vertex-algo2}

![](figures/dcel/vertex2/3/vertex){#fig-dcel-vertex-algo3}

![](figures/dcel/vertex2/4/vertex){#fig-dcel-vertex-algo4}

\(a) The one half-edge referenced in a vertex object is enough to
reconstruct the entire structure of the graph around that vertex.
For example if we follow the reference to (b) the previous
half-edge, "v.he.p", (c) and then its twin, "v.he.p.t", we
recover another half-edge that originates from that vertex.
(d) Repeating this operation recovers all of the half-edges that
originate from that vertex.
:::

Once we can access all of the neighboring half-edges we can then
access additional information about the local structure of the
graph.  Neighboring vertices, for instance, are immediately
given by the originating vertex references of the twin of any
half-edge that originates at a given vertex.

### Faces

Finally each face in the graph is specified by any one half-edge
along its boundary (@fig-dcel-face1).  The provides enough
information to trace out the entire boundary by following the
next or previous references from that one included half-edge
(@fig-dcel-face2, @fig-dcel-face3).

::: {#fig-dcel-face layout="[-0.5, 33, 33, 33, -0.5]"}
![](figures/dcel/face1/face){#fig-dcel-face1}

![](figures/dcel/face2/face){#fig-dcel-face2}

![](figures/dcel/face3/face){#fig-dcel-face3}

\(a) Each face in a doubly-connected edge list includes a
references to any one half-edge along its boundary.  By following
the next and previous references from that one half-edge we can
always trace out the entire boundary of the face, regardless
of whether the face is (b) closed or (c) open.
:::

When working with graphs that contain disconnected faces or faces
that are completely surrounded by other faces, also known as
holes, we would need to include additional half-edge references,
such as one along exterior boundary and one along each internal
boundary.  Because Voronoi graphs do not exhibit these
topological features, however, we don't need to worry about this
additional information.

### Satellite Data

Depending upon the application the half-edge, vertex, and face
data structures can also include additional information in the
form of **satellite data**.  For example when using a
doubly-connected edge list to represent a Voronoi graph it will
be helpful to include information about the associated sites,
three for each vertex objects and one for each face object.  We
don't need to store the two sites associated with an edge, as
these can be recovered from the sites of the left faces of the
corresponding half-edges.

### Bounding Box and Pseudo-Vertices

The visualization of Voronoi graphs is easier if we add a few
more objects to the doubly-connected edge list data structure.
In particular we will add the configuration of a bounding box
as well as **pseudo-vertices** along that bounding box that can
be used to terminate any dangling half-edges.  This ensures that
every half-edge includes a non-empty originated vertex references
which in turns makes visualization functions more uniform.

## Implementing Beach Lines

A beach line consists of parabolic arcs and separated
breakpoints, all of which _ordered_.  As we work through
Fortune's algorithm we'll need to be able to not only add and
remove arcs and breakpoints but also efficiently search through
these ordered objects.  All of these operations can be
implemented with a slightly modified **binary search tree**
[@CormenEtAl:2022] where each node represents either an arc or a
breakpoint (@fig-bst).

::: {#fig-bst layout="[ [-25, 50, -25], [100] ]"}
![](figures/bst/beachline/beachline){#fig-bst-beachline}

![](figures/bst/bst/bst){#fig-bst-bst}

Every (a) beach line can be implemented with a (b) slightly
modified binary search tree where internal nodes represent the
breakpoints in the beach line and leaf nodes represent the arcs.
Critically the ordering of nodes in the binary tree preserves the
spatial relationships of the objects in the beach line.  This
main difference between this binary search tree and a standard
binary search is a lack of static keys within each node.
:::

### Arc Nodes

When using a binary search tree to represent a beach line each
arc is represented by a non-empty leaf node.  I will refer to
these as **arc nodes**.

::: {#fig-bst-arc layout="[ [-25, 50, -25], [100] ]"}
![](figures/bst/beachline_arcs/beachline){#fig-bst-beachline-arc}

![](figures/bst/arc_nodes/arc_nodes){#fig-bst-bst-arc}

\(b) Each leaf node in the binary search tree corresponds to
(a) a parabolic arc in the beach line.  Because of the shape of
these arcs changes with the position of the sweep line these arc
nodes do not include any geometric information.  Instead they
include a reference to the site that generates the arc which then
allows us to calculate the current arc shape whenever necessary.
:::

Beyond a reference to their parent nodes, each of these arc nodes
also contain a reference to the site that generates the arc.
This allows us to, for example, dynamically compute the shape of
the arc for any position of the sweep line.  Because the parabola
generated by a single site can contribute to the beach line
multiple times, multiple arc nodes can share the same site
reference.

To avoid having to search through the event queue we also include
in each arc node a reference to any vertex events that centered
on that arc.  This allows us to, for example, quickly find events
that need to be removed from the event queue.

### Breakpoint Nodes

The internal nodes of a binary search tree represent the
breakpoints between the arcs in a beach line.  I will
unsurprisingly refer to these as **breakpoint nodes**.

::: {#fig-bst-bp layout="[ [-25, 50, -25], [100] ]"}
![](figures/bst/beachline_bps/beachline){#fig-bst-beachline-bp}

![](figures/bst/bp_nodes/bp_nodes){#fig-bst-bst-bp}

\(b) Each internal node in the binary search tree corresponds to
(a) a breakpoint on the beach line.  These nodes are equipped
with references to the two neighboring sites as well as a
reference to one Voronoi half-edge.
:::

As with any binary search tree, each internal node contains
references to theirs parents and children in the tree.  When
implementing Fortune's algorithm we will also need to augment the
internal nodes with additional satellite data.

Firstly we'll need to include a reference to the two sites that
generate the neighboring arcs on each side of the breakpoint.
This information allows us to dynamically calculate the position
of each breakpoint for any position of the sweep line.  Unlike
the site references in the arc nodes, these ordered pairs of
sites uniquely label each breakpoint node.

We will also use the breakpoint nodes to store incomplete Voronoi
edges.  Because there are always two breakpoints that fall on the
same Voronoi edge, however, we can't store a unique edge in each
breakpoint node.  What we can do is split each edge into
half-edges and then store a unique half-edge in each breakpoint
node.  This makes the implementation of Fortune's algorithm
more symmetric.

As the sweep line progresses these incomplete half-edges will not
yet have an originating vertex.  For visualization purposes we
can always display these half-edges as if they originated from
the current position of the corresponding breakpoint
(@fig-bst-edges).

![Initially the half-edges stored in each breakpoint node are not
anchored to any Voronoi vertices.  When visualizing the
progression of Fortune's algorithm we can always anchor each
half-edge to the current position of the corresponding
breakpoint.  This ensures that the half-edges always span at
least a segment of full Voronoi edge, shown here in dark teal.](
figures/bst/bp_edges/edges){
width=50% #fig-bst-edges}

### Nearest-Neighbor Search

Binary search trees feature _predecessor_ and _successor_
operations that take any node and efficiently find the next
smallest and next largest node, respectively (@fig-bst-search).
The predecessor and successor of an arc node are always the two
neighboring breakpoint nodes, while the predecessor and success
of a breakpoint node are always the two neighboring arc nodes.

::: {#fig-bst-search layout="[ [100], [100] ]"}
![](figures/bst/predecessor_search/search){#fig-bst-search-pred}

![](figures/bst/successor_search/search){#fig-bst-search-suc}

Binary search trees are equipped with (a) predecessor and (b)
successor search operations which allow us to navigate across
neighboring objects on the beach line.
:::

To implement Fortune's algorithm, however, we will need to
slightly extend these nearest neighbor search operations.  For
example when considering triplets of neighboring arcs for
potential vertex events we need to be able to find not just the
next smallest and next largest nodes but rather the next smallest
and next largest _arc_ nodes.  That said, operations require just
calling the standard predecessor and successor operations twice
(@fig-bst-arc-search).

::: {#fig-bst-arc-search layout="[ [100], [100] ]"}
![](figures/bst/arc_predecessor_search/search){#fig-bst-arc-search-pred}

![](figures/bst/arc_successor_search/search){#fig-bst-arc-search-suc}

Implementing Fortune's algorithm is a bit easier if we introduce
(a) predecessor arc search and (b) successor arc search
operations which allow us to quickly construct triplets of
neighboring arcs that might spawn new Voronoi vertices.
:::

### Beach Line Search

The key difference (absolutely an intentional pun) between
specialized beach line trees and generic binary search trees is
that, while both are ordered, beach line trees don't feature a
traditional key.  Ordering of the nodes in a beach line tree is
determined not by static key values but rather by the relative
vertical position of the arcs and breakpoints along the
corresponding beach line.  As the sweep line progresses the
absolute positions change but the relative positions, and hence
the structure of the binary tree, remains invariant.

Without explicit keys beach line trees do not feature traditional
key search operations.  They do, however, admit conditional
positional searches.

In particular Fortune's algorithm requires being able to find the
arc in a beach line that intersects with a particular vertical
position $y$.  To find the corresponding arc node in a beach line
tree we can modify the standard key search operation, comparing
current vertical positions at each step instead of keys
(@fig-bst-key-search).

::: {#fig-bst-key-search layout="[ [100], [100], [100] ]"}
![](figures/bst/key_search/1/bst){#fig-bst-key-search1}

![](figures/bst/key_search/2/bst){#fig-bst-key-search2}

![](figures/bst/key_search/3/bst){#fig-bst-key-search3}

Unlike the nodes in standard binary search trees, the nodes in
beach line binary trees are not equipped with static keys.  That
said for the internal breakpoint nodes we can dynamically compute
vertical positions given the current configuration of the beach
line, which allows us to search the beach line for positional
intersections.  To find the arc that intersects $y = 3.832$ we
(a) start at the root node and then (b, c) progress down the
binary tree, evaluating current breakpoint positions as needed
until we reach an arc node.
:::

This operation begins by setting the current position of the
sweep line, $S$, and then accessing the root node of the binary
tree.  If the root node is an arc node then the beach line
consists of a single parabola and we're already done.  Otherwise
the root node is a breakpoint node and we can compute the current
vertical position of the breakpoint given $S$.

If the breakpoint height is larger than $y$ then we know that
the intersecting arc is below the breakpoint in the beach line,
and to the left of the current node in the binary tree.  In that
case we move to the left child of the current node.  When the
breakpoint height is smaller than $y$ then we have to move up
the beach line, which is accomplished by moving to the right
child of the current node.

Regardless of which direction we take, we then just repeat these
operations until we reach a childless arc node.  Because arc
nodes correspond to parabolic curves there is no single height
that we can compare to $y$ at this point.  Fortunately we no
longer have to make comparisons as the first arc node we find
will always represent the arc that intersects with $y$.

### Insertion and Deletion

In a binary search tree the node deletion operation depends on
only the relative ordering of the nodes, and not on any explicit
key values.  Consequently removing nodes is the _mostly_ the same
operation for beach line trees and standard binary search trees.
The only subtlety is that after removing a node we will have to
update the site references in any neighboring breakpoints.

The lack of explicit keys in a beach line tree, however, makes
the standard key insertion operation invalid.  That said,
Fortune's algorithm doesn't actually require a generic key
insertion.  Objects are added to the beach line in one and only
one way: splitting an existing arc into three arcs and two
breakpoints.  This can be compartmentalized into a
_subtree insertion_.

::: {#fig-bst-subtree-insert layout="[ 50, 50 ]"}
![](figures/bst/subtree_insertion/1/bst){#fig-bst-subtree-insert1}

![](figures/bst/subtree_insertion/2/bst){#fig-bst-subtree-insert2}

In Fortune's algorithm beach line trees are expanded through a
subtree insertion operation where (a) an arc node is replaced by
a (b) subtree representing a triplet of arc nodes and the two
breakpoints between them.  Note that the site references in the
initial neighboring breakpoint nodes are still correct, and
hence don't have to be updated, after the subtree insertion.
:::

### Balancing

Number of operations needed for all of the binary tree operations
we have discussed here scales with the height of a given tree.
Now the height of most trees is logarithmic in the number of
nodes, which in the context of Fortune's algorithm is also
logarithmic in the number of sites.  Consequently the tree
operations usually require only a logarithmic number of
operations, and a _typical_ evaluation of Fortune's algorithm
will achieve an overall $N \log N$ cost scaling.

Unfortunately the height of _every_ possible binary tree is not
always logarithmic in the number of nodes.  In the worst case the
height of a tree can be linear in the number of nodes, pushing
up the overall cost scaling of Fortune's algorithm to quadratic.

In order to realize the optimal performance of Fortune's
algorithm we need to be able to avoid excessively tall, or
**unbalanced** trees.  Conveniently this can be accomplished with
the use of **self-balancing binary trees**, in particular
**red-black trees** [@CormenEtAl:2022].  These data structures
modify the standard insertion and deletion operations to ensure
that nodes are well-distributed across the entire tree.  With a
little bit of effort we can adapt these balanced operations to
beach line trees, which then allows us to achieve optimal
performance when implementing Fortune's algorithm.

## The Event Queue

The event queue in Fortune's algorithm can be efficiently
implemented with a **priority queue** data structure
[@CormenEtAl:2022].  Priority queues are equipped with an
operation for "popping" the next element off the queue as well as
operations for inserting and deleting elements while maintaining
the overall order of the event queue.

To implement Fortune's algorithm we'll need to consider two types
of events.  Vertex events contain the position of the sweep line
where an arc will collapse and a reference to that arc.  Site
events, on the other hand, contain the position where sweep line
will encounter a new site as well as a reference to that site.

## Event Processing

Fortune's algorithm proceeds by popping the next event off of
the event queue, using that event to advanced the sweep line and
beach line, and then updating the Voronoi graph as necessary.
With doubly-connected event queues, binary search trees, and
their operations processing each event is relatively
straightforward.

## Processing Vertex Events

The first step when processing a new vertex event is to remove
any other vertex events that will become invalid after we remove
the collapsing arc.  This can be done efficiently by using
the arc predecessor and arc successor operations of the beach
line tree to find the neighboring arcs, and then remove any
vertex events referenced in those arcs.

Next we manage the expansion of the Voronoi graph.  We begin by
adding a new vertex to the doubly-connected edge list at the
point where the arc collapses and the neighboring breakpoints
intersect.

Then we have to deal with the Voronoi edges.  We create a new
half-edge originating at the new vertex and its twin that, for
the moment, is left unanchored (@fig-vertex-event-edges).  Lastly
we link up the previous and next references between these two new
half-edges, the two half-edges in the neighboring breakpoints,
and their twins (@fig-vertex-event-edge-connections).

::: {#fig-vertex-event-edges layout="[ [50, 50], [-25, 50, -25] ]"}
![](figures/fortune/edges/vertex/1/edges){#fig-vertex-event-edges1}

![](figures/fortune/edges/vertex/2/edges){#fig-vertex-event-edges2}

![](figures/fortune/edges/vertex/3/edges){#fig-vertex-event-edges3}

Vertex events require substantial updates to the half-edges of
the Voronoi graph.  (a) The breakpoints that merge in a vertex
event contain references to two half-edges, here `he1` and `he3`.
(b) When processing a vertex event we delete these references.
(c) Then we create two new half-edges, one of which is referenced
by the new vertex, here `he5`, and one of which is referenced by
the new breakpoint, here `he6`.
:::

![The trickiest part of processing a vertex event is making sure
that the four existing half-edges and the two new half-edges all
properly reference each other.  These references ultimately form
the skeleton of the Voronoi graph.](
figures/fortune/edges/connections/edges){
width=50% #fig-vertex-event-edge-connections}

Now we're ready to remove the collapsing arc.  This requires
removing the corresponding arc node as well as one of its
neighboring breakpoint nodes, rebalancing the binary search tree
after each removal (@fig-vertex-event-deletion).  Which
breakpoint node we remove is a matter of convention, but to avoid
any ambiguity I will maintain a convention where we always remove
the smaller breakpoint.

::: {#fig-vertex-event-deletion layout="[ [100], [100], [100] ]"}
![](figures/bst/deletion/1/deletion){#fig-vertex-event-deletion1}

![](figures/bst/deletion/2/deletion){#fig-vertex-event-deletion2}

![](figures/bst/deletion/3/deletion){#fig-vertex-event-deletion3}

\(a, b) Removing an arc from a beach line requires deleting two
nodes from the corresponding binary tree, a leaf node for the
arc itself and an internal node for one of the neighboring
breakpoints.  Which breakpoint we remove is a matter of
convention.  (c) After removing the two nodes from the binary
tree we need to update the site information in the remaining
breakpoint node.  Here the left site has to be updated from
$n_{6}$ to $n_{2}$.
:::

At this point the remaining breakpoint node, the larger
breakpoint in my chosen convention, needs to be updated.
Firstly its left site needs to be updated to the site that
generates its new left neighboring arc; this new neighbor can be
accessed by calling the predecessor operation on the updated
tree.  Secondly we need to set its half-edge references to
the new twin half-edge that we just created.

Lastly there are two new triplets of neighboring arcs that we
have to check for possible new vertex events.  We can efficiently
access these arcs by repeatedly using the arc predecessor and arc
successor operations

### Processing Site Events

Because site events do not update the Voronoi diagram they are a
little bit easier to process.

First things first we have to find the arc that intersects with
the height of the new site by searching through the beach line
tree.  Because splitting this arc will invalidate any existing
vertex event centered on it, we have to remove any corresponding
vertex event from the event queue.

In order to incorporate the new site into the beach line we have
to apply the subtree insertion operation on the intersecting
node, replace it with a subtree that includes two arc nodes
associated with the original site, one arc node associated with
the new site, and two interior breakpoint nodes separating them.
When working with a balanced binary tree we also have to make
sure that this insertion operation properly rebalances the
updated tree.

The two new breakpoint nodes are then filled with two new
half-edges that are twinned to each other
(@fig-site-event-edges).  One half-edge faces the old site and
the other faces the face the new site.

![The introduction of a new site splits an existing arc on the
beach line into three arcs separated by two new breakpoints.
This also creates two new half-edges in the Voronoi graph that
are twinned with each other.  A reference to each of these
half-edges is stored in one of the two new breakpoint nodes that
are inserted into the beach line tree.  Here a reference to the
half-edge `he1` is stored in the breakpoint node `(n1, n2)` while
a reference to `he2` is stored in the breakpoint node `(n2, n1)`.](
figures/fortune/edges/site/edges){
width=50% #fig-site-event-edges}

Finally we have to consider potential new vertex events that.
Incorporating the new site creates up to two new triplets of
neighboring arcs that might generate new vertex events.
Fortunately the arc nodes in these new triplets are once again
readily accessed with the arc predecessor and arc successor
beach line operations.

# Clean Up

Once the event queue is exhausted we are left with a
doubly-connected edge list that contains all of the information
about the Voronoi graph, and hence the Voronoi diagram.

At this point the binary search tree implementing the beach line
will _not_ be empty.  The remaining breakpoint nodes all contain
half-edges with unbounded origins.  If we want to visualize
the Voronoi graph neatly then we'll need to anchor these dangling
edges by creating a bounding box and then adding a pseudo-vector
to the doubly-connected edge list for the intersection of each
dangling half-edge with the bounding box.

# Demonstration

To put all of this discussion into practice let's work through a
full implementation of Fortune's algorithm in `R`.  To be honest
`R` is at best an awkward language for this application, which is
much more natural in languages that are more friendly to
object-oriented programming.

Here I get by with `R6` classes [@Chang:2025], but we'll have to
suffer with some of its clumsy interactions with base `R`
graphics.  At least using `R6` classes gets us pass-by-reference
semantics for class instances instead of `R`'s usual
call-by-need/lazy evaluation semantics.

```{r}
library(R6)
```

Speaking of graphics, we'll start with some graphics settings.

```{r}
par(family="serif", las=1, bty="l",
    cex.axis=1, cex.lab=1, cex.main=1,
    xaxs="i", yaxs="i", mar = c(5, 5, 3, 1))

c_light <- c("#DCBCBC")
c_light_highlight <- c("#C79999")
c_mid <- c("#B97C7C")
c_mid_highlight <- c("#A25050")
c_dark <- c("#8F2727")
c_dark_highlight <- c("#7C0000")

c_light_teal <- c("#6B8E8E")
c_mid_teal <- c("#487575")
c_dark_teal <- c("#1D4F4F")
```

## Beach Line

To define a beach line we'll first need nodes.

This class implements both breakpoint and arc nodes by including
the satellite data for each and using only one type of data at
a time.  In particular the variables `site` and
`scheduled_vertex` are used for only arc nodes while `site_left`,
`site_right`, and `he` are used for only breakpoint nodes.
Note that the site variables are integer indices which will be
used to access site information from global arrays.

The variable `red` denotes the coloring of the node which is used
for balancing binary trees.

```{r}
node <- R6Class("node",
                public = list(
                  site = NULL,
                  scheduled_vertex = NULL,
                  site_left = NULL,
                  site_right = NULL,
                  he = NULL,
                  parent = NULL,
                  left = NULL,
                  right = NULL,
                  red = FALSE,
                  id = NULL,
                  initialize = function(beachline) {
                    if (is.integer(beachline)) {
                      self$id <- beachline
                    } else {
                      self$parent <- beachline$nil
                      self$left <- beachline$nil
                      self$right <- beachline$nil

                      beachline$last_node_id <- beachline$last_node_id + 1
                      self$id <- beachline$last_node_id
                    }
                  }
                )
)
```

The `id` variable allows us to determine if two `node` instances
refer to the same object.  In many programming languages this is
done by directly comparing memory addresses, but `R6` classes
don't seem to play will with memory.


```{r}
`==.node` = function(x, y) { x$id == y$id }
`!=.node` = function(x, y) { x$id != y$id }
```

An actual beach line tree is defined by a root node.  Here we'll
be using a self-balancing red-black binary tree which uses the
coloring of the nodes to limit the height of any given tree.
Red-black binary trees requires a dedicated null, or nil, node.

```{r}
beachline <- R6Class("beachline",
                     public = list(
                       nil = NULL,
                       root = NULL,
                       last_node_id = NULL,
                       last_half_edge_id = NULL,
                       initialize = function() {
                         self$nil <- node$new(as.integer(0))
                         self$root <- self$nil
                         self$last_node_id <- as.integer(1)
                         self$last_half_edge_id <- as.integer(1)
                       }
                     ))
```

The `beachline` class also gives us a place to keep track of the
node indices, as well as half-edge indices that we'll use later.

### Black-Red Binary Tree Operations

The `insert_fixup` function rebalances a binary tree after we
add the node `z` to the tree.

```{r}
beachline$set("private", "left_rotate",
function(x) {
  if (x$right == self$nil)
    return()

  y <- x$right
  x$right <- y$left

  if (y$left != self$nil)
    y$left$parent <- x

  y$parent <- x$parent

  if (x$parent == self$nil) {
    self$root <- y
  } else {
    if (x$parent$left == x)
      x$parent$left <- y
    else
      x$parent$right <- y
  }

  y$left <- x
  x$parent <- y
})
```

```{r}
beachline$set("private", "right_rotate",
function(x) {
  if (x$left == self$nil)
    return()

  y <- x$left
  x$left <- y$right

  if (y$right != self$nil)
    y$right$parent <- x

  y$parent <- x$parent

  if (x$parent == self$nil) {
    self$root <- y
  } else {
    if (x$parent$right == x)
      x$parent$right <- y
    else
      x$parent$left <- y
  }

  y$right <- x
  x$parent <- y
})
```

```{r}
beachline$set("private", "insert_fixup",
function(z) {
  while (z$parent$red) {
    if (z$parent == z$parent$parent$left) {
      y <- z$parent$parent$right
      if (y$red) {
        z$parent$red <- FALSE
        y$red <- FALSE
        z$parent$parent$red <- TRUE
        z <- z$parent$parent
      } else {
        if (z == z$parent$right) {
          z <- z$parent
          private$left_rotate(z)
        }
        z$parent$red <- FALSE
        z$parent$parent$red <- TRUE
        private$right_rotate(z$parent$parent)
      }
    } else {
      y <- z$parent$parent$left

      if (y$red) {
        z$parent$red <- FALSE
        y$red <- FALSE
        z$parent$parent$red <- TRUE
        z <- z$parent$parent
      } else {
        if (z == z$parent$left) {
          z <- z$parent
          private$right_rotate(z)
        }
        z$parent$red <- FALSE
        z$parent$parent$red <- TRUE
        private$left_rotate(z$parent$parent)
      }
    }
  }
  self$root$red <- FALSE
})
```

Deleting a node from a red-black tree takes some care.  To
properly remove a node we'll need to be able to calculate the
boundaries of subtrees.

```{r}
beachline$set("public", "min",
function(x) {
  if (x == self$nil)
    return(self$nil)
  while (x$left != self$nil)
    x <- x$left
  x
})
```

```{r}
beachline$set("public", "max",
function(x) {
  if (x == self$nil)
    return(self$nil)
  while (x$right != self$nil)
    x <- x$right
  x
})
```

Then we'll need to carefully rebalance the tree.

```{r}
beachline$set("private", "transplant",
function(u, v) {
  if (u$parent == self$nil) {
    self$root <- v
  } else {
    if (u == u$parent$left) {
      u$parent$left <- v
    } else {
      u$parent$right <- v
    }
  }
  v$parent <- u$parent
})
```

```{r}
beachline$set("private", "delete_fixup",
function(x) {
  while (x != self$root && !x$red) {
    if (x == x$parent$left) {
      w <- x$parent$right

      if (w$red) {
        w$red <- FALSE
        x$parent$red <- TRUE
        private$left_rotate(x$parent)
        w <- x$parent$right
      }

      if (!w$left$red && !w$right$red) {
        w$red <- TRUE
        x <- x$parent
      } else {
        if (!w$right$red) {
          w$left$red <- FALSE
          w$red <- TRUE
          private$right_rotate(w)
          w <- x$parent$right
        }

        w$red <- x$parent$red
        x$parent$red <- FALSE
        w$right$red <- FALSE
        private$left_rotate(x$parent)
        x <- self$root
      }
    } else {
      w <- x$parent$left

      if (w$red) {
        w$red <- FALSE
        x$parent$red <- TRUE
        private$right_rotate(x$parent)
        w <- x$parent$left
      }

      if (!w$right$red && !w$left$red) {
        w$red <- TRUE
        x <- x$parent
      } else {
        if (!w$left$red) {
          w$right$red <- FALSE
          w$red <- TRUE
          private$left_rotate(w)
          w <- x$parent$left
        }

        w$red <- x$parent$red
        x$parent$red <- FALSE
        w$right$red <- FALSE
        private$right_rotate(x$parent)
        x <- self$root
      }
    }
  }
  x$red <- FALSE
})
```

```{r}
beachline$set("public", "delete",
function(z) {
  y <- z
  y_old_red <- y$red

  if (z$left == self$nil) {
    x <- z$right
    private$transplant(z, z$right)
  } else if (z$right == self$nil) {
    x <- z$left
    private$transplant(z, z$left)
  } else {
    y <- self$min(z$right)
    y_old_red <- y$red
    x <- y$right

    if (y != z$right) {
      private$transplant(y, y$right)
      y$right <- z$right
      y$right$parent <- y
    } else {
      x$parent <- y
    }

    private$transplant(z, y)
    y$left <- z$left
    y$left$parent <- y
    y$red <- z$red
  }

  if (!y_old_red)
    private$delete_fixup(x)
})
```

Fortune's algorithm requires accessing lots of neighboring nodes.

```{r}
beachline$set("public", "arc_successor",
function(x) {
  if (x$right != self$nil) {
    # Stay still
    y <- x
  } else {
    # Ascend left
    y <- x$parent
    while (   y != self$nil
           && x == y$right) {
      x <- y
      y <- y$parent
    }
    # Ascend right
    while (   y != self$nil
           && y$right == self$nil) {
      y <- y$parent
    }
  }
  if (y != self$nil)
    y <- self$min(y$right)
  return(y)
})
```

```{r}
beachline$set("public", "successor",
function(x) {
  if (x$right != self$nil) {
    return(self$min(x$right))
  } else {
    y <- x$parent
    while (   y != self$nil
           && x == y$right) {
      x <- y
      y <- y $parent
    }
    return(y)
  }
})
```

```{r}
beachline$set("public", "arc_predecessor",
function(x) {
  if (x$left != self$nil) {
    # Stay still
    y <- x
  } else {
    # Ascend right
    y <- x$parent
    while (   y != self$nil
           && x == y$left) {
      x <- y
      y <- y$parent
    }
    # Ascend left
    while (   y != self$nil
           && y$left == self$nil) {
      y <- y$parent
    }
  }
  if (y != self$nil)
    y <- self$max(y$left)
  return(y)
})
```

```{r}
beachline$set("public", "predecessor",
function(x) {
  if (x$left != self$nil) {
    return(self$max(x$left))
  } else {
    y <- x$parent
    while (   y != self$nil
              && x == y$left) {
      x <- y
      y <- y$parent
    }
    return(y)
  }
})
```

Tree walks are useful for displaying the current ordering of the
nodes.

```{r}
beachline$set("public", "inorder_arc_walk",
function(x = NULL) {
  if (is.null(x))
    x <- self$root

  if (x != self$nil) {
    self$inorder_arc_walk(x$left)

    if (!is.null(x$site))
      cat(paste0(x$site, ', '))

    self$inorder_arc_walk(x$right)
  }
})
```

```{r}
beachline$set("public", "inorder_walk",
function(x = NULL) {
  if (is.null(x))
    x <- self$root

  if (x != self$nil) {
    self$inorder_walk(x$left)

    if (is.null(x$site)) {
      cat(paste0('(', x$site_left, ',',
                 x$site_right, '), '))
    } else {
      cat(paste0('(', x$site, '), '))
    }

    self$inorder_walk(x$right)
  }
})
```

### Visualization Functions

Visualizing the nodes in a binary search tree is particularly
elegant with recursive functions.  That said any function that
interacts with base `R` graphics objects has to be external to
the `R6` class.

```{r}
beachline$set("public", "height",
function(node=NULL) {
  if (is.null(node))
    node <- self$root

  if (node == self$nil)
    return(0)

  left_height <- self$height(node$left)
  right_height <- self$height(node$right)

  1 + max(left_height, right_height)
})
```

```{r}
plot_tree <- function(tree, delta=0.5, text_cex=0.75) {
  H <- tree$height()

  par(mar=c(0, 0, 0, 0))

  plot(NULL,
       xlab='', ylab='', xaxt="n", yaxt="n", frame.plot=F,
       xlim=c(-delta * (2**(H - 1) + 1), delta * (2**(H - 1) + 1)),
       ylim=c(-(H - 0.5) * delta, 0.5 * delta))

  plot_node(tree$nil, tree$root, 0, 0,
            delta * 2**(H - 2), delta, text_cex)
}
```

```{r}
plot_node <- function(nil, node,
                      x, y,
                      delta_x, delta_y,
                      text_cex) {

  if (node$left != nil) {
    lines(c(x, x - delta_x), c(y, y - delta_y),
          col=c_mid_teal, lwd=2)
    plot_node(nil, node$left,
              x - delta_x, y - delta_y,
              delta_x / 2, delta_y, text_cex)
  }

  if (node$right != nil) {
    lines(c(x, x + delta_x), c(y, y - delta_y),
          col=c_mid_teal, lwd=2)
    plot_node(nil, node$right,
              x + delta_x, y - delta_y,
              delta_x / 2, delta_y, text_cex)
  }

  if (node$red)
    points(x, y, pch=16, col=c_dark, cex=3)
  else
    points(x, y, pch=16, col="black", cex=3)

  if (!is.null(node$site)) {
    text(x, y, col="white", cex=text_cex,
         labels=node$site)
  } else {
    text(x, y, col="white", cex=text_cex,
         labels=paste0('(', node$site_left, ",",
                            node$site_right, ')'))
  }
}
```

### Fortune's Algorithm Geometry Calculations

As we've seen, implementing Fortune's algorithm requires a lot of
geometry calculations.

For example we'll need to be able to compute the vertical
position of the breakpoint separating two neighboring arcs on a
beach line.  Site positions are accessed by indexing the global
arrays `xs` and `ys`.

```{r}
beachline$set("public", "pair_intersection",
function(n1, n2, S) {
  if (xs[n1] == S)
    return(ys[n1])
  if (xs[n2] == S)
    return(ys[n2])

  a <- xs[n2] - xs[n1]
  beta <- ys[n2] * (S - xs[n1]) - ys[n1] * (S - xs[n2])
  c <- (   ys[n2]**2 * (S - xs[n1])
           - ys[n1]**2 * (S - xs[n2])
           - (S - xs[n1]) * (S - xs[n2]) * (xs[n2] - xs[n1]) )

  d <- sqrt(beta**2 - a * c)
  y_int <- (beta + d) / a
  if (y_int < ys[n1] | y_int > ys[n2])
    y_int <- (beta - d) / a

  return(y_int)
})
```

The horizontal position of a breakpoint can be recovered from the
parabola defining each arc.  Here `(fx, fy)` defines the focal
point of the parabola and `S` the vertical directrix.

```{r}
beachline$set("public", "parabola_xs",
function(parabola_ys, fx, fy, S) {
  0.5 * ( (fx + S) - (parabola_ys - fy)**2 / (S - fx) )
})
```

We can't forget the circumcircles.

```{r}
beachline$set("public", "circumcircle",
function(n1, n2, n3) {
  D <- 2 * (     xs[n1] * (ys[n2] - ys[n3])
               + xs[n2] * (ys[n3] - ys[n1])
               + xs[n3] * (ys[n1] - ys[n2]) )
  xc <- (    (xs[n1]**2 + ys[n1]**2) * (ys[n2] - ys[n3])
           + (xs[n2]**2 + ys[n2]**2) * (ys[n3] - ys[n1])
           + (xs[n3]**2 + ys[n3]**2) * (ys[n1] - ys[n2]) ) / D

  yc <- (    (xs[n1]**2 + ys[n1]**2) * (xs[n3] - xs[n2])
           + (xs[n2]**2 + ys[n2]**2) * (xs[n1] - xs[n3])
           + (xs[n3]**2 + ys[n3]**2) * (xs[n2] - xs[n1]) ) / D

  r <- sqrt( (xc - xs[n1])**2 + (yc - ys[n1])**2 )
  c(xc, yc, r)
})
```

We can use these geometric functions to, for example, find the
node representing the arc that intersects with a given vertical
position.

```{r}
beachline$set("private", "arc_search",
function(z, y, S) {
  if (is.null(z$site)) {
    by <- self$pair_intersection(z$site_left,
                                 z$site_right, S)
    if (y <= by)
      z <- private$arc_search(z$left, y, S)
    else
      z <- private$arc_search(z$right, y, S)
  }
  return(z)
})
```

### Fortune's Algorithm Event Functions

With all of these auxiliary functions we are finally ready to
implement event processing functions.

One operating that appears over and over again when processing
events is checking triplets of arc nodes to see if they define a
valid vertex event and, if so, adding that event to the event
queue `queue`.  There is probably a more elegant way to verify
that `alpha_mid` will collapse than comparing `upper_int` to
`lower_int`.

```{r}
beachline$set("private", "schedule_new_vertex_event",
function(queue, alpha_left, alpha_mid, alpha_right, S) {
  # Skip vertex if any of the sites are nil
  if (alpha_left == self$nil || alpha_right == self$nil)
    return()

  # Skip vertex if any of the sites are the same
  if (alpha_left$site == alpha_right$site) {
    return()
  }

  # Construct circumcircle
  int <- self$circumcircle(alpha_left$site,
                           alpha_mid$site,
                           alpha_right$site)
  S_new <- int[1] + int[3]


  lower_int <- self$pair_intersection(alpha_left$site,
                                      alpha_mid$site, S_new)
  upper_int <- self$pair_intersection(alpha_mid$site,
                                      alpha_right$site, S_new)

  if (abs(upper_int - lower_int) < 1e-12) {
    vertex_event <- event$new(NULL, S_new, NULL, alpha_mid)
    alpha_mid$scheduled_vertex <- vertex_event
    queue$insert_event(vertex_event)
  }
})
```

Now we can process site events.

```{r}
beachline$set("private", "process_site_event",
function(queue, new_event) {
  new_site <- new_event$site

  if (self$root == self$nil) {
    # Initialize beach line if empty
    self$root <- node$new(self)
    self$root$site <- new_site
  } else {
    # Find beach line arc that intersects
    # vertical position of new site
    gamma <- private$arc_search(self$root,
                                ys[new_site],
                                new_event$x)
    old_site <- gamma$site

    # Delete any scheduled vertex
    # events at intersecting arc
    if (!is.null(gamma$scheduled_vertex)) {
      queue$delete_event(gamma$scheduled_vertex)
      gamma$scheduled_vertex <- NULL
    }

    # Expand intersecting arc into new subtree
    gamma$site_left <- old_site
    gamma$site_right <- new_site
    gamma$scheduled_vertex <- NULL
    gamma$site <- NULL

    gamma$left <- node$new(self)
    gamma$left$parent <- gamma
    gamma$left$site <- old_site
    gamma$left$red <- TRUE
    private$insert_fixup(gamma$left)

    gamma$right <- node$new(self)
    gamma$right$parent <- gamma
    gamma$right$site_left <- new_site
    gamma$right$site_right <- old_site
    gamma$right$red <- TRUE

    zeta <- gamma$right
    private$insert_fixup(zeta)

    zeta$left <- node$new(self)
    zeta$left$parent <- zeta
    zeta$left$site <- new_site
    zeta$left$red <- TRUE
    private$insert_fixup(zeta$left)

    zeta$right <- node$new(self)
    zeta$right$parent <- zeta
    zeta$right$site <- old_site
    zeta$right$red <- TRUE
    private$insert_fixup(zeta$right)

    # Instantiate and connect new half-edges
    gamma$he <- half_edge$new(self, NULL, old_site,
                              NULL, NULL, NULL)
    gamma$he$twin <- half_edge$new(self, NULL, new_site,
                                   gamma$he, NULL, NULL)
    zeta$he <- gamma$he$twin

    # Attempt to schedule vertex events
    # for the two new arc triplets
    alpha1 <- self$arc_predecessor(zeta$left)
    alpha2 <- self$arc_predecessor(alpha1)
    private$schedule_new_vertex_event(queue,
                                      alpha2, alpha1, zeta$left,
                                      new_event$x)

    beta1 <- self$arc_successor(zeta$left)
    beta2 <- self$arc_successor(beta1)
    private$schedule_new_vertex_event(queue,
                                      zeta$left, beta1, beta2,
                                      new_event$x)
  }
})
```

Vertex event processing follows.

```{r}
beachline$set("private", "process_vertex_event",
function(queue, graph, new_event) {
  gamma <- new_event$arc

  beta1 <- self$arc_successor(gamma)
  beta2 <- self$arc_successor(beta1)

  alpha1 <- self$arc_predecessor(gamma)
  alpha2 <- self$arc_predecessor(alpha1)

  # Clean up deprecated vertex events in the queue
  if (!is.null(beta1$scheduled_vertex)) {
    queue$delete_event(beta1$scheduled_vertex)
    beta1$scheduled_vertex <- NULL
  }

  if (!is.null(alpha1$scheduled_vertex)) {
    queue$delete_event(alpha1$scheduled_vertex)
    alpha1$scheduled_vertex <- NULL
  }

  # Create vertex and update edges
  int <- self$circumcircle(alpha1$site, gamma$site, beta1$site)
  v <- vertex$new(int[1:2], alpha1$site, gamma$site, beta1$site)
  graph$add_vertex(v)

  # Connect half-edges
  bp1 <- self$predecessor(gamma)
  bp2 <- self$successor(gamma)

  new_he <- half_edge$new(self, v, beta1$site,
                          NULL, NULL, NULL)
  new_he$twin <- half_edge$new(self, NULL, alpha1$site,
                               new_he, NULL, NULL)

  v$he <- new_he

  bp1$he$origin <- v
  bp2$he$origin <- v

  bp2$he$prv <- bp1$he$twin
  bp1$he$twin$nxt <- bp2$he

  bp2$he$twin$nxt <- new_he
  new_he$prv <- bp2$he$twin

  new_he$twin$nxt <- bp1$he
  bp1$he$prv <- new_he$twin

  # Delete deprecated nodes
  self$delete(gamma)
  self$delete(bp1)

  # Update remaining breakpoint
  bp2$site_left <- alpha1$site
  bp2$he <- new_he$twin

  # Check for new vertex events
  private$schedule_new_vertex_event(queue,
                                    alpha2, alpha1, beta1,
                                    new_event$x)

  private$schedule_new_vertex_event(queue,
                                    alpha1, beta1, beta2,
                                    new_event$x)
})
```

```{r}
beachline$set("public", "process_next_event",
function(queue, graph) {
  next_event <- queue$pop_next_element()

  if (!is.null(next_event$site)) {
    private$process_site_event(queue,
                               next_event)
  } else {
    private$process_vertex_event(queue,
                                 graph,
                                 next_event)
  }
})
```

### Fortune's Algorithm Visualization Functions

To monitor the progress of Fortune's algorithm we'll lastly
define some functions that allow us to plot the shape of the
beach line for any sweep line position.

```{r}
beachline$set("public", "fill_breakpoints",
function(S, z=NULL) {
  if (is.null(z))
    z <- self$root

  if (is.null(z$site)) {
    by <- self$pair_intersection(z$site_left,
                                 z$site_right, S)
    bx <- self$parabola_xs(by, xs[z$site_left],
                           ys[z$site_left], S)
    z$he$pseudo_origin <- c(bx, by)
    self$fill_breakpoints(S, z$left )
    self$fill_breakpoints(S, z$right)
  }
})
```

```{r}
plot_beachline <- function(bl, S,
                           xlim=c(-4, -2),
                           ylim=c(-2, 2)) {
  bl$fill_breakpoints(S)

  plot(xs, ys,
       axes=FALSE, ann=FALSE, frame.plot=F,
       col=c_dark, pch=16,
       xlim=xlim, ylim=ylim,
       main="")

  # Plot sweep line
  abline(v=S, col=c_mid_teal, lty=2, lwd=2)

  # Plot site parabolas given the sweep line
  for (i in which(xs < S)) {
    text(xs[i] + 0.1, ys[i], i)
    parabola_ys <- seq(ylim[1], ylim[2], 0.01)
    lines(bl$parabola_xs(parabola_ys, xs[i], ys[i], S),
          parabola_ys, lwd=2, col="#DDDDDD")
  }

  # Plot beach line
  if (bl$root != bl$nil)
    plot_arc(bl, bl$root, ylim[1], ylim[2], S)
}
```

```{r}
plot_arc <- function(bl, z, y_min, y_max, S) {
  if (!is.null(z$site)) {
    # Leaf node
    if (y_max > y_min + 1e-3) {
      parabola_ys <- seq(y_min, y_max, 1e-3)
      lines(bl$parabola_xs(parabola_ys, xs[z$site],
                           ys[z$site], S),
            parabola_ys,
            lty=2, lwd=2, col=c_mid)
    }
  } else {
    # Internal node
    by <- bl$pair_intersection(z$site_left,
                               z$site_right, S)
    bx <- bl$parabola_xs(by, xs[z$site_left],
                         ys[z$site_left], S)
    points(bx, by, col=c_light, pch=16)
    text(bx + 0.1, by,
         labels=paste0(z$site_left, ",", z$site_right))

    if (is.null(z$he$origin)) {
      p1 <- z$he$pseudo_origin
    } else {
      p1 <- z$he$origin$pos
    }
    if (is.null(z$he$twin$origin)) {
      p2 <- z$he$twin$pseudo_origin
    } else {
      p2 <- z$he$twin$origin$pos
    }

    lines(c(p1[1], p2[1]), c(p1[2], p2[2]),
          lwd=2, col=c_mid_teal)

    plot_arc(bl, z$left,  y_min, by, S)
    plot_arc(bl, z$right, by, y_max, S)
  }
}
```

## Event Queue

Similarly to how we handled the `node` class, we'll overload the
`event` class to handle both site and vertex events at the same
time.  The variable `site` is used for only site events while the
variable `arc` is used for only vertex events.

```{r}
event <- R6Class("event",
                 public=list(
                   x=NULL,
                   site=NULL,
                   arc=NULL,
                   queue_idx=NULL,
                   initialize = function(i, x, site, arc) {
                     self$queue_idx <- i
                     self$x <- x
                     self$site <- site
                     self$arc <- arc
                   }))
```

The event queue itself is a pretty straightforward implementation
of a priority queue data structure.  When creating a new event
queue instance we pass in a vector of indices for the available
sites which are then used to create the site events that
initially populate the queue.

```{r}
event_queue <- R6Class("event_queue",
                       public=list(
                         idxs=NULL,
                         events=NULL,
                         length=0,
                         initialize = function(sites, xs) {
                           self$length <- length(sites)
                           self$idxs <- 1:self$length
                           for (i in seq_along(sites)) {
                             new_event <- event$new(i, xs[i], i, NULL)
                             self$events <- append(self$events,
                                                   new_event)
                           }
                           for (i in (self$length %/% 2):1) {
                             private$min_heapify(i)
                           }
                         },
                         is_empty = function() {
                           self$length == 0
                         },
                         is_not_empty = function() {
                           self$length > 0
                         }))
```

```{r}
event_queue$set("private", "min_heapify",
function(i) {
  l <- 2 * i
  r <- l + 1

  if (   l <= self$length
      &&   self$events[[self$idxs[l]]]$x
         < self$events[[self$idxs[i]]]$x) {
    smallest <- l
  } else {
    smallest <- i
  }

  if (   r <= self$length
         &&   self$events[[self$idxs[r]]]$x
         < self$events[[self$idxs[smallest]]]$x) {
    smallest <- r
  }

  if (smallest != i) {
    n <- self$events[[self$idxs[i]]]$queue_idx
    self$events[[self$idxs[i]]]$queue_idx <-
      self$events[[self$idxs[smallest]]]$queue_idx
    self$events[[self$idxs[smallest]]]$queue_idx <- n

    n <- self$idxs[i]
    self$idxs[i] <- self$idxs[smallest]
    self$idxs[smallest] <- n

    private$min_heapify(smallest)
  }
})
```

```{r}
event_queue$set("public", "insert_event",
function(new_event) {
  self$length <- self$length + 1
  self$idxs <- c(self$idxs, length(self$events) + 1)

  new_event$queue_idx <- self$length
  self$events <- append(self$events, new_event)

  i <- self$length
  while (   i > 1
            &&   self$events[[self$idxs[i %/% 2]]]$x
            > self$events[[self$idxs[i]]]$x) {
    n <- self$events[[self$idxs[i]]]$queue_idx
    self$events[[self$idxs[i]]]$queue_idx <-
      self$events[[self$idxs[i %/% 2]]]$queue_idx
    self$events[[self$idxs[i %/% 2]]]$queue_idx <- n

    n <- self$idxs[i]
    self$idxs[i] <- self$idxs[i %/% 2]
    self$idxs[i %/% 2] <- n

    i <- i %/% 2
  }
})
```

```{r}
event_queue$set("public", "insert_site_event",
function(site, xs) {
  self$insert_event(event$new(NULL, xs[site], site, NULL))
})
```

```{r}
event_queue$set("public", "insert_vertex_event",
function(S, arc) {
  self$insert_event(event$new(NULL, S, NULL, arc))
})
```

```{r}
event_queue$set("public", "pop_next_element",
function() {
  if (self$length == 0) {
    print("The queue is empty!")
    return(NULL)
  }

  next_element <- self$events[[self$idxs[1]]]

  self$idxs[1] <- self$idxs[self$length]
  self$events[[self$idxs[1]]]$queue_idx <- 1

  self$idxs <- self$idxs[-self$length]
  self$length <- self$length - 1

  private$min_heapify(1)

  next_element
})
```

```{r}
event_queue$set("public", "delete_event",
function(event) {
  event$x <- self$events[[self$idxs[1]]]$x - 1

  i <- event$queue_idx
  while (   i > 1
         &&   self$events[[self$idxs[i %/% 2]]]$x
            > self$events[[self$idxs[i]]]$x) {
    n <- self$events[[self$idxs[i]]]$queue_idx
    self$events[[self$idxs[i]]]$queue_idx <-
      self$events[[self$idxs[i %/% 2]]]$queue_idx
    self$events[[self$idxs[i %/% 2]]]$queue_idx <- n

    n <- self$idxs[i]
    self$idxs[i] <- self$idxs[i %/% 2]
    self$idxs[i %/% 2] <- n

    i <- i %/% 2
  }
  e <- self$pop_next_element()
})
```

```{r}
event_queue$set("public", "check_min_heapify",
function() {
  if (self$length > 0) {
    for (i in 1:self$length)  {
      if (i %/% 2 == 0) next
      print(   self$events[[self$idxs[i %/% 2]]]$x
            <= self$events[[self$idxs[i]]]$x)
    }
  }
})
```

```{r}
event_queue$set("public", "next_event",
function() {
  if (self$length > 0)
    self$events[[self$idxs[1]]]
})
```

## Doubly-Connected Edge List

In order to represent a Voronoi graph we'll need classes for
half_edges, vertices, and faces and then the doubly-connected
edge list that encapsulates them.  Like the beach line nodes the
half-edges uses a global indexing to allow for equality
comparisons.

```{r}
half_edge <- R6Class("half_edge",
                     public=list(
                       pseudo_origin=NULL,
                       origin=NULL,
                       left_face=NULL,
                       twin=NULL,
                       prv=NULL,
                       nxt=NULL,
                       visited=FALSE,
                       id=NULL,
                       initialize = function(tree, o, lf, t, p, n) {
                         self$origin <- o
                         self$left_face <- lf
                         self$twin   <- t
                         self$prv    <- p
                         self$nxt    <- n

                         tree$last_half_edge_id <- tree$last_half_edge_id + 1
                         self$id <- tree$last_half_edge_id
                       }))
```

```{r}
`==.half_edge` = function(x, y) { x$id == y$id }
`!=.half_edge` = function(x, y) { x$id != y$id }
```

```{r}
vertex <- R6Class("vertex",
                  public=list(
                    pos=NULL,
                    n1=NULL,
                    n2=NULL,
                    n3=NULL,
                    he=NULL,
                    initialize = function(p, n1, n2, n3) {
                      self$pos <- p
                      self$n1 <- n1
                      self$n2 <- n2
                      self$n3 <- n3
                    }))
```

```{r}
face <- R6Class("face",
                public=list(
                  site=NULL,
                  he=NULL,
                  initialize = function(site, he) {
                    self$site <- site
                    self$he <- he
                  }))
```

```{r}
dcel <- R6Class("dcel",
                public=list(
                  vertices=NULL,
                  pseudo_vertices=NULL,
                  faces=NULL,
                  xlim=NULL,
                  ylim=NULL,
                  initialize = function() {
                    self$vertices <- list()
                    self$pseudo_vertices <- list()
                    self$faces <- list()
                  },
                  add_vertex = function(v) {
                    self$vertices <-
                      append(self$vertices, v)
                  },
                  N_vertices = function() {
                    length(self$vertices)
                  },
                  add_pseudo_vertex = function(pv) {
                    self$pseudo_vertices <-
                      append(self$pseudo_vertices, pv)
                  },
                  N_pseudo_vertices = function() {
                    length(self$pseudo_vertices)
                  },
                  add_face = function(f) {
                    self$faces <-
                      append(self$faces, f)
                  },
                  N_faces = function() {
                    length(self$faces)
                  }))
```

We can do so much with a doubly-connected edge list.  For example
we can automatically compute a bounding box that contains all of
the vertices.  The parameter `q` specifies how much
multiplicative margin we add around the smallest bounding box.

```{r}
dcel$set("public", "compute_bounding_box",
function(q=0.05) {
  # Set bounding box edges to
  # extreme vertex positions
  self$xlim <- c(self$vertices[[1]]$pos[1],
                 self$vertices[[1]]$pos[1])
  self$ylim <- c(self$vertices[[1]]$pos[2],
                 self$vertices[[1]]$pos[2])

  for (v in self$vertices[-1]) {
    if (v$pos[1] > self$xlim[2]) {
      self$xlim[2] <- v$pos[1]
    } else if (v$pos[1] < self$xlim[1]) {
      self$xlim[1] <- v$pos[1]
    }

    if (v$pos[2] > self$ylim[2]) {
      self$ylim[2] <- v$pos[2]
    } else if (v$pos[2] < self$ylim[1]) {
      self$ylim[1] <- v$pos[2]
    }
  }

  # Add proportional margins to bounding box
  dx <- self$xlim[2] - self$xlim[1]
  self$xlim[1] <- self$xlim[1] - q * dx
  self$xlim[2] <- self$xlim[2] + q * dx

  dy <- self$ylim[2] - self$ylim[1]
  self$ylim[1] <- self$ylim[1] - q * dy
  self$ylim[2] <- self$ylim[2] + q * dy
})
```

Given a bounding box we can then introduce pseudo-vertices to
anchor any dangling half-edges.

```{r}
dcel$set("public", "anchor_dangling_edges",
function(z, xs, ys) {
  if (is.null(z$site)) {
    # Dangling edge
    he <- z$he

    nu <- he$twin$origin$pos
    zeta <- -nu

    i_left <- he$left_face
    i_right <- he$twin$left_face
    dangling_he <- he

    # Anchored edges
    he <- he$nxt
    zeta <- zeta + he$twin$origin$pos

    he <- he$twin$nxt
    zeta <- zeta + he$twin$origin$pos

    # Compute bisecting vector in the opposite
    # direction of the anchored edges
    delta <- c(ys[i_right] - ys[i_left],
               xs[i_left]  - xs[i_right])

    # Compute closest bounding box intersection
    bh <- ifelse(delta[1] > 0, self$xlim[2], self$xlim[1])
    bv <- ifelse(delta[2] > 0, self$ylim[2], self$ylim[1])

    pv_x <- bh
    pv_y <- nu[2] + delta[2] * (pv_x - nu[1]) / delta[1]
    if (pv_y < self$ylim[1] || pv_y > self$ylim[2]) {
      pv_y <- bv
      pv_x <- nu[1] + delta[1] * (pv_y - nu[2]) / delta[2]
    }

    # Anchor dangling edge to a new pseudo-vertex
    # positioned at the bounding box intersection
    pv <- vertex$new(c(pv_x, pv_y), NULL, NULL, NULL)
    dangling_he$origin <- pv
    self$add_pseudo_vertex(pv)

    # Recurse
    self$anchor_dangling_edges(z$left, xs, ys)
    self$anchor_dangling_edges(z$right, xs, ys)
  }
})
```

Most relevant to the construction of a Voronoi diagrams we can
follow the half-edges to automatically identify all of the faces
in the graph, and hence cells in the diagram.

```{r}
dcel$set("private", "reset_visitation",
function() {
  for (v in self$vertices) {
    he <- v$he
    he$visited <- FALSE
    he$twin$visited <- FALSE
    he <- he$prv
    he$visited <- FALSE
    he$twin$visited <- FALSE
    he <- he$twin$prv
    he$visited <- FALSE
    he$twin$visited <- FALSE
  }
})
```

```{r}
dcel$set("public", "find_faces",
function(he=NULL) {

  if (is.null(he)) {
    private$reset_visitation()
    he <- self$vertices[[1]]$he
  }

  init_he <- he
  open_face <- FALSE
  closed_face <- FALSE

  while (!he$visited) {
    he$visited <- TRUE
    he <- he$nxt
    if (is.null(he)) {
      open_face <- TRUE
      break
    }
    if (he == init_he) {
      closed_face <- TRUE
      break
    }
  }

  he <- init_he
  if (open_face) {
    # Reverse to beginning of open boundary
    while(!is.null(he$prv)) {
      he <- he$prv
    }

    self$add_face(face$new(he$left_face, he))

    # Shoot along open boundary
    while(!is.null(he$nxt)) {
      if(!he$twin$visited) self$find_faces(he$twin)
      he <- he$nxt
    }
  }
  if (closed_face) {
    self$add_face(face$new(he$left_face, he))

    # Shoot along closed boundary
    while(1) {
      if(!he$twin$visited) self$find_faces(he$twin)
      he <- he$nxt
      if (he == init_he) break
    }
  }
})
```

```{r}
dcel$set("public", "trace_face",
function(f) {
  print(paste('Face', f$site))
  he <- f$he

  perimeter_xs <- c()
  perimeter_ys <- c()

  while (1) {
    perimeter_xs <- c(perimeter_xs, he$origin$pos[1])
    perimeter_ys <- c(perimeter_ys, he$origin$pos[2])

    if (is.null(he$nxt)) {
      perimeter_xs <- c(perimeter_xs, he$twin$origin$pos[1])
      perimeter_ys <- c(perimeter_ys, he$twin$origin$pos[2])
      break
    }

    he <- he$nxt

    if (he == f$he) {
      # Close the perimeter
      perimeter_xs <- c(perimeter_xs, he$origin$pos[1])
      perimeter_ys <- c(perimeter_ys, he$origin$pos[2])
      break
    }
  }

  cat('  ', sprintf('%.3f/%.3f,', perimeter_xs, perimeter_ys), '\n')
})
```

Why spend all of this time creating graphs if we can't visualize
them with aesthetic flair?

```{r}
plot_half_edge <- function(pi, pf, lwd, col=c_mid_teal,
                           delta, eps, theta, gamma) {
  r <- pf - pi
  v <- r / sqrt(sum(r**2))

  r_perp <- c(r[2], -r[1])
  w <- r_perp / sqrt(sum(r_perp**2))

  # Perpendicular offset
  gammax <- -gamma * w[1]
  gammay <- -gamma * w[2]

  # Arrow body
  lines(c(pi[1], pf[1]) + gammax,
        c(pi[2], pf[2]) + gammay,
        lwd=lwd, col=col)

  # Arrow head
  phi_rad <- (90 - theta) * (3.14159265 / 180)
  p3 <- pf - (delta + eps * tan(phi_rad)) * v - eps * w
  p4 <- pf - delta * v

  polygon(c(pf[1], p3[1], p4[1], pf[1]) + gammax,
          c(pf[2], p3[2], p4[2], pf[2]) + gammay,
          col=col, border=NA)
}
```

```{r}
plot_vertex_edges <- function(v, lwd=2, delta=0.1, eps=0.05,
                              theta=60, gamma=0.025) {
  points(v$pos[1], v$pos[2],
         pch=16, col=c_light_teal)

  he <- v$he
  if (   !is.null(he$origin)
         && !is.null(he$twin$origin)) {
    plot_half_edge(he$origin$pos, he$twin$origin$pos,
                   lwd, ifelse(he$visited, c_mid, c_mid_teal),
                   delta, eps, theta, gamma)
    plot_half_edge(he$twin$origin$pos, he$origin$pos,
                   lwd, ifelse(he$twin$visited, c_mid, c_mid_teal),
                   delta, eps, theta, gamma)
  }

  he <- he$prv
  if (   !is.null(he$origin)
         && !is.null(he$twin$origin)) {
    plot_half_edge(he$origin$pos, he$twin$origin$pos,
                   lwd, ifelse(he$visited, c_mid, c_mid_teal),
                   delta, eps, theta, gamma)
    plot_half_edge(he$twin$origin$pos, he$origin$pos,
                   lwd, ifelse(he$twin$visited, c_mid, c_mid_teal),
                   delta, eps, theta, gamma)
  }

  he <- he$twin$prv
  if (   !is.null(he$origin)
         && !is.null(he$twin$origin)) {
    plot_half_edge(he$origin$pos, he$twin$origin$pos,
                   lwd, ifelse(he$visited, c_mid, c_mid_teal),
                   delta, eps, theta, gamma)
    plot_half_edge(he$twin$origin$pos, he$origin$pos,
                   lwd, ifelse(he$twin$visited, c_mid, c_mid_teal),
                   delta, eps, theta, gamma)
  }
}
```

```{r}
plot_face_boundary <- function(f, lwd=2, col=c_mid_teal,
                               delta=0.1, eps=0.05,
                               theta=60, gamma=0.025) {
  he <- f$he

  while (1) {
    points(he$origin$pos[1], he$origin$pos[2],
           pch=16, col=col)

    plot_half_edge(he$origin$pos, he$twin$origin$pos,
                   lwd, col,
                   delta, eps, theta, gamma)

    he <- he$nxt
    if (is.null(he) || he == f$he) break
  }
}
```

```{r}
plot_face <- function(f, col=c_mid_teal) {
  # Trace boundary
  he <- f$he

  perimeter_xs <- c()
  perimeter_ys <- c()

  while (1) {
    perimeter_xs <- c(perimeter_xs, he$origin$pos[1])
    perimeter_ys <- c(perimeter_ys, he$origin$pos[2])

    if (is.null(he$nxt)) {
      perimeter_xs <- c(perimeter_xs, he$twin$origin$pos[1])
      perimeter_ys <- c(perimeter_ys, he$twin$origin$pos[2])
      break
    }

    he <- he$nxt

    if (he == f$he) {
      # Close the perimeter
      perimeter_xs <- c(perimeter_xs, he$origin$pos[1])
      perimeter_ys <- c(perimeter_ys, he$origin$pos[2])
      break
    }
  }

  polygon(perimeter_xs, perimeter_ys, col=col, border=NULL)
}
```

## Running Fortune's Algorithm

Alright, let's cut the pedagogical tension and finally build a
Voronoi diagram.

We start by defining a collection of sites.

```{r}
N <- 10

set.seed(54482248)
xs <- runif(N, -1, 1)
ys <- runif(N, -1, 1)


par(mfrow=c(1, 1), mar=c(5, 5, 4, 1))
plot(xs, ys, col=c_dark, pch=16)
```

Next we initialize the event queue and beach line tree.

```{r}
queue <- event_queue$new(1:N, xs)

bl <- beachline$new()
voronoi <- dcel$new()

cat('Initial queue:\n')
for (i in queue$idxs) {
  arc <- queue$events[[i]]$arc
  if (is.null(queue$events[[i]]$arc)) {
    cat(paste('  S =', sprintf('%.3f', queue$events[[i]]$x),
              ':',
              'Site Event at', queue$events[[i]]$site),
              '\n')
  } else {
    cat(paste('  S =', sprintf('%.3f', queue$events[[i]]$x),
              ':',
              'Vertex Event at', queue$events[[i]]$arc$site),
              '\n')
  }
}
```

We can completely exhaust the event queue by running.

```{r}
#| eval: false
while (queue$is_not_empty()) {
  bl$process_next_event(queue, voronoi)
}
```

Here, however, we'll proceed a little bit more deliberately.
This function processes the next event with extreme verbosity,
communicating the type of the event, plotting both the current
beach line tree and the current beach line, and then displaying
the status of the event queue.

```{r}
process_next_event <- function() {
  if (is.null(queue$next_event()$arc)) {
    cat(paste0('Processing Site Event (',
               queue$next_event()$site, ')\n'))
  } else {
    cat(paste('Processing Vertex Event (',
              queue$next_event()$arc$site, ')\n'))
  }

  bl$process_next_event(queue, voronoi)

  par(mfrow=c(2, 1))
  plot_tree(bl, text_cex=0.6)
  if (is.null(queue$next_event()$x)) {
    plot_beachline(bl, 2,
                   c(-4, 2), c(-2, 2))
  } else {
    plot_beachline(bl, queue$next_event()$x,
                   c(-4, 2), c(-2, 2))
  }

  cat('\n')
  cat('Remaining queue:\n')
  for (i in queue$idxs) {
    arc <- queue$events[[i]]$arc
    if (is.null(queue$events[[i]]$arc)) {
      cat(paste('  S =', sprintf('%.3f', queue$events[[i]]$x),
                ':',
                'Site Event at', queue$events[[i]]$site),
                '\n')
    } else {
      cat(paste('  S =', sprintf('%.3f', queue$events[[i]]$x),
                ':',
                'Vertex Event at', queue$events[[i]]$arc$site),
                '\n')
    }
  }
}
```

Initially the beach line is empty.  The first site event
populates the beach line with a single parabola.

```{r}
process_next_event()
```

The next site event splits this monolithic parabola up into three
arcs.  Note that the breakpoint $b_{26}$ is too far left to be
seen in this plot.

```{r}
process_next_event()
```

At this point we'll jump through the next few events until we're
about to encounter our first vertex event.

```{r}
for (i in 1:3) {
  bl$process_next_event(queue, voronoi)
}

par(mfrow=c(2, 1))
plot_tree(bl, text_cex=0.6)
if (is.null(queue$next_event()$x)) {
  plot_beachline(bl, 2,
                 c(-4, 2), c(-2, 2))
} else {
  plot_beachline(bl, queue$next_event()$x,
                 c(-4, 2), c(-2, 2))
}

cat('')
cat('Remaining queue:')
for (i in queue$idxs) {
  arc <- queue$events[[i]]$arc
  if (is.null(queue$events[[i]]$arc)) {
    cat(paste('  S =', sprintf('%.3f', queue$events[[i]]$x),
              ':',
              'Site Event at', queue$events[[i]]$site),
              '\n')
  } else {
    cat(paste('  S =', sprintf('%.3f', queue$events[[i]]$x),
              ':',
              'Vertex Event at', queue$events[[i]]$arc$site),
              '\n')
  }
}
```

In the first vertex event we start paring the beach line down.
Here the arc node generated by site $2$ that sits between the
breakpoints $b_{32}$ and $b_{27}$ is removed while our first
Voronoi vertex is created.

```{r}
process_next_event()
```

To not draw this process out _too_ long let's run Fortune's
algorithm to completion.

```{r}
while (queue$is_not_empty()) {
  bl$process_next_event(queue, voronoi)
}
```

The remaining breakpoint nodes each contain a dangling Voronoi
half-edge.

```{r}
par(mfrow=c(2, 1))
plot_tree(bl, text_cex=0.6)
if (is.null(queue$next_event()$x)) {
  plot_beachline(bl, 2,
                 c(-4, 2), c(-2, 2))
} else {
  plot_beachline(bl, queue$next_event()$x,
                 c(-4, 2), c(-2, 2))
}
```

To visualize the entire Voronoi graph within the confines of a
finite plot we first compute a bounding box.

```{r}
voronoi$compute_bounding_box(0.25)
```

Note that we can always specify a bounding box of our own.  This
can be useful, for example, if we want larger or more symmetric
bounding boxes for visualization purposes.

```{r}
#| eval: false
voronoi$xlim <- c(-3, 3)
voronoi$ylim <- c(-3, 3)
```

Given a bounding box geometry we can anchor dangling half-edges
to pseudo-vertices on the bounding box.

```{r}
voronoi$anchor_dangling_edges(bl$root, xs, ys)
```

Perhaps the simplest way to plot the Voronoi graph is to plot
each Voronoi vertex along with the half-edges originating from
and terminating at those vertices.

```{r}
par(mfrow=c(1, 1), mar=c(0, 0, 0, 0))
plot(xs, ys, axes=FALSE, ann=FALSE, col=c_dark, pch=16,
     xlim=voronoi$xlim, ylim=voronoi$ylim,
     main="")

for (n in 1:N)
  text(xs[n], ys[n] - 0.1, n)

for (v in voronoi$vertices) {
  plot_vertex_edges(v, delta=0.08, eps=0.04, gamma=0.0075)
}
```

The real content of the Voronoi diagram, however, is in the
faces.

```{r}
voronoi$find_faces()
```

We can always examine the face boundaries directly, which is
useful for exporting the Voronoi geometry to other environments.

```{r}
for (f in voronoi$faces) {
  voronoi$trace_face(f)
}
```

Or we can just plot the faces directly.

```{r}
par(mfrow=c(1, 1), mar=c(0, 0, 0, 0))
plot(xs, ys, axes=FALSE, ann=FALSE, col=c_light, pch=16,
     xlim=voronoi$xlim, ylim=voronoi$ylim,
     main="")

for (n in 1:N)
  text(xs[n], ys[n] - 0.1, n)

for (f in voronoi$faces) {
  plot_face_boundary(f, col=c_light_teal,
                     delta=0.08, eps=0.04, gamma=0.01)
}

f <- voronoi$faces[[1]]
plot_face(f, col=c_light_teal)
plot_face_boundary(f, col=c_dark_teal,
                   delta=0.08, eps=0.04, gamma=0.01)
points(xs[f$site], ys[f$site], col=c_dark, pch=16)
text(xs[f$site], ys[f$site] - 0.1, f$site)
```

# Conclusion

Voronoi diagrams are straightforward to define but subtle to
actually derive in practice.  Fortune's algorithm is able to
systematize the construction of Voronoi diagrams.  Understanding
how the algorithm actually achieves this task, however, requires
some careful planar geometry.  Moreover, effectively implementing
Fortune's algorithm in practice requires familiarity with a
variety of nontrivial data structures.

From a more optimistic perspective, Fortune's algorithm provides
a wealth of learning opportunities...

# Acknowledgements {-}

A very special thanks to everyone supporting me on Patreon:

# References {-}

<div id="refs"></div>

# License {-}

A repository containing all of the files used to generate this chapter is
available on [GitHub](
https://github.com/betanalpha/quarto_misc/tree/main/voronoi).

The code in this case study is copyrighted by Michael Betancourt and licensed
under the new BSD (3-clause) license:

[https://opensource.org/licenses/BSD-3-Clause](https://opensource.org/licenses/BSD-3-Clause)

The text and figures in this chapter are copyrighted by Michael Betancourt
and licensed under the CC BY-NC 4.0 license:

[https://creativecommons.org/licenses/by-nc/4.0/](https://creativecommons.org/licenses/by-nc/4.0/)

# Original Computing Environment {-}

```{r, comment=NA}
sessionInfo()
```
